---
title: "\\textbf{Lab 11: Synthetic Control}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, November 5"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Construct a synthetic control and read the weight table.
2.  Judge the design by pre-treatment fit, not by the size of the gap.
3.  Run placebo inference in space and in time.
4.  Check robustness by leaving donors out.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(tidysynth)

data("smoking", package = "tidysynth")
c(rows = nrow(smoking), states = n_distinct(smoking$state),
  years = paste(range(smoking$year), collapse = "-"))
```

# The design

**Question.** What did California's Proposition 99 (1988) do to cigarette
consumption?

**Treated unit.** California. **Donor pool.** 38 states that adopted no
comparable programme. **Outcome.** Per-capita cigarette sales.
**Pre-period.** 1970–1988.

::: callout-note
The donor pool was restricted *on substantive grounds before estimating*: states
with their own large tobacco taxes or control programmes are excluded, because
they are treated too. That decision is part of the design, not a robustness
check.
:::

```{r}
#| label: fig-raw
#| fig-cap: "Raw trajectories. California is one line among 39; the eye cannot pick out a treatment effect."
smoking |>
  mutate(ca = state == "California") |>
  ggplot(aes(year, cigsale, group = state, colour = ca, linewidth = ca)) +
  geom_line(alpha = .8) +
  scale_colour_manual(values = c("grey78", "firebrick"), guide = "none") +
  scale_linewidth_manual(values = c(.3, 1.1), guide = "none") +
  geom_vline(xintercept = 1988, linetype = 2) +
  labs(x = NULL, y = "packs per capita") +
  theme_minimal(base_size = 10)
```

# Building the synthetic control

```{r}
#| label: build
sc <- smoking |>
  synthetic_control(outcome = cigsale, unit = state, time = year,
                    i_unit = "California", i_time = 1988,
                    generate_placebos = TRUE) |>
  # predictors measured over the pre-period
  generate_predictor(time_window = 1980:1988,
                     ln_income = mean(lnincome, na.rm = TRUE),
                     ret_price = mean(retprice, na.rm = TRUE),
                     youth     = mean(age15to24, na.rm = TRUE)) |>
  generate_predictor(time_window = 1984:1988,
                     beer_sales = mean(beer, na.rm = TRUE)) |>
  # lagged outcomes: these do most of the work
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = 1988, cigsale_1988 = cigsale) |>
  generate_weights(optimization_window = 1970:1988) |>
  generate_control()
```

## Who is synthetic California?

```{r}
#| label: weights
sc |> grab_unit_weights() |> filter(weight > 0.001) |>
  arrange(desc(weight)) |> knitr::kable(digits = 3)
```

```{r}
#| label: predweights
sc |> grab_predictor_weights() |> arrange(desc(weight)) |> knitr::kable(digits = 3)
```

::: callout-important
**Print the weight table.** It is the single most informative output of a
synthetic control study: readers can decide for themselves whether a weighted
average of these states is a plausible California. No regression coefficient
offers that.
:::

## Balance

```{r}
#| label: balance
sc |> grab_balance_table() |> knitr::kable(digits = 2)
```

# The result

```{r}
#| label: fig-trend
#| fig-cap: "California and synthetic California. The pre-1988 lines nearly coincide, which is what licenses reading the post-1988 gap as an effect."
plot_trends(sc)
```

```{r}
#| label: fig-diff
#| fig-cap: "The gap between California and its synthetic control."
plot_differences(sc)
```

```{r}
#| label: effect-size
gaps <- sc |> grab_synthetic_control() |>
  mutate(gap = real_y - synth_y)
c(mean_gap_post = mean(gaps$gap[gaps$time_unit > 1988]),
  gap_2000 = gaps$gap[gaps$time_unit == 2000],
  rmspe_pre = sqrt(mean(gaps$gap[gaps$time_unit <= 1988]^2)))
```

The pre-treatment RMSPE is the number to look at first. If it were large relative
to the post-treatment gap, there would be nothing to report.

# Inference

## Placebos in space

```{r}
#| label: fig-placebo
#| fig-cap: "Each grey line is a donor state treated as if it had adopted Proposition 99. California is in black."
plot_placebos(sc)
```

`plot_placebos()` drops donors whose pre-treatment fit is much worse than
California's — otherwise states the method could never match would dominate the
picture for reasons unrelated to treatment.

## The RMSPE ratio

```{r}
#| label: fig-mspe
#| fig-cap: "Ratio of post- to pre-treatment RMSPE. California is the largest of all 39 units."
plot_mspe_ratio(sc)
```

```{r}
#| label: pvalue
sig <- sc |> grab_significance()
sig |> select(unit_name, type, mspe_ratio, rank, fishers_exact_pvalue) |>
  head(6) |> knitr::kable(digits = 3)
```

The `fishers_exact_pvalue` column is the unit's rank divided by the number of
units. With 38 donors the smallest attainable value is $1/39 \approx 0.026$ — a
hard floor worth stating when you report it. Strictly this is a *placebo rank*
rather than an exact test: exactness would require treatment to have been
randomly assigned among these 39 states.

## A placebo in time

Pretend the programme began in 1980, eight years early. Nothing should happen.

```{r}
#| label: intime
sc_1980 <- smoking |>
  filter(year <= 1988) |>
  synthetic_control(outcome = cigsale, unit = state, time = year,
                    i_unit = "California", i_time = 1980,
                    generate_placebos = FALSE) |>
  generate_predictor(time_window = 1970:1980,
                     ln_income = mean(lnincome, na.rm = TRUE),
                     ret_price = mean(retprice, na.rm = TRUE)) |>
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_weights(optimization_window = 1970:1980) |>
  generate_control()

g80 <- sc_1980 |> grab_synthetic_control() |> mutate(gap = real_y - synth_y)
c(mean_gap_1981_1988 = mean(g80$gap[g80$time_unit > 1980]),
  rmspe_pre = sqrt(mean(g80$gap[g80$time_unit <= 1980]^2)))
```

A gap near zero after the fake intervention is reassuring. A large one would mean
the method manufactures effects.

## Leave-one-donor-out

```{r}
#| label: loo
donors <- sc |> grab_unit_weights() |> filter(weight > 0.01) |> pull(unit)

## refit the EXACT original pipeline after dropping each donor, so that any
## change is attributable to the donor and not to a specification change
loo <- lapply(donors, function(dd) {
  s <- smoking |> filter(state != dd) |>
    synthetic_control(outcome = cigsale, unit = state, time = year,
                      i_unit = "California", i_time = 1988,
                      generate_placebos = FALSE) |>
    generate_predictor(time_window = 1980:1988,
                       ln_income = mean(lnincome, na.rm = TRUE),
                       ret_price = mean(retprice, na.rm = TRUE),
                       youth     = mean(age15to24, na.rm = TRUE)) |>
    generate_predictor(time_window = 1984:1988,
                       beer_sales = mean(beer, na.rm = TRUE)) |>
    generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
    generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
    generate_predictor(time_window = 1988, cigsale_1988 = cigsale) |>
    generate_weights(optimization_window = 1970:1988) |>
    generate_control()
  g <- s |> grab_synthetic_control() |> mutate(gap = real_y - synth_y)
  tibble::tibble(dropped = dd, mean_gap = mean(g$gap[g$time_unit > 1988]))
}) |> bind_rows()

knitr::kable(loo, digits = 2)
```

If dropping any single donor moved the estimate substantially, the result would
rest on one state rather than on a comparison group.

# Exercises

1.  **Predictor sensitivity.** Re-estimate using only lagged outcomes as
    predictors (drop income, price, beer, youth). How much do the weights and the
    estimate move? What does that tell you about how much work the covariates are
    doing?

2.  **Pre-period length.** Re-estimate using only 1980–1988 to fit the weights.
    Does the pre-treatment fit worsen? Does the estimate change? Connect this to
    the factor-model justification from lecture.

3.  **A bad donor pool.** Deliberately include a state with its own tobacco
    programme. How does that contaminate the comparison, and would any diagnostic
    catch it?

4.  **DID for comparison.** Estimate a plain 2×2 DID comparing California to the
    unweighted average of all donors, before and after 1988. How does it differ
    from the synthetic control estimate, and why?

5.  **A different treated unit.** Using `set.seed(723)`, pick a donor state at
    random, treat it as treated, and run the whole pipeline. Write the paragraph you would write if
    this were your result — then explain what stops you from believing it.

# Session information

```{r}
#| label: session
sessionInfo()
```
