---
title: "\\textbf{Lab 6: Matching, Weighting, and Doubly Robust Estimation}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, October 1"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{setspace}
        \setstretch{1.12}
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\},breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999, width = 76)
```

# The question

**Estimand.** The average difference in log personal earnings between completing
a bachelor's degree and not completing one, among the full-time workers aged
25--64 in our GSS analytic sample, 2010--2022.

::: callout-note
Note the deliberately modest population. The GSS is a complex sample, and doing
this properly for the U.S. population would require the survey weights (`wt`) and
PSU/stratum-aware inference. We work unweighted here so that the adjustment
methods are visible on their own; that is a simplification, and you should say so
if you write it up.
:::

We also exclude `hours` from the adjustment set. Hours worked is plausibly a
*consequence* of completing college, so conditioning on it would be the
post-treatment error from Week 5; it would answer a different question.

This is an *observational* question with obvious confounding: people who complete
college differ from those who do not in family background, measured ability, and
much else. We are going to assume conditional ignorability given the covariates
we have, do the adjustment five different ways, and then ask how much an
unobserved confounder would have to matter to overturn the answer.

::: callout-warning
Assuming conditional ignorability here is almost certainly wrong. That is the
point of the last section. Treat today's numbers as an exercise in *method*, not
as an estimate of the return to college.
:::

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(MatchIt)
library(cobalt)
library(sensemakr)
library(sandwich); library(lmtest)
library(boot)

gss <- readRDS("../Data/gss_earnings.rds") |>
  mutate(D = ba, Y = lnearn)

covs <- c("pareduc", "wordsum", "female", "black", "otherace", "exper")
c(n = nrow(gss), treated = sum(gss$D), control = sum(1 - gss$D))
```

# Where we start: the naive comparison

```{r}
#| label: naive
naive <- lm(Y ~ D, data = gss)
reg   <- lm(Y ~ D + pareduc + wordsum + female + black + otherace + exper +
              I(exper^2), data = gss)

c(naive = coef(naive)["D"], regression = coef(reg)["D"])
```

Adjustment moves the estimate substantially. But regression is only one of the
adjustment strategies available, and it makes assumptions we have not examined.

```{r}
#| label: imbalance
bal.tab(reg_formula <- D ~ pareduc + wordsum + female + black + otherace +
          exper, data = gss, un = TRUE, thresholds = c(m = 0.1))
```

The `Diff.Un` column is the standardized mean difference. Values well above
$0.1$, here `pareduc` and `wordsum` especially, mean the two groups are not
comparable as they stand.

# Exact matching on cells

The lecture's matching section began with the cleanest estimator there is:
coarsen the covariates into cells, take the difference in means inside every
cell that holds both graduates and non-graduates, and average the cell
differences with the weights of the estimand you named. This section
reproduces the lecture's numbers by hand, then checks them against `MatchIt`.

## Building the cells

```{r}
#| label: exact-cells
## coarsen the three continuous covariates at rough terciles; keep the binaries as they are
gss <- gss |>
  mutate(pareduc_b = cut(pareduc, c(-Inf, 11, 13, Inf), labels = c("<12", "12-13", "14+")),
         wordsum_b = cut(wordsum, c(-Inf, 5, 7, Inf),   labels = c("0-5", "6-7", "8-10")),
         exper_b   = cut(exper,   c(-Inf, 16, 29, Inf), labels = c("0-16", "17-29", "30+")),
         cell = interaction(female, black, pareduc_b, wordsum_b, exper_b))
## one row per cell: how many treated and untreated, and their mean outcomes
cells <- gss |>
  group_by(cell) |>
  summarise(n1 = sum(D), n0 = sum(1 - D),
            m1 = mean(Y[D == 1]), m0 = mean(Y[D == 0]), .groups = "drop") |>
  mutate(n = n1 + n0, tau = m1 - m0, both = n1 > 0 & n0 > 0)
c(possible = nlevels(gss$cell), occupied = nrow(cells), both_arms = sum(cells$both),
  treated_only = sum(cells$n1 > 0 & cells$n0 == 0), untreated_only = sum(cells$n0 > 0 & cells$n1 == 0))
## coverage: who is lost when the one-armed cells are dropped
c(treated_dropped = sum(cells$n1[!cells$both]), untreated_dropped = sum(cells$n0[!cells$both]))
```

## The three estimands, and what OLS returns

```{r}
#| label: exact-estimates
m <- filter(cells, both)
## ATE: cell shares among matched units; ATT: shares among matched treated; ATU: among matched untreated
c(naive = mean(gss$Y[gss$D == 1]) - mean(gss$Y[gss$D == 0]),
  ATE = sum(m$tau * m$n)  / sum(m$n),
  ATT = sum(m$tau * m$n1) / sum(m$n1),
  ATU = sum(m$tau * m$n0) / sum(m$n0),
  ## OLS saturated in the cells: the variance-weighted average from the lecture
  OLS_cells = coef(lm(Y ~ D + cell, data = gss))["D"])
```

The four adjusted numbers are the same 92 cell differences under four sets of
weights. Bootstrap the whole procedure for standard errors; resampling units is
valid here because nothing is nearest-neighbour.

```{r}
#| label: exact-boot
boot_exact <- function(d) {
  b <- d[sample(nrow(d), replace = TRUE), ]
  t <- b |> group_by(cell) |>
    summarise(n1 = sum(D), n0 = sum(1 - D), y1 = sum(Y * D), y0 = sum(Y * (1 - D)), .groups = "drop") |>
    filter(n1 > 0, n0 > 0) |> mutate(tau = y1 / n1 - y0 / n0)
  c(ATE = sum(t$tau * (t$n1 + t$n0)) / sum(t$n1 + t$n0), ATT = sum(t$tau * t$n1) / sum(t$n1))
}
set.seed(723)
apply(replicate(200, boot_exact(gss)), 1, sd)
```

## The checks: coverage and balance

```{r}
#| label: exact-balance
## standardized mean difference: (weighted mean treated - weighted mean untreated) / pooled SD
covs <- c("pareduc", "wordsum", "exper", "female", "black", "otherace")
sd_pool <- sapply(covs, function(v) sqrt((var(gss[gss$D == 1, v]) + var(gss[gss$D == 0, v])) / 2))
smd <- function(w1, w0) sapply(covs, function(v)
  (weighted.mean(gss[[v]], w1) - weighted.mean(gss[[v]], w0)) / sd_pool[v])
## weights that reproduce the matched estimators: each unit gets (target share of its cell) / (units of its arm in the cell)
cw <- as.character(gss$cell)
share <- setNames(m$n / sum(m$n), as.character(m$cell))        # ATE target: matched-population cell shares
share1 <- setNames(m$n1 / sum(m$n1), as.character(m$cell))     # ATT target: treated cell shares
n1c <- setNames(m$n1, as.character(m$cell)); n0c <- setNames(m$n0, as.character(m$cell))
get <- function(v, k) ifelse(is.na(v[k]), 0, v[k])
w1_ate <- gss$D * get(share, cw) / pmax(get(n1c, cw), 1)
w0_ate <- (1 - gss$D) * get(share, cw) / pmax(get(n0c, cw), 1)
w1_att <- gss$D * get(share1, cw) / pmax(get(n1c, cw), 1)
w0_att <- (1 - gss$D) * get(share1, cw) / pmax(get(n0c, cw), 1)
round(rbind(raw = smd(gss$D, 1 - gss$D), ATE_matched = smd(w1_ate, w0_ate), ATT_matched = smd(w1_att, w0_att)), 3)
```

Exactly matched variables balance by construction; the coarsened ones keep the
within-band imbalance the bands allow; `otherace`, which we did not match on,
moves on its own. None of these numbers is a test.

## The same thing in `MatchIt`

```{r}
#| label: exact-matchit
## method = "exact" builds the same cells; estimand chooses the weights
for (est in c("ATE", "ATT")) {
  mm <- matchit(D ~ female + black + pareduc_b + wordsum_b + exper_b,
                data = gss, method = "exact", estimand = est)
  md <- match.data(mm)
  cat(est, ":", round(coef(lm(Y ~ D, data = md, weights = weights))["D"], 3),
      " matched n =", nrow(md), "\n")
}
```

# The propensity score

```{r}
#| label: ps
ps_mod <- glm(D ~ pareduc + wordsum + female + black + otherace +
                exper + I(exper^2),
              data = gss, family = binomial)
gss$ps <- fitted(ps_mod)

summary(gss$ps)
```

```{r}
#| label: fig-overlap
#| fig-cap: "Distribution of the estimated propensity score by treatment status. Overlap is imperfect in both tails."
ggplot(gss, aes(ps, fill = factor(D))) +
  geom_density(alpha = .45, colour = NA) +
  scale_fill_manual(values = c("0" = "navy", "1" = "firebrick"),
                    labels = c("no BA", "BA"), name = NULL) +
  labs(x = "estimated propensity score", y = "density") +
  theme_minimal(base_size = 10)
```

```{r}
#| label: overlap-tails
## how many treated units sit where controls are scarce, and vice versa?
c(treated_above_0.9 = sum(gss$D == 1 & gss$ps > 0.9),
  controls_above_0.9 = sum(gss$D == 0 & gss$ps > 0.9),
  controls_below_0.1 = sum(gss$D == 0 & gss$ps < 0.1),
  treated_below_0.1 = sum(gss$D == 1 & gss$ps < 0.1))
```

::: callout-note
Note the trade-off from lecture: a propensity model that predicts treatment
*well* is a model that separates the groups, which is exactly the situation in
which comparison is hardest. Judge the model by **balance**, not by AUC.
:::

# Matching

## Nearest-neighbour on the propensity score

```{r}
#| label: match-nn
m_nn <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                  exper + I(exper^2),
                data = gss, method = "nearest", distance = "glm",
                estimand = "ATT", replace = FALSE)
m_nn
```

```{r}
#| label: fig-balance
#| fig-cap: "Covariate balance before and after nearest-neighbour matching. Everything should land inside the dashed lines."
love.plot(m_nn, thresholds = c(m = 0.1), binary = "std",
          colors = c("firebrick", "navy")) +
  theme_minimal(base_size = 9)
```

```{r}
#| label: match-est
md_nn <- match.data(m_nn)
fit_nn <- lm(Y ~ D + pareduc + wordsum + female + black + otherace +
               exper + I(exper^2),
             data = md_nn, weights = weights)
round(coeftest(fit_nn, vcov = vcovCL, cluster = ~ subclass)["D", ], 4)
```

Two things to notice. First, we still run a regression on the matched data, 
matching is *preprocessing*, and the regression mops up residual imbalance.
Second, the standard errors are clustered on the matched pair, because the
matching induces dependence.

## Mahalanobis distance

Instead of collapsing the covariates to a propensity score, match on the
covariates directly, scaling by their covariance so that no variable dominates
through its units.

```{r}
#| label: match-mahal
m_mah <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper,
                 data = gss, method = "nearest", distance = "mahalanobis",
                 estimand = "ATT")
summary(m_mah)$nn
```

## A caliper

A caliper refuses matches beyond a set distance on the propensity score,
measured in standard deviations of the linear predictor.

```{r}
#| label: match-caliper
m_cal <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper + I(exper^2),
                 data = gss, method = "nearest", distance = "glm",
                 estimand = "ATT", caliper = 0.1, std.caliper = TRUE)
summary(m_cal)$nn
```

Look at the **Unmatched** row for the treated. Those respondents have no
comparable counterpart in the data, and dropping them silently redefines the
estimand; you are now describing the subpopulation of degree-holders who have
plausible non-degree counterparts.

## Coarsened exact matching

```{r}
#| label: cem
m_cem <- matchit(D ~ pareduc + wordsum + female + black + exper,
                 data = gss, method = "cem", estimand = "ATT")
summary(m_cem)$nn
```

CEM is the most transparent method: it matches exactly within coarsened bins, so
balance is guaranteed by construction. The price is written plainly in the
number of discarded observations.

# Weighting

## IPW by hand

```{r}
#| label: ipw
e  <- gss$ps
D  <- gss$D
Y  <- gss$Y

## Horvitz-Thompson
ate_ht <- mean(D * Y / e) - mean((1 - D) * Y / (1 - e))

## Hajek / stabilized -- divide by the sum of the weights
w1 <- D / e
w0 <- (1 - D) / (1 - e)
ate_sipw <- sum(w1 * Y) / sum(w1) - sum(w0 * Y) / sum(w0)

c(horvitz_thompson = ate_ht, stabilized = ate_sipw)
```

```{r}
#| label: weight-diagnostics
w <- ifelse(D == 1, 1 / e, 1 / (1 - e))
ess <- sum(w)^2 / sum(w^2)
c(max_weight = max(w), ess = ess, n = length(w), ess_share = ess / length(w))
```

The **effective sample size** is what your standard errors should reflect. If a
handful of observations carry most of the weight, you have far less information
than $n$ suggests.

## ATT and overlap weights

```{r}
#| label: other-weights
att_w  <- ifelse(D == 1, 1, e / (1 - e))          # ATT weights
ato_w  <- ifelse(D == 1, 1 - e, e)                # overlap weights

wmean_diff <- function(wt) {
  sum(wt * Y * D) / sum(wt * D) - sum(wt * Y * (1 - D)) / sum(wt * (1 - D))
}
c(ATE_stabilized = ate_sipw,
  ATT = wmean_diff(att_w),
  ATO_overlap = wmean_diff(ato_w))
```

Three different numbers, because they are three different **estimands**, not
three attempts at one. The overlap-weighted estimate describes the subpopulation
where comparison is most informative, and its weights are bounded by
construction.

## Does weighting actually balance the covariates?

```{r}
#| label: weight-balance
bal.tab(D ~ pareduc + wordsum + female + black + otherace + exper,
        data = gss, weights = ifelse(D == 1, 1 / e, 1 / (1 - e)),
        method = "weighting", un = TRUE, thresholds = c(m = 0.1))
```

# G-computation and AIPW

```{r}
#| label: gcomp
## fit separate outcome models in each arm, then impute both potential outcomes
out_form <- Y ~ pareduc + wordsum + female + black + otherace + exper +
  I(exper^2)
m1 <- lm(out_form, data = gss[D == 1, ])
m0 <- lm(out_form, data = gss[D == 0, ])

mu1 <- predict(m1, newdata = gss)
mu0 <- predict(m0, newdata = gss)

ate_gcomp <- mean(mu1 - mu0)
ate_gcomp
```

```{r}
#| label: aipw
aipw_terms <- (mu1 - mu0) +
  D * (Y - mu1) / e -
  (1 - D) * (Y - mu0) / (1 - e)

ate_aipw <- mean(aipw_terms)

## the influence-function standard error comes free
se_aipw <- sd(aipw_terms) / sqrt(length(aipw_terms))

c(estimate = ate_aipw, se = se_aipw,
  lower = ate_aipw - 1.96 * se_aipw, upper = ate_aipw + 1.96 * se_aipw)
```

::: callout-tip
The AIPW estimator is an average of $n$ i.i.d. terms, so its standard error is
just the standard deviation of those terms over $\sqrt n$. That is not a
coincidence, `aipw_terms` is the *efficient influence function*, and this is
the pattern that Week 13 generalizes.
:::

## Everything in one place

```{r}
#| label: compare-all
results <- tibble::tibble(
  method = c("Naive difference", "OLS adjustment", "NN matching (ATT)",
             "IPW stabilized (ATE)", "ATT weights", "Overlap weights (ATO)",
             "G-computation (ATE)", "AIPW (ATE)"),
  estimate = c(coef(naive)["D"], coef(reg)["D"], coef(fit_nn)["D"],
               ate_sipw, wmean_diff(att_w), wmean_diff(ato_w),
               ate_gcomp, ate_aipw)
)
knitr::kable(results, digits = 4)
```

If these had disagreed sharply, that itself would be the finding: it would tell
us the answer is coming from modelling choices rather than from comparison.

## Bootstrapping the whole procedure

Standard errors must account for the fact that the propensity score was
*estimated*. The safest route is to bootstrap the entire pipeline.

```{r}
#| label: boot-aipw
aipw_fun <- function(data, idx) {
  d <- data[idx, ]
  psm <- glm(D ~ pareduc + wordsum + female + black + otherace +
               exper + I(exper^2), data = d, family = binomial)
  ee  <- fitted(psm)
  f1 <- lm(out_form, data = d[d$D == 1, ]); f0 <- lm(out_form, data = d[d$D == 0, ])
  m1p <- predict(f1, newdata = d); m0p <- predict(f0, newdata = d)
  mean((m1p - m0p) + d$D * (d$Y - m1p) / ee -
         (1 - d$D) * (d$Y - m0p) / (1 - ee))
}
set.seed(723)
bo <- boot(gss, aipw_fun, R = 500)
c(estimate = bo$t0, boot_se = sd(bo$t), analytic_se = se_aipw)
```

The two standard errors agree closely, which is reassuring: the influence-function
formula is doing its job.

# Sensitivity analysis

All of the above assumed we measured every confounder. We did not. So ask
instead: **how strong would an unobserved confounder have to be?** Two
answers, both computed from the long regression `reg` and both on the slides.

## Oster's coefficient-stability bound

Oster (2019) compares the coefficient *without* controls (short regression)
to the coefficient *with* them (long regression), and scales the movement by
how much $R^2$ the controls added. If controls moved the coefficient little
but also explained little, that is no comfort.

```{r}
#| label: oster
## the short regression: Y on D alone; the long one is `reg` from above
short <- lm(Y ~ D, data = gss)
b_dot <- unname(coef(short)["D"]); r_dot <- summary(short)$r.squared   ## uncontrolled
b_til <- unname(coef(reg)["D"]);   r_til <- summary(reg)$r.squared     ## controlled
c(beta_short = b_dot, r2_short = r_dot, beta_long = b_til, r2_long = r_til)

## Oster's assumption: selection on unobservables = delta x selection on observables,
## and Rmax = the R^2 we would reach with the unobservables in.
## The movement per unit of R^2 gained is extrapolated to the R^2 still to gain:
oster_beta  <- function(delta, rmax) b_til - delta * (b_dot - b_til) * (rmax - r_til) / (r_til - r_dot)
## delta*: the delta at which beta* hits zero
oster_delta <- function(rmax) b_til * (r_til - r_dot) / ((b_dot - b_til) * (rmax - r_til))

## Oster's rule of thumb is Rmax = 1.3 x R^2 of the long regression; show two harsher choices too
rmax <- c(rmax_1.3 = 1.3 * r_til, rmax_2 = 2 * r_til, rmax_1 = 1)
rbind(beta_star_delta1 = sapply(rmax, oster_beta, delta = 1),
      delta_star       = sapply(rmax, oster_delta))
```

Read the two rows. `beta_star_delta1` is the estimate if unobservables were
exactly as important as the observables ($\delta = 1$); the finding is called
robust if the interval from the controlled estimate to this number excludes
zero. `delta_star` is how many times stronger than the observables the
unobservables would have to be to erase the effect; Oster's standard is
$\delta^* > 1$. Here even $R_{\max} = 1$ leaves $\delta^* = 1.8$. This is the
simplified formula from the paper; the `robomit` package computes the exact
one.

## Partial $R^2$ and the robustness value

```{r}
#| label: sensemakr
sens <- sensemakr(model = reg, treatment = "D",
                  benchmark_covariates = "wordsum",
                  kd = 1:3)
sens
```

```{r}
#| label: fig-sens
#| fig-cap: "Contours of the adjusted estimate as a function of how much residual variation in treatment and outcome an unobserved confounder explains. Benchmarks show multiples of the vocabulary score."
plot(sens)
```

Read the **robustness value**: it is the share of residual variation in both the
treatment and the outcome that a confounder would need to explain to reduce the
estimate to zero. Compare it to the benchmark: if a confounder three times as
strong as measured verbal ability would be needed, the finding is reasonably
robust; if half of one would do it, it is not.

```{r}
#| label: rv-by-hand
## the robustness value from the t statistic alone: RV = (sqrt(f^4 + 4 f^2) - f^2) / 2, f = |t| / sqrt(df)
tt <- coef(summary(reg))["D", "t value"]; df <- reg$df.residual
f  <- abs(tt) / sqrt(df)
c(t = tt, df = df, RV_by_hand = (sqrt(f^4 + 4 * f^2) - f^2) / 2,
  RV_sensemakr = sens$sensitivity_stats$rv_q)
```

# Selection on unobservables: the Heckman correction

Everything above conditioned on $X$. The last section of the lecture asked what
can be done when units select on something we did *not* measure. Heckman's
answer is to write down the selection rule and the joint distribution of the two
errors, and then compute the gap $\E[\epsilon_i \mid D_i = 1] -
\E[\epsilon_i \mid D_i = 0]$ instead of assuming it away.

::: callout-note
The GSS offers no credible *excluded variable*, a variable that moves college
completion but has no direct effect on earnings. Without one, the lecture showed
that the correction rests on the curvature of $\phi/\Phi$ alone. So this section
works on simulated data where the true effect is known, exactly as the lecture's
table did (`figs/w6_heckman.R`).
:::

## The world

```{r}
#| label: heck-sim
set.seed(723)
n <- 2000
rho <- 0.5; sigma <- 1; tau <- 1
X <- rnorm(n)      ## the observed control, in both equations
Z <- rnorm(n)      ## the excluded variable: shifts selection, not the outcome
u <- rnorm(n)      ## selection-equation error, variance 1
## outcome error correlated rho with u: that correlation IS selection on unobservables
eps <- sigma * (rho * u + sqrt(1 - rho^2) * rnorm(n))
D <- as.integer(0.5 * X + Z + u > 0)   ## the selection rule
Y <- 1 + 0.5 * X + tau * D + eps       ## the outcome, one constant effect tau = 1
dat <- data.frame(Y, D, X, Z)
## OLS with X in the regression: biased upward, because X is not the confounder
coef(lm(Y ~ X + D, data = dat))["D"]
```

## The two steps by hand

```{r}
#| label: heck-2step
## step 1: probit of D on the selection regressors, X and the excluded Z
step1 <- glm(D ~ X + Z, family = binomial(link = "probit"), data = dat)
c_hat <- predict(step1, type = "link")     ## Z_i' gamma-hat
## step 2: the correction term, one expression for both groups
dat$lambda <- ifelse(dat$D == 1,  dnorm(c_hat) / pnorm(c_hat),
                                 -dnorm(c_hat) / (1 - pnorm(c_hat)))
step2 <- lm(Y ~ X + D + lambda, data = dat)
## D recovers tau; the coefficient on lambda estimates rho * sigma = 0.5
coef(summary(step2))[, c("Estimate", "Std. Error", "t value")]
```

The $t$-test on `lambda` is the test of selection on unobservables. But the
standard errors printed above are wrong: `lambda` was estimated in step 1, and
$\epsilon_i$ is heteroskedastic given $D_i$. Bootstrap both steps together.

```{r}
#| label: heck-boot
two_step <- function(d, selection = D ~ X + Z) {
  s1 <- glm(selection, family = binomial(link = "probit"), data = d)
  ch <- predict(s1, type = "link")
  d$lambda <- ifelse(d$D == 1, dnorm(ch) / pnorm(ch), -dnorm(ch) / (1 - pnorm(ch)))
  coef(lm(Y ~ X + D + lambda, data = d))[c("D", "lambda")]
}
B <- 200
boot <- replicate(B, two_step(dat[sample(n, replace = TRUE), ]))
apply(boot, 1, sd)    ## compare with the step-2 OLS standard errors above
```

## Both equations by maximum likelihood

The two-step estimator is the Week 3 idea of "estimate, then plug in." Fitting
the probit and the outcome equation jointly by maximum likelihood (FIML) uses the
bivariate normal likelihood directly, returns $\rho$ and $\sigma$ separately, and
gives standard errors that need no bootstrap.

```{r}
#| label: heck-fiml
library(sampleSelection)
## treatReg: selection equation first, then the outcome equation with D in it
fiml <- treatReg(D ~ X + Z, Y ~ X + D, data = dat)
summary(fiml)
```

The wage version, where $Y$ is observed only for the $D_i = 1$ rows, is
`heckit(selection = D ~ X + Z, outcome = Y ~ X, data = ...)` in the same package.

## Without an excluded variable

```{r}
#| label: heck-noZ
## the selection equation now has only X: lambda is a smooth function of X and D
s1 <- glm(D ~ X, family = binomial(link = "probit"), data = dat)
ch <- predict(s1, type = "link")
dat$lambda_noZ <- ifelse(dat$D == 1, dnorm(ch) / pnorm(ch), -dnorm(ch) / (1 - pnorm(ch)))
## how much of the correction term do X and D already explain linearly?
summary(lm(lambda_noZ ~ X + D, data = dat))$r.squared
## what is left identifies tau: the curvature of phi/Phi, and nothing else
coef(summary(lm(Y ~ X + D + lambda_noZ, data = dat)))[c("D", "lambda_noZ"), ]
```

Compare the standard error on `D` with the one from the two-step with `Z`.
The excluded variable must move treatment and have no direct effect on $Y$.
That is next week's instrument.

## When the errors are not normal

```{r}
#| label: heck-skew
## the common shock is a centred, unit-variance, right-skewed chi-square(1)
v <- (rchisq(n, 1) - 1) / sqrt(2)
eps_s <- sigma * (rho * v + sqrt(1 - rho^2) * rnorm(n))
D_s <- as.integer(0.5 * X + Z + v > 0)
Y_s <- 1 + 0.5 * X + tau * D_s + eps_s
dat_s <- data.frame(Y = Y_s, D = D_s, X, Z)
two_step(dat_s)                          ## with the excluded variable
two_step(dat_s, selection = D ~ X)       ## without it
```

With an excluded variable the normal formula for the gap is only slightly off.
Without one, the estimator *is* the normal formula, and it can land anywhere.

# Exercises

1.  **Estimand discipline.** Redo the analysis targeting the **ATT** throughout
    (matching, ATT weights, and an ATT version of AIPW). Do the answers move?
    Explain why the ATE and ATT differ here in terms of who is on the margin of
    completing college.

2.  **Trimming.** Restrict to $0.1 < \hat e < 0.9$ and recompute IPW and AIPW.
    How much does the estimate move, and how many observations did you drop?
    Write one sentence stating what population your new estimate describes.

3.  **Deliberate misspecification, on simulated data.** Double robustness cannot
    be "confirmed" against observational data, because the causal truth is
    unknown. Simulate a world with a known ATE, then fit (a) a correct outcome
    model with a badly wrong propensity model and (b) the reverse. Confirm that
    AIPW stays close to the truth in both cases while IPW and g-computation each
    fail in one. This is double robustness in action.

4.  **A bad control.** Add `prestige` (occupational prestige) to the covariate
    set. It is plausibly a *consequence* of holding a degree. Report how the
    estimate changes and explain, using the Week 5 vocabulary, why the new number
    answers a different question.

5.  **Balance versus fit.** Compute the AUC of the propensity model. Then add
    interactions until balance improves. Does the AUC go up or down? Discuss what
    this means for the "propensity score paradox" from lecture.

6.  **Effective sample size.** Plot the IPW weights against $\hat e$. Identify the
    ten most heavily weighted observations. Refit without them, how much does
    the estimate move relative to its standard error?

7.  **An excluded variable for college.** Propose a variable in the GSS, or one
    you would collect, that shifts completing a bachelor's degree but has no
    direct effect on earnings. Write the argument for "no direct effect" in
    three sentences, then list what would break it. Explain, using the last
    section, why the correction is worth little without such a variable.

# Session information

```{r}
#| label: session
## R version and the packages this lab loaded
c(R = R.version.string, sapply(sessionInfo()$otherPkgs, \(p) as.character(p$Version)))
```
