---
title: "\\textbf{Lab 6: Matching, Weighting, and Doubly Robust Estimation}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, October 1"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{setspace}
        \setstretch{1.12}
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\},breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999, width = 76)
```

# The question

**Estimand.** The average difference in log personal earnings between completing
a bachelor's degree and not completing one, among the full-time workers aged
25--64 in our GSS analytic sample, 2010--2022.

::: callout-note
Note the deliberately modest population. The GSS is a complex sample, and doing
this properly for the U.S. population would require the survey weights (`wt`) and
PSU/stratum-aware inference. We work unweighted here so that the adjustment
methods are visible on their own; that is a simplification, and you should say so
if you write it up.
:::

We also exclude `hours` from the adjustment set. Hours worked is plausibly a
*consequence* of completing college, so conditioning on it would be the
post-treatment error from Week 5; it would answer a different question.

This is an *observational* question with obvious confounding: people who complete
college differ from those who do not in family background, measured ability, and
much else. We are going to assume conditional ignorability given the covariates
we have, do the adjustment five different ways, and then ask how much an
unobserved confounder would have to matter to overturn the answer.

::: callout-warning
Assuming conditional ignorability here is almost certainly wrong. That is the
point of the last section. Treat today's numbers as an exercise in *method*, not
as an estimate of the return to college.
:::

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(MatchIt)
library(cobalt)
library(sensemakr)
library(sandwich); library(lmtest)
library(boot)

gss <- readRDS("../Data/gss_earnings.rds") |>
  mutate(D = ba, Y = lnearn)

covs <- c("pareduc", "wordsum", "female", "black", "otherace", "exper")
c(n = nrow(gss), treated = sum(gss$D), control = sum(1 - gss$D))
```

# Where we start: the naive comparison

```{r}
#| label: naive
naive <- lm(Y ~ D, data = gss)
reg   <- lm(Y ~ D + pareduc + wordsum + female + black + otherace + exper +
              I(exper^2), data = gss)

c(naive = coef(naive)["D"], regression = coef(reg)["D"])
```

Adjustment moves the estimate substantially. But regression is only one of the
adjustment strategies available, and it makes assumptions we have not examined.

```{r}
#| label: imbalance
bal.tab(reg_formula <- D ~ pareduc + wordsum + female + black + otherace +
          exper, data = gss, un = TRUE, thresholds = c(m = 0.1))
```

The `Diff.Un` column is the standardized mean difference. Values well above
$0.1$, here `pareduc` and `wordsum` especially, mean the two groups are not
comparable as they stand.

# Exact matching on cells

The lecture's matching section began with the cleanest estimator there is:
coarsen the covariates into cells, take the difference in means inside every
cell that holds both graduates and non-graduates, and average the cell
differences with the weights of the estimand you named. This section
reproduces the lecture's numbers by hand, then checks them against `MatchIt`.

## Building the cells

```{r}
#| label: exact-cells
## coarsen the three continuous covariates at rough terciles; keep the binaries as they are
gss <- gss |>
  mutate(pareduc_b = cut(pareduc, c(-Inf, 11, 13, Inf), labels = c("<12", "12-13", "14+")),
         wordsum_b = cut(wordsum, c(-Inf, 5, 7, Inf),   labels = c("0-5", "6-7", "8-10")),
         exper_b   = cut(exper,   c(-Inf, 16, 29, Inf), labels = c("0-16", "17-29", "30+")),
         cell = interaction(female, black, pareduc_b, wordsum_b, exper_b))
## one row per cell: how many treated and untreated, and their mean outcomes
cells <- gss |>
  group_by(cell) |>
  summarise(n1 = sum(D), n0 = sum(1 - D),
            m1 = mean(Y[D == 1]), m0 = mean(Y[D == 0]), .groups = "drop") |>
  mutate(n = n1 + n0, tau = m1 - m0, both = n1 > 0 & n0 > 0)
c(possible = nlevels(gss$cell), occupied = nrow(cells), both_arms = sum(cells$both),
  treated_only = sum(cells$n1 > 0 & cells$n0 == 0), untreated_only = sum(cells$n0 > 0 & cells$n1 == 0))
## coverage: who is lost when the one-armed cells are dropped
c(treated_dropped = sum(cells$n1[!cells$both]), untreated_dropped = sum(cells$n0[!cells$both]))
```

## The three estimands, and what OLS returns

```{r}
#| label: exact-estimates
m <- filter(cells, both)
## ATE: cell shares among matched units; ATT: shares among matched treated; ATU: among matched untreated
est_hand <- c(ATE = sum(m$tau * m$n)  / sum(m$n),
              ATT = sum(m$tau * m$n1) / sum(m$n1),
              ATU = sum(m$tau * m$n0) / sum(m$n0))
c(naive = mean(gss$Y[gss$D == 1]) - mean(gss$Y[gss$D == 0]),
  est_hand,
  ## OLS saturated in the cells: the variance-weighted average from the lecture
  OLS_cells = coef(lm(Y ~ D + cell, data = gss))["D"])
```

The four adjusted numbers are the same 92 cell differences under four sets of
weights.

### The standard error by hand

The lecture's formula: the estimate is a weighted sum of cell differences, so
its variance is the weighted sum of the cells' variances,
$\Var(\widehat{\text{ATE}}) = \sum_x (n_x/n)^2\,(s^2_{1x}/n_{1x} + s^2_{0x}/n_{0x})$,
with the ATT and ATU weights in place of $n_x/n$.

```{r}
#| label: exact-se
## within-cell variances of Y in each arm (NA when a cell has one unit in that arm)
v1 <- tapply(gss$Y[gss$D == 1], gss$cell[gss$D == 1], var)
v0 <- tapply(gss$Y[gss$D == 0], gss$cell[gss$D == 0], var)
m$v1 <- v1[as.character(m$cell)]; m$v0 <- v0[as.character(m$cell)]
m$v1[is.na(m$v1)] <- 0; m$v0[is.na(m$v0)] <- 0
## the variance of one cell difference, then the weighted sum with each target's weights
cell_var <- m$v1 / m$n1 + m$v0 / m$n0
se_hand <- c(ATE = sqrt(sum((m$n  / sum(m$n))^2  * cell_var)),
             ATT = sqrt(sum((m$n1 / sum(m$n1))^2 * cell_var)),
             ATU = sqrt(sum((m$n0 / sum(m$n0))^2 * cell_var)))
round(se_hand, 4)
```

Now the bootstrap: resample units, rebuild the cells, recompute all three.
Resampling is valid here because the estimate is a smooth function of cell
means; nothing is nearest-neighbour.

```{r}
#| label: exact-boot
boot_exact <- function(d) {
  b <- d[sample(nrow(d), replace = TRUE), ]
  t <- b |> group_by(cell) |>
    summarise(n1 = sum(D), n0 = sum(1 - D), y1 = sum(Y * D), y0 = sum(Y * (1 - D)), .groups = "drop") |>
    filter(n1 > 0, n0 > 0) |> mutate(tau = y1 / n1 - y0 / n0)
  c(ATE = sum(t$tau * (t$n1 + t$n0)) / sum(t$n1 + t$n0),
    ATT = sum(t$tau * t$n1) / sum(t$n1),
    ATU = sum(t$tau * t$n0) / sum(t$n0))
}
set.seed(723)
se_boot <- apply(replicate(200, boot_exact(gss)), 1, sd)
round(rbind(formula = se_hand, bootstrap = se_boot), 4)
```

The two rows agree to within a couple of thousandths: the formula is doing
what the bootstrap does, without the 200 refits.

## The same thing in `MatchIt`, with standard errors

`method = "exact"` builds the same cells; `estimand` chooses the weights
(`"ATC"` is MatchIt's name for the ATU). `match.data()` returns the matched
units with a `weights` column, and a weighted regression of `Y` on `D` is a
difference of weighted means, so its coefficient reproduces the by-hand
estimate exactly. For the standard error use a sandwich (`vcovHC`): the
default `lm` standard error ignores the weights.

```{r}
#| label: exact-matchit
mi <- list()
for (est in c("ATE", "ATT", "ATC")) {
  mm <- matchit(D ~ female + black + pareduc_b + wordsum_b + exper_b,
                data = gss, method = "exact", estimand = est)
  md <- match.data(mm)
  f  <- lm(Y ~ D, data = md, weights = weights)
  mi[[est]] <- c(estimate = unname(coef(f)["D"]),
                 se_HC1   = unname(sqrt(vcovHC(f, type = "HC1")["D", "D"])))
}
mi_tab <- do.call(rbind, mi); rownames(mi_tab) <- c("ATE", "ATT", "ATU")
## by hand (section 3.2) next to MatchIt + WLS + sandwich
round(cbind(hand_estimate = c(est_hand["ATE"], est_hand["ATT"], est_hand["ATU"]),
            matchit_estimate = mi_tab[, "estimate"],
            hand_se = se_hand, matchit_se_HC1 = mi_tab[, "se_HC1"]), 4)
```

The estimates are identical. The sandwich standard errors sit a little above
the by-hand ones: the sandwich measures each unit's deviation from its *arm's*
weighted mean, so it counts between-cell variation in $Y$ as noise, whereas the
formula uses only within-cell variances. Both treat the cell weights as fixed.

### The balance check

`love.plot()` draws the standardized mean difference of every covariate
before and after matching. The variables we matched on are the coarsened
bins, which balance exactly by construction; `addl` adds the underlying
continuous variables, where the within-band imbalance shows, and `otherace`,
which we did not match on at all.

```{r}
#| label: fig-exact-balance
#| fig-cap: "Covariate balance before and after exact matching on the cells (ATE weights). The bins balance exactly; the continuous variables keep the within-band imbalance the bins allow."
mm_ate <- matchit(D ~ female + black + pareduc_b + wordsum_b + exper_b,
                  data = gss, method = "exact", estimand = "ATE")
love.plot(mm_ate, data = gss, addl = ~ pareduc + wordsum + exper + otherace,
          thresholds = c(m = 0.1), binary = "std",
          colors = c("firebrick", "navy")) +
  theme_minimal(base_size = 9)
```

# The propensity score

```{r}
#| label: ps
## a logit for Pr(D = 1 | X); family = binomial("logit") says so explicitly
## (binomial() alone would also default to the logit link, but say what you mean)
ps_mod <- glm(D ~ pareduc + wordsum + female + black + otherace +
                exper + I(exper^2),
              data = gss, family = binomial("logit"))
gss$ps <- fitted(ps_mod)

summary(gss$ps)
```

```{r}
#| label: fig-overlap
#| fig-cap: "Distribution of the estimated propensity score by treatment status. Overlap is imperfect in both tails."
ggplot(gss, aes(ps, fill = factor(D))) +
  geom_density(alpha = .45, colour = NA) +
  scale_fill_manual(values = c("0" = "navy", "1" = "firebrick"),
                    labels = c("no BA", "BA"), name = NULL) +
  labs(x = "estimated propensity score", y = "density") +
  theme_minimal(base_size = 10)
```

```{r}
#| label: overlap-tails
## how many treated units sit where controls are scarce, and vice versa?
c(treated_above_0.9 = sum(gss$D == 1 & gss$ps > 0.9),
  controls_above_0.9 = sum(gss$D == 0 & gss$ps > 0.9),
  controls_below_0.1 = sum(gss$D == 0 & gss$ps < 0.1),
  treated_below_0.1 = sum(gss$D == 1 & gss$ps < 0.1))
```

::: callout-note
Note the trade-off from lecture: a propensity model that predicts treatment
*well* is a model that separates the groups, which is exactly the situation in
which comparison is hardest. Judge the model by **balance**, not by AUC.
:::

# Matching

## Nearest-neighbour on the propensity score

Each treated unit takes the control with the closest propensity score.
**With replacement** (`replace = TRUE`) the same control may serve several
treated units: every treated unit gets its best match, so balance is far
better, and the price is the reuse counts $K_j$ from lecture, which inflate the
variance and rule out the ordinary bootstrap. Without replacement, once the
good controls are used up the remaining treated units get poor matches.

```{r}
#| label: match-nn
m_nn <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                  exper + I(exper^2),
                data = gss, method = "nearest", distance = "glm",
                estimand = "ATT", replace = TRUE)
m_nn
```

```{r}
#| label: fig-balance
#| fig-cap: "Covariate balance before and after nearest-neighbour matching on the propensity score, with replacement. Everything lands inside the dashed lines."
love.plot(m_nn, thresholds = c(m = 0.1), binary = "std",
          colors = c("firebrick", "navy")) +
  theme_minimal(base_size = 9)
```

```{r}
#| label: match-nn-norep
## the same matching without replacement: 1,441 treated compete for 2,068 controls
## and the late ones get poor matches. Compare the adjusted standardized differences.
m_nn0 <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper + I(exper^2),
                 data = gss, method = "nearest", distance = "glm",
                 estimand = "ATT", replace = FALSE)
round(cbind(with_replacement = bal.tab(m_nn)$Balance$Diff.Adj,
            without          = bal.tab(m_nn0)$Balance$Diff.Adj)[-1, ], 3)
```

```{r}
#| label: match-est
## get_matches() returns one row per matched pair member, so a reused control
## appears once per treated unit it serves; `unit` identifies the person,
## `subclass` the matched pair. The data already have a column called id,
## hence the id = "unit" argument.
md_nn <- get_matches(m_nn, id = "unit")
c(rows = nrow(md_nn), distinct_controls = length(unique(md_nn$unit[md_nn$D == 0])),
  max_reuse = max(table(md_nn$unit[md_nn$D == 0])))
fit_nn <- lm(Y ~ D + pareduc + wordsum + female + black + otherace +
               exper + I(exper^2),
             data = md_nn, weights = weights)
## standard errors clustered on both the pair and the person (MatchIt's recommendation
## for matching with replacement); the formal answer is the Abadie-Imbens variance
## from lecture, and the ordinary bootstrap is NOT valid here
round(coeftest(fit_nn, vcov = vcovCL, cluster = ~ subclass + unit)["D", ], 4)
```

Two things to notice. First, we still run a regression on the matched data:
matching is *preprocessing*, and the regression mops up residual imbalance.
Second, the standard errors are clustered on the matched pair and on the
person, because a control reused $K_j$ times appears $K_j$ times.

The lecture's variance formula, with the $K_j^2$ term, is what
`Matching::Match()` reports (`install.packages("Matching")`). It matches on the
same propensity score and prints the Abadie-Imbens standard error directly,
with no regression afterwards.

```{r}
#| label: match-ai
library(Matching)
## one-to-one nearest neighbour on the propensity score, with replacement, ATT:
## the same design as m_nn above; m_nn$distance is MatchIt's estimated propensity score.
## Match() keeps every control tied at the closest distance, at equal weight (MatchIt picks
## one); the GSS has many identical covariate profiles, so ties are common here
mm <- Match(Y = gss$Y, Tr = gss$D, X = m_nn$distance, M = 1, estimand = "ATT",
            replace = TRUE)
summary(mm)
## how many distinct controls were used, and the largest K_j
c(distinct_controls = length(unique(mm$index.control)),
  max_reuse = max(table(mm$index.control)))
```

The `AI SE` line is the Abadie-Imbens standard error: it counts a control used
$K_j$ times $K_j^2$ times, which is why it is larger than a naive SE that treats
the matched rows as independent. It is close to the clustered SE above. The
matched sample is larger than MatchIt's because tied controls are all kept;
the estimate barely moves. (With `ties = FALSE` Match() drops the
Abadie-Imbens calculation and prints a naive SE, so leave the default.)

## Mahalanobis distance

Instead of collapsing the covariates to a propensity score, match on the
covariates directly, scaling by their covariance so that no variable dominates
through its units.

```{r}
#| label: match-mahal
m_mah <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper,
                 data = gss, method = "nearest", distance = "mahalanobis",
                 estimand = "ATT")
summary(m_mah)$nn
```

## A caliper

A caliper refuses matches beyond a set distance on the propensity score,
measured in standard deviations of the linear predictor.

```{r}
#| label: match-caliper
m_cal <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper + I(exper^2),
                 data = gss, method = "nearest", distance = "glm",
                 estimand = "ATT", caliper = 0.1, std.caliper = TRUE)
summary(m_cal)$nn
```

Look at the **Unmatched** row for the treated. Those respondents have no
comparable counterpart in the data, and dropping them silently redefines the
estimand; you are now describing the subpopulation of degree-holders who have
plausible non-degree counterparts.

## Coarsened exact matching

```{r}
#| label: cem
## MatchIt's default coarsening: each continuous variable is cut into bins by
## Sturges' rule (about log2(n) + 1 of them), binaries as they are
m_cem <- matchit(D ~ pareduc + wordsum + female + black + exper,
                 data = gss, method = "cem", estimand = "ATT")
summary(m_cem)$nn
round(bal.tab(m_cem, un = TRUE)$Balance[, c("Diff.Un", "Diff.Adj")], 3)
```

CEM balances the *coarsened* variables exactly, and nothing more: how much
imbalance remains on the underlying continuous variables depends entirely on
how wide the bins are. With MatchIt's default bins, which are narrow, the
residual differences above are essentially zero, but look at the `Unmatched`
row: 598 of the 1,441 treated units found no control in their bin and were
dropped, so the estimand is now the effect on the matchable treated. Section 3
used three wide bins per variable and lost 4 treated units, at the price of
standardized differences near $0.1$ on the continuous variables. The same
trade-off, chosen by the bins:

```{r}
#| label: cem-coarse
## our section-3 bins, handed to MatchIt as cutpoints
m_cem3 <- matchit(D ~ pareduc + wordsum + female + black + exper,
                  data = gss, method = "cem", estimand = "ATT",
                  cutpoints = list(pareduc = c(11.5, 13.5),
                                   wordsum = c(5.5, 7.5),
                                   exper   = c(16.5, 29.5)))
summary(m_cem3)$nn
round(bal.tab(m_cem3)$Balance[, "Diff.Adj", drop = FALSE], 3)
```

# Weighting

## IPW by hand

```{r}
#| label: ipw
e  <- gss$ps
D  <- gss$D
Y  <- gss$Y

## Horvitz-Thompson
ate_ht <- mean(D * Y / e) - mean((1 - D) * Y / (1 - e))

## Hajek / stabilized -- divide by the sum of the weights
w1 <- D / e
w0 <- (1 - D) / (1 - e)
ate_sipw <- sum(w1 * Y) / sum(w1) - sum(w0 * Y) / sum(w0)

c(horvitz_thompson = ate_ht, stabilized = ate_sipw)
```

```{r}
#| label: weight-diagnostics
w <- ifelse(D == 1, 1 / e, 1 / (1 - e))
ess <- sum(w)^2 / sum(w^2)
c(max_weight = max(w), ess = ess, n = length(w), ess_share = ess / length(w))
```

The **effective sample size** is what your standard errors should reflect. If a
handful of observations carry most of the weight, you have far less information
than $n$ suggests.

## ATT and overlap weights

```{r}
#| label: other-weights
att_w  <- ifelse(D == 1, 1, e / (1 - e))          # ATT weights
ato_w  <- ifelse(D == 1, 1 - e, e)                # overlap weights

wmean_diff <- function(wt) {
  sum(wt * Y * D) / sum(wt * D) - sum(wt * Y * (1 - D)) / sum(wt * (1 - D))
}
c(ATE_stabilized = ate_sipw,
  ATT = wmean_diff(att_w),
  ATO_overlap = wmean_diff(ato_w))
```

Three different numbers, because they are three different **estimands**, not
three attempts at one. The overlap-weighted estimate describes the subpopulation
where comparison is most informative, and its weights are bounded by
construction.

## The same thing as a weighted regression

The coefficient on a dummy in weighted least squares is the difference of the
weighted means, so `lm(Y ~ D, weights = w)` must reproduce each Hájek estimate
exactly. It also hands us a standard error: a sandwich (`HC1`), which treats
the weights as known. Nothing here accounts for the propensity score having
been estimated; the bootstrap in section 7 does.

```{r}
#| label: weight-wls
w_ate <- ifelse(D == 1, 1 / e, 1 / (1 - e))
## the argument is called wvec, not wt: gss has a survey-weight column named wt,
## and lm() would quietly pick up that column instead of our vector
wls <- function(wvec) {
  f <- lm(Y ~ D, data = gss, weights = wvec)
  c(WLS = unname(coef(f)["D"]), se_HC1 = unname(sqrt(vcovHC(f, type = "HC1")["D", "D"])))
}
wls_tab <- rbind(ATE = wls(w_ate), ATT = wls(att_w), ATO = wls(ato_w))
cbind(by_hand = c(ate_sipw, wmean_diff(att_w), wmean_diff(ato_w)), wls_tab)
```

The `by_hand` and `WLS` columns agree to every printed digit, as they must.

## Does weighting actually balance the covariates?

```{r}
#| label: weight-balance
bal.tab(D ~ pareduc + wordsum + female + black + otherace + exper,
        data = gss, weights = ifelse(D == 1, 1 / e, 1 / (1 - e)),
        method = "weighting", un = TRUE, thresholds = c(m = 0.1))
```

# G-computation and AIPW

```{r}
#| label: gcomp
## fit separate outcome models in each arm, then impute both potential outcomes
out_form <- Y ~ pareduc + wordsum + female + black + otherace + exper +
  I(exper^2)
m1 <- lm(out_form, data = gss[D == 1, ])
m0 <- lm(out_form, data = gss[D == 0, ])

mu1 <- predict(m1, newdata = gss)
mu0 <- predict(m0, newdata = gss)

ate_gcomp <- mean(mu1 - mu0)
ate_gcomp
```

```{r}
#| label: aipw
aipw_terms <- (mu1 - mu0) +
  D * (Y - mu1) / e -
  (1 - D) * (Y - mu0) / (1 - e)

ate_aipw <- mean(aipw_terms)

## the influence-function standard error comes free
se_aipw <- sd(aipw_terms) / sqrt(length(aipw_terms))

c(estimate = ate_aipw, se = se_aipw,
  lower = ate_aipw - 1.96 * se_aipw, upper = ate_aipw + 1.96 * se_aipw)
```

::: callout-tip
The AIPW estimator is an average of $n$ i.i.d. terms, so its standard error is
just the standard deviation of those terms over $\sqrt n$. That is not a
coincidence, `aipw_terms` is the *efficient influence function*, and this is
the pattern that Week 13 generalizes.
:::

## Bootstrapping the whole procedure

Standard errors must account for the fact that the propensity score was
*estimated*. The safest route is to bootstrap the entire pipeline.

```{r}
#| label: boot-aipw
aipw_fun <- function(data, idx) {
  d <- data[idx, ]
  psm <- glm(D ~ pareduc + wordsum + female + black + otherace +
               exper + I(exper^2), data = d, family = binomial)
  ee  <- fitted(psm)
  f1 <- lm(out_form, data = d[d$D == 1, ]); f0 <- lm(out_form, data = d[d$D == 0, ])
  m1p <- predict(f1, newdata = d); m0p <- predict(f0, newdata = d)
  mean((m1p - m0p) + d$D * (d$Y - m1p) / ee -
         (1 - d$D) * (d$Y - m0p) / (1 - ee))
}
set.seed(723)
bo <- boot(gss, aipw_fun, R = 500)
c(estimate = bo$t0, boot_se = sd(bo$t), analytic_se = se_aipw)
```

The two standard errors agree closely, which is reassuring: the influence-function
formula is doing its job.

# Sensitivity analysis

All of the above assumed we measured every confounder. We did not. So ask
instead: **how strong would an unobserved confounder have to be?** Two
answers, both computed from the long regression `reg` and both on the slides.

## Oster's coefficient-stability bound

Oster (2019) starts from the old habit of comparing the coefficient *without*
controls to the coefficient *with* them. A small move is reassuring only if the
controls explained a good share of the outcome; otherwise unobservables that
explain more could still move the coefficient a long way. Her method judges the
move relative to the gain in $R^2$ and extrapolates. The `robomit` package does
the arithmetic (`install.packages("robomit")`).

```{r}
#| label: oster
library(robomit)
## two inputs. R2max: how much of Y all confounders together could explain; Oster's rule
## of thumb is 1.3 x the R^2 of the regression with controls. delta: how important the
## unobservables are relative to the controls (1 = equally important)
r2_long <- summary(reg)$r.squared
con <- "pareduc + wordsum + female + black + otherace + exper + I(exper^2)"  ## the controls

## beta*: the effect if the unobservables were as important as the controls (delta = 1)
o_beta(y = "Y", x = "D", con = con, delta = 1, R2max = 1.3 * r2_long, type = "lm", data = gss)

## delta*: how many times as important as the controls the unobservables would have to be
## to drive the effect to zero
o_delta(y = "Y", x = "D", con = con, beta = 0, R2max = 1.3 * r2_long, type = "lm", data = gss)
```

Read the first row of each table. Oster's standard: the finding is robust if
the interval from the controlled coefficient to `beta*` excludes zero and
`delta*` is above one. Here `beta*` is 0.58 and `delta*` is 2.6: unobservables
would have to matter more than twice as much as everything we controlled for to
erase the BA premium.

## Partial $R^2$ and the robustness value

```{r}
#| label: sensemakr
sens <- sensemakr(model = reg, treatment = "D",
                  benchmark_covariates = "wordsum",
                  kd = 1:3)
sens
```

```{r}
#| label: fig-sens
#| fig-cap: "Contours of the adjusted estimate as a function of how much residual variation in treatment and outcome an unobserved confounder explains. Benchmarks show multiples of the vocabulary score."
plot(sens)
```

Read the **robustness value**: it is the share of residual variation in both the
treatment and the outcome that a confounder would need to explain to reduce the
estimate to zero. Compare it to the benchmark: if a confounder three times as
strong as measured verbal ability would be needed, the finding is reasonably
robust; if half of one would do it, it is not.

# Exercises

1.  **Estimand discipline.** Redo the analysis targeting the **ATT** throughout
    (matching, ATT weights, and an ATT version of AIPW). Do the answers move?
    Explain why the ATE and ATT differ here in terms of who is on the margin of
    completing college.

2.  **Trimming.** Restrict to $0.1 < \hat e < 0.9$ and recompute IPW and AIPW.
    How much does the estimate move, and how many observations did you drop?
    Write one sentence stating what population your new estimate describes.

3.  **Deliberate misspecification, on simulated data.** Double robustness cannot
    be "confirmed" against observational data, because the causal truth is
    unknown. Simulate a world with a known ATE, then fit (a) a correct outcome
    model with a badly wrong propensity model and (b) the reverse. Confirm that
    AIPW stays close to the truth in both cases while IPW and g-computation each
    fail in one. This is double robustness in action.

4.  **A bad control.** Add `prestige` (occupational prestige) to the covariate
    set. It is plausibly a *consequence* of holding a degree. Report how the
    estimate changes and explain, using the Week 5 vocabulary, why the new number
    answers a different question.

5.  **Balance versus fit.** Compute the AUC of the propensity model. Then add
    interactions until balance improves. Does the AUC go up or down? Discuss what
    this means for the "propensity score paradox" from lecture.

6.  **Effective sample size.** Plot the IPW weights against $\hat e$. Identify the
    ten most heavily weighted observations. Refit without them, how much does
    the estimate move relative to its standard error?

# Session information

```{r}
#| label: session
## R version and the packages this lab loaded
c(R = R.version.string, sapply(sessionInfo()$otherPkgs, \(p) as.character(p$Version)))
```
