---
title: "\\textbf{Lab 6: Matching, Weighting, and Doubly Robust Estimation}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, October 1"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# The question

**Estimand.** The average difference in log personal earnings between completing
a bachelor's degree and not completing one, among the full-time workers aged
25--64 in our GSS analytic sample, 2010--2022.

::: callout-note
Note the deliberately modest population. The GSS is a complex sample, and doing
this properly for the U.S. population would require the survey weights (`wt`) and
PSU/stratum-aware inference. We work unweighted here so that the adjustment
methods are visible on their own; that is a simplification, and you should say so
if you write it up.
:::

We also exclude `hours` from the adjustment set. Hours worked is plausibly a
*consequence* of completing college, so conditioning on it would be the
post-treatment error from Week 5; it would answer a different question.

This is an *observational* question with obvious confounding: people who complete
college differ from those who do not in family background, measured ability, and
much else. We are going to assume conditional ignorability given the covariates
we have, do the adjustment five different ways, and then ask how much an
unobserved confounder would have to matter to overturn the answer.

::: callout-warning
Assuming conditional ignorability here is almost certainly wrong. That is the
point of the last section. Treat today's numbers as an exercise in *method*, not
as an estimate of the return to college.
:::

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(MatchIt)
library(cobalt)
library(sensemakr)
library(sandwich); library(lmtest)
library(boot)

gss <- readRDS("../Data/gss_earnings.rds") |>
  mutate(D = ba, Y = lnearn)

covs <- c("pareduc", "wordsum", "female", "black", "otherace", "exper")
c(n = nrow(gss), treated = sum(gss$D), control = sum(1 - gss$D))
```

# Where we start: the naive comparison

```{r}
#| label: naive
naive <- lm(Y ~ D, data = gss)
reg   <- lm(Y ~ D + pareduc + wordsum + female + black + otherace + exper +
              I(exper^2), data = gss)

c(naive = coef(naive)["D"], regression = coef(reg)["D"])
```

Adjustment moves the estimate substantially. But regression is only one of the
adjustment strategies available, and it makes assumptions we have not examined.

```{r}
#| label: imbalance
bal.tab(reg_formula <- D ~ pareduc + wordsum + female + black + otherace +
          exper, data = gss, un = TRUE, thresholds = c(m = 0.1))
```

The `Diff.Un` column is the standardized mean difference. Values well above
$0.1$, here `pareduc` and `wordsum` especially, mean the two groups are not
comparable as they stand.

# The propensity score

```{r}
#| label: ps
ps_mod <- glm(D ~ pareduc + wordsum + female + black + otherace +
                exper + I(exper^2),
              data = gss, family = binomial)
gss$ps <- fitted(ps_mod)

summary(gss$ps)
```

```{r}
#| label: fig-overlap
#| fig-cap: "Distribution of the estimated propensity score by treatment status. Overlap is imperfect in both tails."
ggplot(gss, aes(ps, fill = factor(D))) +
  geom_density(alpha = .45, colour = NA) +
  scale_fill_manual(values = c("0" = "navy", "1" = "firebrick"),
                    labels = c("no BA", "BA"), name = NULL) +
  labs(x = "estimated propensity score", y = "density") +
  theme_minimal(base_size = 10)
```

```{r}
#| label: overlap-tails
## how many treated units sit where controls are scarce, and vice versa?
c(treated_above_0.9 = sum(gss$D == 1 & gss$ps > 0.9),
  controls_above_0.9 = sum(gss$D == 0 & gss$ps > 0.9),
  controls_below_0.1 = sum(gss$D == 0 & gss$ps < 0.1),
  treated_below_0.1 = sum(gss$D == 1 & gss$ps < 0.1))
```

::: callout-note
Note the trade-off from lecture: a propensity model that predicts treatment
*well* is a model that separates the groups, which is exactly the situation in
which comparison is hardest. Judge the model by **balance**, not by AUC.
:::

# Matching

## Nearest-neighbour on the propensity score

```{r}
#| label: match-nn
m_nn <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                  exper + I(exper^2),
                data = gss, method = "nearest", distance = "glm",
                estimand = "ATT", replace = FALSE)
m_nn
```

```{r}
#| label: fig-balance
#| fig-cap: "Covariate balance before and after nearest-neighbour matching. Everything should land inside the dashed lines."
love.plot(m_nn, thresholds = c(m = 0.1), binary = "std",
          colors = c("firebrick", "navy")) +
  theme_minimal(base_size = 9)
```

```{r}
#| label: match-est
md_nn <- match.data(m_nn)
fit_nn <- lm(Y ~ D + pareduc + wordsum + female + black + otherace +
               exper + I(exper^2),
             data = md_nn, weights = weights)
coeftest(fit_nn, vcov = vcovCL, cluster = ~ subclass)["D", ]
```

Two things to notice. First, we still run a regression on the matched data, 
matching is *preprocessing*, and the regression mops up residual imbalance.
Second, the standard errors are clustered on the matched pair, because the
matching induces dependence.

## Mahalanobis distance

Instead of collapsing the covariates to a propensity score, match on the
covariates directly, scaling by their covariance so that no variable dominates
through its units.

```{r}
#| label: match-mahal
m_mah <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper,
                 data = gss, method = "nearest", distance = "mahalanobis",
                 estimand = "ATT")
summary(m_mah)$nn
```

## A caliper

A caliper refuses matches beyond a set distance on the propensity score,
measured in standard deviations of the linear predictor.

```{r}
#| label: match-caliper
m_cal <- matchit(D ~ pareduc + wordsum + female + black + otherace +
                   exper + I(exper^2),
                 data = gss, method = "nearest", distance = "glm",
                 estimand = "ATT", caliper = 0.1, std.caliper = TRUE)
summary(m_cal)$nn
```

Look at the **Unmatched** row for the treated. Those respondents have no
comparable counterpart in the data, and dropping them silently redefines the
estimand; you are now describing the subpopulation of degree-holders who have
plausible non-degree counterparts.

## Coarsened exact matching

```{r}
#| label: cem
m_cem <- matchit(D ~ pareduc + wordsum + female + black + exper,
                 data = gss, method = "cem", estimand = "ATT")
summary(m_cem)$nn
```

CEM is the most transparent method: it matches exactly within coarsened bins, so
balance is guaranteed by construction. The price is written plainly in the
number of discarded observations.

# Weighting

## IPW by hand

```{r}
#| label: ipw
e  <- gss$ps
D  <- gss$D
Y  <- gss$Y

## Horvitz-Thompson
ate_ht <- mean(D * Y / e) - mean((1 - D) * Y / (1 - e))

## Hajek / stabilized -- divide by the sum of the weights
w1 <- D / e
w0 <- (1 - D) / (1 - e)
ate_sipw <- sum(w1 * Y) / sum(w1) - sum(w0 * Y) / sum(w0)

c(horvitz_thompson = ate_ht, stabilized = ate_sipw)
```

```{r}
#| label: weight-diagnostics
w <- ifelse(D == 1, 1 / e, 1 / (1 - e))
ess <- sum(w)^2 / sum(w^2)
c(max_weight = max(w), ess = ess, n = length(w), ess_share = ess / length(w))
```

The **effective sample size** is what your standard errors should reflect. If a
handful of observations carry most of the weight, you have far less information
than $n$ suggests.

## ATT and overlap weights

```{r}
#| label: other-weights
att_w  <- ifelse(D == 1, 1, e / (1 - e))          # ATT weights
ato_w  <- ifelse(D == 1, 1 - e, e)                # overlap weights

wmean_diff <- function(wt) {
  sum(wt * Y * D) / sum(wt * D) - sum(wt * Y * (1 - D)) / sum(wt * (1 - D))
}
c(ATE_stabilized = ate_sipw,
  ATT = wmean_diff(att_w),
  ATO_overlap = wmean_diff(ato_w))
```

Three different numbers, because they are three different **estimands**, not
three attempts at one. The overlap-weighted estimate describes the subpopulation
where comparison is most informative, and its weights are bounded by
construction.

## Does weighting actually balance the covariates?

```{r}
#| label: weight-balance
bal.tab(D ~ pareduc + wordsum + female + black + otherace + exper,
        data = gss, weights = ifelse(D == 1, 1 / e, 1 / (1 - e)),
        method = "weighting", un = TRUE, thresholds = c(m = 0.1))
```

# G-computation and AIPW

```{r}
#| label: gcomp
## fit separate outcome models in each arm, then impute both potential outcomes
out_form <- Y ~ pareduc + wordsum + female + black + otherace + exper +
  I(exper^2)
m1 <- lm(out_form, data = gss[D == 1, ])
m0 <- lm(out_form, data = gss[D == 0, ])

mu1 <- predict(m1, newdata = gss)
mu0 <- predict(m0, newdata = gss)

ate_gcomp <- mean(mu1 - mu0)
ate_gcomp
```

```{r}
#| label: aipw
aipw_terms <- (mu1 - mu0) +
  D * (Y - mu1) / e -
  (1 - D) * (Y - mu0) / (1 - e)

ate_aipw <- mean(aipw_terms)

## the influence-function standard error comes free
se_aipw <- sd(aipw_terms) / sqrt(length(aipw_terms))

c(estimate = ate_aipw, se = se_aipw,
  lower = ate_aipw - 1.96 * se_aipw, upper = ate_aipw + 1.96 * se_aipw)
```

::: callout-tip
The AIPW estimator is an average of $n$ i.i.d. terms, so its standard error is
just the standard deviation of those terms over $\sqrt n$. That is not a
coincidence, `aipw_terms` is the *efficient influence function*, and this is
the pattern that Week 13 generalizes.
:::

## Everything in one place

```{r}
#| label: compare-all
results <- tibble::tibble(
  method = c("Naive difference", "OLS adjustment", "NN matching (ATT)",
             "IPW stabilized (ATE)", "ATT weights", "Overlap weights (ATO)",
             "G-computation (ATE)", "AIPW (ATE)"),
  estimate = c(coef(naive)["D"], coef(reg)["D"], coef(fit_nn)["D"],
               ate_sipw, wmean_diff(att_w), wmean_diff(ato_w),
               ate_gcomp, ate_aipw)
)
knitr::kable(results, digits = 4)
```

If these had disagreed sharply, that itself would be the finding: it would tell
us the answer is coming from modelling choices rather than from comparison.

## Bootstrapping the whole procedure

Standard errors must account for the fact that the propensity score was
*estimated*. The safest route is to bootstrap the entire pipeline.

```{r}
#| label: boot-aipw
aipw_fun <- function(data, idx) {
  d <- data[idx, ]
  psm <- glm(D ~ pareduc + wordsum + female + black + otherace +
               exper + I(exper^2), data = d, family = binomial)
  ee  <- fitted(psm)
  f1 <- lm(out_form, data = d[d$D == 1, ]); f0 <- lm(out_form, data = d[d$D == 0, ])
  m1p <- predict(f1, newdata = d); m0p <- predict(f0, newdata = d)
  mean((m1p - m0p) + d$D * (d$Y - m1p) / ee -
         (1 - d$D) * (d$Y - m0p) / (1 - ee))
}
set.seed(723)
bo <- boot(gss, aipw_fun, R = 500)
c(estimate = bo$t0, boot_se = sd(bo$t), analytic_se = se_aipw)
```

The two standard errors agree closely, which is reassuring: the influence-function
formula is doing its job.

# Sensitivity analysis

All of the above assumed we measured every confounder. We did not. So ask
instead: **how strong would an unobserved confounder have to be?**

```{r}
#| label: sensemakr
sens <- sensemakr(model = reg, treatment = "D",
                  benchmark_covariates = "wordsum",
                  kd = 1:3)
sens
```

```{r}
#| label: fig-sens
#| fig-cap: "Contours of the adjusted estimate as a function of how much residual variation in treatment and outcome an unobserved confounder explains. Benchmarks show multiples of the vocabulary score."
plot(sens)
```

Read the **robustness value**: it is the share of residual variation in both the
treatment and the outcome that a confounder would need to explain to reduce the
estimate to zero. Compare it to the benchmark: if a confounder three times as
strong as measured verbal ability would be needed, the finding is reasonably
robust; if half of one would do it, it is not.

# Exercises

1.  **Estimand discipline.** Redo the analysis targeting the **ATT** throughout
    (matching, ATT weights, and an ATT version of AIPW). Do the answers move?
    Explain why the ATE and ATT differ here in terms of who is on the margin of
    completing college.

2.  **Trimming.** Restrict to $0.1 < \hat e < 0.9$ and recompute IPW and AIPW.
    How much does the estimate move, and how many observations did you drop?
    Write one sentence stating what population your new estimate describes.

3.  **Deliberate misspecification, on simulated data.** Double robustness cannot
    be "confirmed" against observational data, because the causal truth is
    unknown. Simulate a world with a known ATE, then fit (a) a correct outcome
    model with a badly wrong propensity model and (b) the reverse. Confirm that
    AIPW stays close to the truth in both cases while IPW and g-computation each
    fail in one. This is double robustness in action.

4.  **A bad control.** Add `prestige` (occupational prestige) to the covariate
    set. It is plausibly a *consequence* of holding a degree. Report how the
    estimate changes and explain, using the Week 5 vocabulary, why the new number
    answers a different question.

5.  **Balance versus fit.** Compute the AUC of the propensity model. Then add
    interactions until balance improves. Does the AUC go up or down? Discuss what
    this means for the "propensity score paradox" from lecture.

6.  **Effective sample size.** Plot the IPW weights against $\hat e$. Identify the
    ten most heavily weighted observations. Refit without them, how much does
    the estimate move relative to its standard error?

# Session information

```{r}
#| label: session
sessionInfo()
```
