---
title: "\\textbf{Lab 5: Simulating Confounding, Colliders, and DAGs}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 24"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

The point of this lab is that **you already know the right answer**. We build
worlds where the causal effect is a number we chose, then see which analyses
recover it and which do not. That is the only way to develop reliable intuition
about identification.

1.  Simulate confounding and watch adjustment fix it.
2.  Simulate a collider and watch adjustment *break* a correct answer.
3.  Simulate post-treatment (mediator) control.
4.  See selection bias as collider bias.
5.  Draw and query DAGs in `R` with `dagitty`.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(dagitty)
library(ggdag)
set.seed(723)
```

# Confounding

The world: $X$ causes both $D$ and $Y$; the true effect of $D$ on $Y$ is
$\tau = 2$.

```{r}
#| label: confound-dgp
sim_confound <- function(n = 2000, tau = 2) {
  x <- rnorm(n)                       # confounder
  d <- 1.2 * x + rnorm(n)             # treatment
  y <- tau * d + 1.5 * x + rnorm(n)   # outcome
  data.frame(y, d, x)
}
dat <- sim_confound()

c(naive    = coef(lm(y ~ d, data = dat))["d"],
  adjusted = coef(lm(y ~ d + x, data = dat))["d"],
  truth    = 2)
```

The naive estimate is badly biased upward. Check the omitted variable bias
formula from Week 1:

```{r}
#| label: confound-ovb
b_x   <- coef(lm(y ~ d + x, data = dat))["x"]   # effect of the omitted variable
delta <- coef(lm(x ~ d, data = dat))["d"]       # omitted on included
c(predicted_bias = b_x * delta,
  actual_bias = coef(lm(y ~ d, data = dat))["d"] - coef(lm(y ~ d + x, data = dat))["d"])
```

Exactly as predicted. Over many replications the adjusted estimator is centred
on the truth and the naive one is not:

```{r}
#| label: fig-confound
#| fig-cap: "1,000 replications. Adjustment recovers the true effect of 2; the naive estimator does not."
reps <- t(replicate(1000, {
  d0 <- sim_confound()
  c(naive = coef(lm(y ~ d, d0))["d"], adjusted = coef(lm(y ~ d + x, d0))["d"])
}))

as.data.frame(reps) |>
  tidyr::pivot_longer(everything()) |>
  ggplot(aes(value, fill = name)) +
  geom_histogram(bins = 60, alpha = .7, position = "identity") +
  geom_vline(xintercept = 2, colour = "black", linetype = 2) +
  scale_fill_manual(values = c("naive.d" = "firebrick", "adjusted.d" = "navy")) +
  labs(x = expression(hat(tau)), y = "count", fill = NULL) +
  theme_minimal(base_size = 10)
```

# Colliders

Now the world has **no confounding at all**: $D$ and $Y$ share no common cause.
But both cause $C$.

```{r}
#| label: collider-dgp
sim_collider <- function(n = 2000, tau = 2) {
  d <- rnorm(n)
  y <- tau * d + rnorm(n)
  c_ <- 1.0 * d + 1.0 * y + rnorm(n)     # collider: caused by BOTH
  data.frame(y, d, c = c_)
}
dat2 <- sim_collider()

c(correct   = coef(lm(y ~ d, data = dat2))["d"],
  controlled = coef(lm(y ~ d + c, data = dat2))["d"],
  truth     = 2)
```

::: callout-important
Read that table again. The **unadjusted** regression is right; adding a control
makes it wrong. Nothing about the data would tell you this. Only the DAG does.
:::

```{r}
#| label: fig-collider
#| fig-cap: "Controlling for a collider introduces bias where none existed."
reps2 <- t(replicate(1000, {
  d0 <- sim_collider()
  c(correct = coef(lm(y ~ d, d0))["d"], controlled = coef(lm(y ~ d + c, d0))["d"])
}))

as.data.frame(reps2) |>
  tidyr::pivot_longer(everything()) |>
  ggplot(aes(value, fill = name)) +
  geom_histogram(bins = 60, alpha = .7, position = "identity") +
  geom_vline(xintercept = 2, colour = "black", linetype = 2) +
  scale_fill_manual(values = c("correct.d" = "navy", "controlled.d" = "firebrick")) +
  labs(x = expression(hat(tau)), y = "count", fill = NULL) +
  theme_minimal(base_size = 10)
```

## Berkson's paradox, directly

Two genuinely independent traits become correlated once you condition on their
common effect.

```{r}
#| label: berkson
n <- 5000
test <- rnorm(n); athl <- rnorm(n)
admit <- (test + athl) > 1.5          # admitted if either is high enough

c(all_applicants = cor(test, athl),
  admitted_only  = cor(test[admit], athl[admit]))
```

```{r}
#| label: fig-berkson
#| fig-cap: "Test score and athletic talent are independent among applicants and negatively related among admitted students."
tibble::tibble(test, athl, admit) |>
  ggplot(aes(test, athl, colour = admit)) +
  geom_point(alpha = .25, size = .5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = .7) +
  scale_colour_manual(values = c("grey70", "navy")) +
  labs(colour = "admitted") +
  theme_minimal(base_size = 10)
```

# Post-treatment control

$D$ affects $Y$ both directly and through a mediator $M$. Controlling for $M$
answers a different question.

```{r}
#| label: mediator
sim_med <- function(n = 2000) {
  d <- rnorm(n)
  m <- 0.8 * d + rnorm(n)              # mediator
  y <- 1.0 * d + 1.5 * m + rnorm(n)    # total effect = 1.0 + 0.8*1.5 = 2.2
  data.frame(y, d, m)
}
dat3 <- sim_med()

c(total_effect   = coef(lm(y ~ d, data = dat3))["d"],
  direct_effect  = coef(lm(y ~ d + m, data = dat3))["d"],
  truth_total    = 2.2,
  truth_direct   = 1.0)
```

Both numbers are "correct", they estimate *different estimands*. The mistake
is not computing the second; it is reporting it as though it were the first.

::: callout-warning
This is the single most common error in applied sociology: controlling for
occupation when studying the effect of education on earnings, controlling for
income when studying the effect of education on health, controlling for marriage
when studying the effect of a policy on fertility.
:::

# Selection bias

Now the sample itself is selected on a collider. Only people who are employed
appear in the data, and employment depends on both the treatment and the
unobserved determinants of wages.

```{r}
#| label: selection
sim_select <- function(n = 5000, tau = 1) {
  educ <- rnorm(n)
  u    <- rnorm(n)                        # unobserved ability
  wage <- tau * educ + u + rnorm(n)
  employed <- (0.8 * educ + 1.2 * u + rnorm(n)) > 0.5
  data.frame(wage, educ, employed)
}
d4 <- sim_select()

c(full_population = coef(lm(wage ~ educ, data = d4))["educ"],
  employed_only   = coef(lm(wage ~ educ, data = d4, subset = employed))["educ"],
  truth = 1)
```

The regression on the observed (employed) sample is biased **downward**, even
though the specification is exactly the one that was unbiased in the full
population. Restricting to the employed opens a path through the collider, so
unobserved ability becomes correlated with education *within the sample*. The
sample restriction *is* the conditioning.

# DAGs in `R`

`dagitty` lets you state a graph and ask it what to condition on.

```{r}
#| label: dagitty-basic
g <- dagitty('dag {
  U [unobserved]
  X -> D
  X -> Y
  U -> X
  U -> C
  D -> M
  M -> Y
  Y -> C
  D [exposure]
  Y [outcome]
}')

adjustmentSets(g, exposure = "D", outcome = "Y", effect = "total")
```

```{r}
#| label: fig-dag
#| fig-cap: "The DAG from lecture."
#| fig-height: 3.2
ggdag(tidy_dagitty(g, layout = "sugiyama"), text_size = 3,
      node_size = 12) +
  theme_dag_blank()
```

Ask it directly whether a candidate adjustment set works:

```{r}
#| label: dagitty-queries
isAdjustmentSet(g, "X", exposure = "D", outcome = "Y")
isAdjustmentSet(g, c("X", "C"), exposure = "D", outcome = "Y")   # collider!
isAdjustmentSet(g, c("X", "M"), exposure = "D", outcome = "Y")   # mediator!
```

And which paths exist, open or blocked:

```{r}
#| label: paths
paths(g, from = "D", to = "Y")$paths |> head(6)
paths(g, from = "D", to = "Y", Z = "X")$open |> head(6)
```

## Testable implications

A DAG implies conditional independencies. Those *are* testable, and they are the
only part of a DAG that data can speak to.

```{r}
#| label: implications
impliedConditionalIndependencies(g) |> head(5)
```

```{r}
#| label: test-implications
## simulate from the DAG and check
sim <- simulateSEM(g, N = 5000)
localTests(g, sim, type = "cis.loess", R = 50) |> round(3) |> head(5)
```

::: callout-note
A DAG that passes its testable implications is not thereby "correct", many
different DAGs imply the same independencies. But a DAG that *fails* them is
wrong, and that is worth knowing.
:::

# Exercises

1.  **Bad control that predicts treatment.** Simulate $Z \to D \to Y$ with $Z$
    having no effect on $Y$, plus an unobserved confounder $U \to D$, $U \to Y$.
    Compare the bias in $\hat\tau$ with and without controlling for $Z$. You
    should find that controlling for $Z$ makes the bias *worse*; this is
    **bias amplification**. Explain why using the FWL logic from Week 1.

2.  **Positivity.** Modify `sim_confound()` so that everyone with $x > 1$ is
    treated with probability $0.99$. Estimate the ATE with and without those
    observations. What does regression do in the region with no comparison
    cases?

3.  **Heterogeneous effects.** Make $\tau_i$ depend on $x$: $\tau_i = 1 + x_i$.
    Compute the true ATE and the true ATT by simulation, then see what
    `lm(y ~ d + x)` reports. Connect your answer to the variance-weighting
    result from Week 1.

4.  **Your own DAG.** Draw the DAG for a causal question in your own research in
    `dagitty`. Mark unobserved variables. Ask for `adjustmentSets()`. Is there
    any set you can actually measure? If not, that is a finding, and it points
    you toward Weeks 7--11.

5.  **The birth-weight paradox.** Simulate: smoking $\to$ birth weight, an
    unobserved severe condition $\to$ birth weight and $\to$ infant mortality,
    smoking $\to$ mortality (a genuine harmful effect). Show that among
    low-birth-weight infants, smoking appears protective.

6.  **State an estimand.** For the GSS earnings data, write one sentence in the
    Lundberg--Johnson--Stewart form: a unit-specific quantity and a target
    population. Then say what would have to be true for
    `lm(lnearn ~ educ + ...)` to estimate it.

# Session information

```{r}
#| label: session
sessionInfo()
```
