---
title: "\\textbf{Lab 5: Potential Outcomes, Confounding, Colliders, and DAGs}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 24"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    toccolor: blue
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \usepackage{xcolor}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta,positioning}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\indep}{\perp\!\!\!\perp}
        \usepackage{setspace}
        \setstretch{1.12}
        %% the DAG style from the lecture slides, verbatim
        \definecolor{dukeblue}{RGB}{1,33,105}
        \definecolor{accent}{RGB}{200,78,0}
        \tikzset{
          dag/.style={>=Stealth,thick,
            every node/.style={circle,draw=dukeblue,fill=white,inner sep=1pt,
                               minimum size=9mm,font=\small}},
          unobs/.style={draw=gray,dashed},
          cond/.style={draw=accent,fill=accent!12},
          lab/.style={draw=none,fill=none,rectangle,inner sep=2pt,font=\footnotesize},
          arr/.style={->,dukeblue,thick},
          uarr/.style={->,gray,thick}}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

The point of this lab is that **you already know the right answer**. We build
worlds where every unit has both potential outcomes, so the ATE is a number we
chose, then see which analyses recover it and which do not.

1.  Build a population with both potential outcomes; compute the ATE, ATT,
    ATU; verify the selection-bias decomposition.
2.  Randomize, then apply the adjustment formula by hand, and watch both work.
3.  Simulate confounding, colliders, mediators, bias amplification, and
    selection; run Baron and Kenny's mechanism test on the GSS and break it.
4.  Draw and query DAGs: `dagitty` for the logic, TikZ for the picture.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(dagitty)
library(ggdag)
library(mediation)  # mediate, medsens
set.seed(723)
```

# Potential outcomes on a population you built

## Both columns, then the switching equation

Two thousand people. $X_i = 1$ if the parents went to college. The untreated
outcome is higher for $X_i = 1$, and the treatment effect is *lower* for
$X_i = 1$: the negative-selection pattern of Brand and Xie (2010) from lecture, built in.

```{r}
#| label: population
n <- 2000
X   <- rbinom(n, 1, 0.5)
Y0  <- 10 + 4 * X + rnorm(n, 0, 2)            ## baseline: 4 higher if X = 1
tau <- 6 - 4 * X + rnorm(n, 0, 1)             ## effect: 6 if X = 0, 2 if X = 1
Y1  <- Y0 + tau
pop <- data.frame(X, Y0, Y1, tau)
head(pop, 4)

## the estimands are averages of a column nobody ever sees in full
ATE <- mean(pop$tau)
ATE
```

Now assign treatment with selection: $X_i = 1$ is treated with probability
$0.7$, $X_i = 0$ with probability $0.3$. The switching equation reveals one
column per person.

```{r}
#| label: switching
pop$D <- rbinom(n, 1, ifelse(pop$X == 1, 0.7, 0.3))
pop$Y <- pop$D * pop$Y1 + (1 - pop$D) * pop$Y0     ## the switching equation

## ATT and ATU exist because WE still see both columns; a researcher does not
ATT <- mean(pop$tau[pop$D == 1]); ATU <- mean(pop$tau[pop$D == 0])
pi  <- mean(pop$D)
c(ATE = ATE, ATT = ATT, ATU = ATU, check = pi * ATT + (1 - pi) * ATU)
```

The naive comparison and its decomposition from lecture, term by term:

```{r}
#| label: decomposition
naive    <- mean(pop$Y[pop$D == 1]) - mean(pop$Y[pop$D == 0])
baseline <- mean(pop$Y0[pop$D == 1]) - mean(pop$Y0[pop$D == 0])   ## E[Y0|D=1] - E[Y0|D=0]
c(naive = naive, ATT_plus_baseline = ATT + baseline,
  ATE_plus_both = ATE + baseline + (1 - pi) * (ATT - ATU))
```

Both identities hold to the last decimal. The treated have a higher baseline
(they are mostly $X_i = 1$) and a smaller effect, and the naive comparison
adds the first error to the ATT.

```{r}
#| label: naive-is-ols
## regression of Y on D returns the naive comparison exactly; next week opens
## with why (the error term holds the baseline and the excess gain)
coef(lm(Y ~ D, data = pop))["D"]
```

## Randomization

Same population, treatment by coin flip. Now $D$ is independent of both
columns, the baseline term is zero on average, and the difference in means
is centred on the ATE.

```{r}
#| label: randomize
pop$Dr <- rbinom(n, 1, 0.5)
pop$Yr <- pop$Dr * pop$Y1 + (1 - pop$Dr) * pop$Y0
c(estimate = mean(pop$Yr[pop$Dr == 1]) - mean(pop$Yr[pop$Dr == 0]), ATE = ATE,
  baseline_gap = mean(pop$Y0[pop$Dr == 1]) - mean(pop$Y0[pop$Dr == 0]))
```

The estimate misses the ATE by sampling noise only; the baseline gap is
close to zero because a coin does not read $Y_i(0)$. The figure at the end of
this section repeats the draw 1,000 times.

## Selection on observables: the adjustment formula by hand

Back to the selected treatment `D`. Within each value of $X$, treatment was a
coin flip (probability $0.7$ or $0.3$, unrelated to the potential outcomes),
so conditional independence holds by construction. The lecture's formula:
difference within each stratum, then weight by the share of the stratum.

```{r}
#| label: adjustment-formula
cell <- pop |>
  group_by(X) |>
  summarise(share = n() / nrow(pop),
            diff  = mean(Y[D == 1]) - mean(Y[D == 0]),   ## CATE(x), estimable
            true_cate = mean(tau), .groups = "drop")
cell
c(adjusted = sum(cell$diff * cell$share), ATE = ATE)

## the regression version. It agrees with the formula up to sampling noise
## because Var(D | X) is the same in both strata (0.7 * 0.3 = 0.3 * 0.7), so
## OLS weights the two CATEs by the stratum shares. Week 6 shows what happens
## when the variances differ (exercise 3)
coef(lm(Y ~ D + X, data = pop))["D"]
```

Repeat the whole thing 1,000 times with fresh treatment draws, for all three
designs.

```{r}
#| label: fig-three-designs
#| fig-cap: "One population, treatment redrawn 1,000 times. The naive comparison under selection is centred on the wrong number; randomization and the adjustment formula are centred on the ATE (dashed). All three have about the same spread: bias is not imprecision."
#| fig-height: 2.6
three <- t(replicate(1000, {
  D <- rbinom(n, 1, ifelse(pop$X == 1, 0.7, 0.3)); Y <- D * pop$Y1 + (1 - D) * pop$Y0
  Dr <- rbinom(n, 1, 0.5); Yr <- Dr * pop$Y1 + (1 - Dr) * pop$Y0
  cell <- sapply(0:1, function(x) mean(Y[D == 1 & pop$X == x]) - mean(Y[D == 0 & pop$X == x]))
  c(`selection, naive` = mean(Y[D == 1]) - mean(Y[D == 0]),
    `randomized, naive` = mean(Yr[Dr == 1]) - mean(Yr[Dr == 0]),
    `selection, adjusted` = sum(cell * table(pop$X) / n))
}))
as.data.frame(three) |>
  tidyr::pivot_longer(everything(), names_to = "design") |>
  mutate(design = factor(design, levels = colnames(three))) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "grey75") +
  geom_vline(xintercept = ATE, linetype = 2) +
  facet_wrap(~ design) +
  labs(x = "estimate of the ATE", y = NULL) +
  theme_minimal(base_size = 10)
round(rbind(mean = colMeans(three), sd = apply(three, 2, sd)), 3)
```

# Confounding

The world: $X$ causes both $D$ and $Y$; the true effect of $D$ on $Y$ is
$\tau = 2$. This is the fork of the lecture's three building blocks, and the
back-door path $D \leftarrow X \to Y$ is the only one.

```{r}
#| label: confound-dgp
sim_confound <- function(n = 2000, tau = 2) {
  x <- rnorm(n)                       ## confounder
  d <- 1.2 * x + rnorm(n)             ## treatment
  y <- tau * d + 1.5 * x + rnorm(n)   ## outcome
  data.frame(y, d, x)
}
dat <- sim_confound()

c(naive    = coef(lm(y ~ d, data = dat))["d"],
  adjusted = coef(lm(y ~ d + x, data = dat))["d"],
  truth    = 2)
```

The naive estimate is biased upward. Check the omitted variable bias formula
from Week 1:

```{r}
#| label: confound-ovb
b_x   <- coef(lm(y ~ d + x, data = dat))["x"]   ## effect of the omitted variable
delta <- coef(lm(x ~ d, data = dat))["d"]       ## omitted on included
c(predicted_bias = b_x * delta,
  actual_bias = coef(lm(y ~ d, data = dat))["d"] - coef(lm(y ~ d + x, data = dat))["d"])
```

# Colliders

Now the world has **no confounding at all**: $D$ and $Y$ share no common cause.
But both cause $C$.

```{r}
#| label: collider-dgp
sim_collider <- function(n = 2000, tau = 2) {
  d <- rnorm(n)
  y <- tau * d + rnorm(n)
  c_ <- 1.0 * d + 1.0 * y + rnorm(n)     ## collider: caused by BOTH
  data.frame(y, d, c = c_)
}
dat2 <- sim_collider()

c(correct    = coef(lm(y ~ d, data = dat2))["d"],
  controlled = coef(lm(y ~ d + c, data = dat2))["d"],
  truth      = 2)
```

::: callout-important
Read that table again. The **unadjusted** regression is right; adding a control
makes it wrong. Nothing about the data would tell you this. Only the DAG does.
:::

```{r}
#| label: fig-collider
#| fig-cap: "Controlling for a collider introduces bias where none existed."
reps2 <- t(replicate(1000, {
  d0 <- sim_collider()
  c(correct = coef(lm(y ~ d, d0))["d"], controlled = coef(lm(y ~ d + c, d0))["d"])
}))

as.data.frame(reps2) |>
  tidyr::pivot_longer(everything()) |>
  ggplot(aes(value, fill = name)) +
  geom_histogram(bins = 60, alpha = .7, position = "identity") +
  geom_vline(xintercept = 2, colour = "black", linetype = 2) +
  scale_fill_manual(values = c("correct.d" = "navy", "controlled.d" = "firebrick")) +
  labs(x = expression(hat(tau)), y = "count", fill = NULL) +
  theme_minimal(base_size = 10)
```

## Berkson's paradox, directly

Two genuinely independent traits become correlated once you condition on their
common effect. This is the admissions example from lecture.

```{r}
#| label: berkson
n_b <- 5000
test <- rnorm(n_b); athl <- rnorm(n_b)
admit <- (test + athl) > 1.5          ## admitted if either is high enough

c(all_applicants = cor(test, athl),
  admitted_only  = cor(test[admit], athl[admit]))
```

```{r}
#| label: fig-berkson
#| fig-cap: "Test score and athletic talent are independent among applicants and negatively related among admitted students."
tibble::tibble(test, athl, admit) |>
  ggplot(aes(test, athl, colour = admit)) +
  geom_point(alpha = .25, size = .5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = .7) +
  scale_colour_manual(values = c("grey70", "navy")) +
  labs(colour = "admitted") +
  theme_minimal(base_size = 10)
```

# Post-treatment control

$D$ affects $Y$ both directly and through a mediator $M$. Controlling for $M$
answers a different question.

```{r}
#| label: mediator
sim_med <- function(n = 2000) {
  d <- rnorm(n)
  m <- 0.8 * d + rnorm(n)              ## mediator
  y <- 1.0 * d + 1.5 * m + rnorm(n)    ## total effect = 1.0 + 0.8*1.5 = 2.2
  data.frame(y, d, m)
}
dat3 <- sim_med()

c(total_effect   = coef(lm(y ~ d, data = dat3))["d"],
  direct_effect  = coef(lm(y ~ d + m, data = dat3))["d"],
  truth_total    = 2.2,
  truth_direct   = 1.0)
```

Both numbers are "correct", they estimate *different estimands*. The mistake
is not computing the second; it is reporting it as though it were the first.

::: callout-warning
This is the single most common error in applied sociology: controlling for
occupation when studying the effect of education on earnings, controlling for
income when studying the effect of education on health, controlling for marriage
when studying the effect of a policy on fertility.
:::

# The mechanism test: Baron and Kenny (1986)

The lecture's three regressions, on the mediator world above. The truth is
$a = 0.8$, $b = 1.5$, $c' = 1$, so the indirect effect is $ab = 1.2$ and the
total effect is $c = 2.2$.

```{r}
#| label: baron-kenny
bk <- function(dat) {
  m1 <- lm(y ~ d, data = dat)          ## step 1: total effect c
  m2 <- lm(m ~ d, data = dat)          ## step 2: D moves M, coefficient a
  m3 <- lm(y ~ d + m, data = dat)      ## step 3: b for M, c' for D
  a <- coef(m2)["d"]; b <- coef(m3)["m"]
  ## Sobel SE for ab: delta method, a and b from different regressions so
  ## their covariance is taken as zero
  se_ab <- sqrt(b^2 * vcov(m2)["d", "d"] + a^2 * vcov(m3)["m", "m"])
  c(c = coef(m1)["d"], a = a, b = b, c_prime = coef(m3)["d"],
    ab = a * b, se_ab = se_ab, share = a * b / coef(m1)["d"])
}
round(bk(dat3), 3)
## the identity c = c' + ab holds exactly for OLS: it is the omitted-variable
## formula with M as the omitted variable
with(as.list(bk(dat3)), c(c = c.d, c_prime_plus_ab = c_prime.d + ab.d))
```

Now the same test on real data: parents' education, own education, log
earnings, with sex, race, and age as covariates in every equation.

```{r}
#| label: bk-gss
gss <- readRDS("../Data/gss_earnings.rds")
c_tot <- coef(lm(lnearn ~ pareduc + female + black + age, data = gss))["pareduc"]
a     <- coef(lm(educ   ~ pareduc + female + black + age, data = gss))["pareduc"]
f     <- coef(lm(lnearn ~ pareduc + educ + female + black + age, data = gss))
round(c(c = c_tot, a = a, b = f["educ"], c_prime = f["pareduc"],
        ab = a * f["educ"], share = a * f["educ"] / c_tot), 4)
```

A paper would write: "63 percent of the effect of parents' education on
earnings runs through the respondent's own schooling." Step 3 is a regression
of earnings on schooling, and the lecture's back-door question applies to it:
is there an unmeasured cause of both own education and earnings? Ability is
one. Nothing in the three regressions can detect it. Build it in and watch:

```{r}
#| label: bk-broken
sim_med_u <- function(n = 20000) {
  d <- rnorm(n); u <- rnorm(n)             ## u: unmeasured, into M and Y
  m <- 0.8 * d + u + rnorm(n)
  y <- 1.0 * d + 1.5 * m + u + rnorm(n)    ## same a, b, c' as before
  data.frame(y, d, m)
}
round(rbind(truth = c(c = 2.2, a = 0.8, b = 1.5, c_prime = 1, ab = 1.2, se_ab = NA, share = 1.2 / 2.2),
            baron_kenny = bk(sim_med_u())), 3)
```

The total effect $c$ and the first-stage $a$ survive, because $D$ has no back
door. The mediator coefficient $b$ absorbs $U$, so the indirect effect is
overstated, the direct effect understated, and the "share mediated" moves
from 55 to 73 percent. The identity $c = c' + ab$ still holds; only the split
is wrong, and the split was the point. Modern mediation analysis names the
assumption sequential ignorability (Imai, Keele, and Tingley 2010) and pairs
the estimate with a sensitivity analysis for $U$.

## The same test in a package, plus the sensitivity analysis

The `mediation` package takes the mediator model and the outcome model and
returns the indirect effect (it calls it the ACME, average causal mediation
effect), the direct effect (ADE), and the share mediated, with simulation
intervals. With linear models and no interaction the ACME *is* $ab$ and the
ADE *is* $c'$, so this is Baron and Kenny with better intervals. What the
package adds is `medsens()`: how strong would the unmeasured $U$ have to be
to erase the indirect effect?

```{r}
#| label: mediation-pkg
m_med <- lm(educ   ~ pareduc + female + black + age, data = gss)          ## step 2
m_out <- lm(lnearn ~ pareduc + educ + female + black + age, data = gss)   ## step 3
## sims: draws of the coefficients for the intervals (Week 3's parametric bootstrap)
med <- mediate(m_med, m_out, treat = "pareduc", mediator = "educ", sims = 500)
summary(med)

## sensitivity: rho is the correlation between the errors of the mediator and
## outcome equations that an unmeasured U would induce. At what rho is ACME 0?
sens <- medsens(med, rho.by = 0.1, sims = 200)
summary(sens)
```

The ACME of $0.039$ and ADE of $0.023$ are the $ab$ and $c'$ computed by
hand. The sensitivity table says the indirect effect reaches zero when the
error correlation is $0.3$, or when an unmeasured $U$ explains about 9
percent of the variance in both schooling and earnings. Whether ability is
that strong is the substantive question; the package puts a number on it
where Baron and Kenny said nothing.

```{r}
#| label: fig-medsens
#| fig-cap: "Sensitivity of the indirect effect to an unmeasured confounder of schooling and earnings."
#| fig-height: 3
plot(sens, main = "", ylim = c(-0.02, 0.06))
```

# Bias amplification

The lecture's third single-cause case: $Z$ causes $D$ only, and an unmeasured
$U$ confounds $D$ and $Y$. $Z$ is not on any back-door path, so controlling
for it cannot remove bias. It makes the bias worse.

```{r}
#| label: amplification
sim_amp <- function(n = 5000, tau = 1) {
  z <- rnorm(n); u <- rnorm(n)              ## z: cause of D only; u: unmeasured confounder
  d <- 1.5 * z + u + rnorm(n)
  y <- tau * d + 2 * u + rnorm(n)
  data.frame(y, d, z)
}
dat5 <- sim_amp()
c(without_z = coef(lm(y ~ d, data = dat5))["d"],
  with_z    = coef(lm(y ~ d + z, data = dat5))["d"],
  truth     = 1)
```

Why: by FWL (Week 1), controlling for $Z$ regresses $Y$ on the part of $D$
that $Z$ does not explain. That residual is $U + \epsilon_D$, in which $U$ is
a *larger* share than it was in $D$. The clean variation from $Z$ was the
part of $D$ that was safe to use, and the control threw it away. Variables
like $Z$ are instruments (Week 7).

# Selection bias

Now the sample itself is selected on a collider. Only people who are employed
appear in the data, and employment depends on both the treatment and the
unobserved determinants of wages.

```{r}
#| label: selection
sim_select <- function(n = 5000, tau = 1) {
  educ <- rnorm(n)
  u    <- rnorm(n)                        ## unobserved ability
  wage <- tau * educ + u + rnorm(n)
  employed <- (0.8 * educ + 1.2 * u + rnorm(n)) > 0.5
  data.frame(wage, educ, employed)
}
d4 <- sim_select()

c(full_population = coef(lm(wage ~ educ, data = d4))["educ"],
  employed_only   = coef(lm(wage ~ educ, data = d4, subset = employed))["educ"],
  truth = 1)
```

The regression on the employed is biased **downward**, even though the
specification is the one that was unbiased in the full population. Restricting
to the employed conditions on a collider, so unobserved ability becomes
correlated with education *within the sample*. The sample restriction *is* the
conditioning.

::: callout-important
`gss_earnings.rds` is built this way: `build_gss_extract.R` keeps respondents
working full time with positive earnings. Every earnings regression in this
course runs on a sample selected on a descendant of schooling and of the
unobserved determinants of earnings. Say so when you report one.
:::

# DAGs in `R`

`dagitty` lets you state a graph and ask it what to condition on. This is the
lecture's worked example: degree $D$, earnings $Y$, family background $X$, a
measured test score $A$, unmeasured school quality $U$, and occupation $M$ as
one channel from degree to earnings.

```{r}
#| label: dagitty-basic
g <- dagitty('dag {
  U [unobserved]
  X -> D   X -> Y
  A -> D   A -> Y
  U -> A   U -> Y
  D -> M   M -> Y
  D -> Y
  D [exposure]  Y [outcome]
}')

adjustmentSets(g, exposure = "D", outcome = "Y", effect = "total")
```

```{r}
#| label: fig-dag
#| fig-cap: "The lecture's college-and-earnings DAG, drawn by ggdag."
#| fig-height: 3.2
ggdag(tidy_dagitty(g, layout = "sugiyama"), text_size = 3, node_size = 12) +
  theme_dag_blank()
```

Ask it directly whether a candidate adjustment set works, and why not:

```{r}
#| label: dagitty-queries
isAdjustmentSet(g, c("X", "A"), exposure = "D", outcome = "Y")
isAdjustmentSet(g, "X", exposure = "D", outcome = "Y")             ## A's back door open
isAdjustmentSet(g, c("X", "A", "M"), exposure = "D", outcome = "Y") ## mediator
```

Every path, and which of them $\{X, A\}$ leaves open:

```{r}
#| label: paths
paths(g, from = "D", to = "Y")$paths
paths(g, from = "D", to = "Y", Z = c("X", "A"))$open   ## TRUE = still open
```

Only the two causal paths, $D \to Y$ and $D \to M \to Y$, stay open under
$\{X, A\}$: the back-door criterion, checked by machine. `U` needed no
adjustment because every path through it passes through `A`, a chain node
that `A` blocks.

## Testable implications

A DAG implies conditional independencies among the *observed* variables.
Those are testable, and they are the only part of a DAG that data can speak
to.

```{r}
#| label: implications
impliedConditionalIndependencies(g)
```

```{r}
#| label: test-implications
## simulate from the DAG and check
sim <- simulateSEM(g, N = 5000)
localTests(g, sim, type = "cis.loess", R = 50) |> round(3)
```

::: callout-note
A DAG that passes its testable implications is not thereby "correct": many
different DAGs imply the same independencies. A DAG that *fails* them is wrong,
and that is worth knowing. Note what is absent from the list: nothing about
$Y(d)$. Identification assumptions are not among the testable implications.
:::

# Drawing DAGs

`ggdag` is fine for a quick look. For a paper or a slide you want control over
placement, and the standard tool is TikZ. Every DAG in this week's slides is
drawn with one style block and a handful of commands, and the same code works
in a Quarto PDF document. The style, placed once in the preamble (here, in the
YAML `include-in-header`), defines the node shape, the arrow, and three
variants:

```latex
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}
\tikzset{
  dag/.style={>=Stealth,thick,
    every node/.style={circle,draw=dukeblue,fill=white,inner sep=1pt,
                       minimum size=9mm,font=\small}},
  unobs/.style={draw=gray,dashed},       % unmeasured variable
  cond/.style={draw=accent,fill=accent!12}, % a variable conditioned on
  lab/.style={draw=none,fill=none,rectangle,inner sep=2pt,font=\footnotesize},
  arr/.style={->,dukeblue,thick},
  uarr/.style={->,gray,thick}}            % arrow from an unmeasured variable
```

A picture is then a list of nodes at coordinates, and a list of arrows
between them. This is the college-and-earnings DAG, exactly as it appears in
the slides:

```latex
\begin{tikzpicture}[dag]
  \node (D) at (0,0)      {$D$};
  \node (M) at (2.6,0.9)  {$M$};
  \node (Y) at (5.2,0)    {$Y$};
  \node (X) at (1.0,-2.0) {$X$};
  \node (A) at (4.2,-2.0) {$A$};
  \node[unobs] (U) at (6.2,-3.4) {$U$};
  \draw[arr] (D)--(M);  \draw[arr] (M)--(Y);  \draw[arr] (D)--(Y);
  \draw[arr] (X)--(D);  \draw[arr] (X)--(Y);
  \draw[arr] (A)--(D);  \draw[arr] (A)--(Y);
  \draw[uarr] (U)--(A); \draw[uarr] (U)--(Y);
\end{tikzpicture}
```

In a Quarto document, put that code inside a raw LaTeX block, which starts
with ```` ```{=latex} ```` and is passed straight to the PDF engine. The
result:

```{=latex}
\begin{center}
\begin{tikzpicture}[dag]
  \node (D) at (0,0)      {$D$};
  \node (M) at (2.6,0.9)  {$M$};
  \node (Y) at (5.2,0)    {$Y$};
  \node (X) at (1.0,-2.0) {$X$};
  \node (A) at (4.2,-2.0) {$A$};
  \node[unobs] (U) at (6.2,-3.4) {$U$};
  \draw[arr] (D)--(M);  \draw[arr] (M)--(Y);  \draw[arr] (D)--(Y);
  \draw[arr] (X)--(D);  \draw[arr] (X)--(Y);
  \draw[arr] (A)--(D);  \draw[arr] (A)--(Y);
  \draw[uarr] (U)--(A); \draw[uarr] (U)--(Y);
\end{tikzpicture}
\end{center}
```

Three more things you will want. A conditioned node uses `\node[cond]`; a
curved arrow uses `to[bend left=30]` in place of `--`; a text label uses
`\node[lab]`. The collider example from lecture uses all three:

```latex
\begin{tikzpicture}[dag]
  \node (T) at (0,0)   {$T$};   \node (A) at (3.2,0) {$A$};
  \node[cond] (C) at (1.6,-1.8) {$C$};
  \draw[arr] (T)--(C); \draw[arr] (A)--(C);
  \draw[<->,accent,dashed,thick] (T) to[bend left=30] (A);
  \node[lab] at (0,0.8) {test score};  \node[lab] at (3.2,0.8) {athletic talent};
  \node[lab] at (1.6,-2.6) {admitted};
\end{tikzpicture}
```

```{=latex}
\begin{center}
\begin{tikzpicture}[dag]
  \node (T) at (0,0)   {$T$};   \node (A) at (3.2,0) {$A$};
  \node[cond] (C) at (1.6,-1.8) {$C$};
  \draw[arr] (T)--(C); \draw[arr] (A)--(C);
  \draw[<->,accent,dashed,thick] (T) to[bend left=30] (A);
  \node[lab] at (0,0.8) {test score};  \node[lab] at (3.2,0.8) {athletic talent};
  \node[lab] at (1.6,-2.6) {admitted};
\end{tikzpicture}
\end{center}
```

The same code goes into a beamer frame unchanged; add
`scale=0.85, every node/.append style={transform shape}` to the options to
shrink a large graph. The complete style is in the YAML header of this file
and in `slides.tex`.

# Exercises

1.  **A precision variable.** Simulate $Z \to Y$ with no arrow into $D$, plus
    $D \to Y$. Compare the estimate and its standard error with and without
    $Z$ over 500 replications. Which case of the lecture's single-cause frame
    is this, and does the DAG say to include $Z$?

2.  **Positivity.** Modify `sim_confound()` so that everyone with $x > 1$ is
    treated with probability $0.99$. Estimate the ATE with and without those
    observations. What does regression do in the region with no comparison
    cases?

3.  **Heterogeneous effects.** In Section 2, change the treatment
    probabilities to $0.9$ for $X_i = 1$ and $0.3$ for $X_i = 0$, so that
    $\Var(D_i \mid X_i)$ differs across strata. Recompute the adjustment
    formula and `lm(Y ~ D + X)`. They now disagree; say which one is the ATE
    and what the other one is. Week 6 derives the weights.

4.  **Your own DAG.** Draw the DAG for a causal question in your own research,
    first in `dagitty` and then in TikZ. Mark unobserved variables. Ask for
    `adjustmentSets()`. Is there any set you can actually measure? If not,
    that is a finding, and it points you toward Weeks 7 to 11.

5.  **The birth-weight paradox.** Simulate: smoking $\to$ birth weight, an
    unobserved severe condition $\to$ birth weight and $\to$ infant mortality,
    smoking $\to$ mortality (a genuine harmful effect). Show that among
    low-birth-weight infants, smoking appears protective. Then draw the DAG
    with birth weight as a conditioned node.

6.  **State an estimand.** For the GSS earnings data, write one sentence in
    the Lundberg, Johnson, and Stewart (2021) form: a unit-specific quantity
    and a target population. Then say what would have to be true for
    `lm(lnearn ~ ba + ...)` to estimate it. Remember that the extract keeps
    only full-time workers (Section 8).

# Session information

```{r}
#| label: session
sessionInfo()
```
