---
title: "\\textbf{Lab 4: Limited Dependent Variable Models}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 17"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Fit a logit, then read it on the probability scale: predicted probabilities
    first, average marginal effects second, odds ratios as a footnote.
2.  Handle an interaction in a nonlinear model: when to include the product,
    and how to report the interaction.
3.  Fit a multinomial logit on a nominal outcome and an ordered logit on an
    ordered one; check proportional odds.
4.  Fit count models: Poisson, the three overdispersion repairs, the hurdle
    model; compare them by AIC and BIC.
5.  Fit a random-intercept model, read the intraclass correlation and
    shrinkage, and compare it with fixed effects (Thursday's multilevel
    session).

```{r}
#| label: packages
library(lme4)      # lmer, glmer
library(dplyr)
library(ggplot2)
library(marginaleffects)
library(sandwich); library(lmtest)
library(MASS)      # glm.nb, polr
library(AER)       # dispersiontest
library(nnet)      # multinom
library(pscl)      # hurdle, zeroinfl
library(brglm2)    # Firth penalized likelihood
library(pROC)

gss <- readRDS("../Data/gss_earnings.rds")
```

# Binary outcomes

Outcome: `ba`, whether the respondent holds a bachelor's degree or more. The
model is Tuesday's: parents' education, vocabulary score, female, black.

```{r}
#| label: fit-binary
f <- ba ~ pareduc + wordsum + female + black

m_logit  <- glm(f, data = gss, family = binomial(link = "logit"))
m_probit <- glm(f, data = gss, family = binomial(link = "probit"))
m_lpm    <- lm(f, data = gss)

## the logit table from the slides: coefficient, SE, z, exp(coefficient)
round(cbind(coef(summary(m_logit))[, 1:3], exp_b = exp(coef(m_logit))), 3)
```

## Three scales, one coefficient

The `wordsum` coefficient is $0.391$ on the log-odds scale, $e^{0.391} = 1.48$
on the odds scale, and *not one number* on the probability scale. The first
two are exact and uninterpretable to most readers; the third is what the
reader wants and depends on where the person starts.

```{r}
#| label: three-scales
## one person at a time: log-odds, odds, probability for wordsum = 3, 6, 9
## with the other covariates at pareduc = 12, female = 0, black = 0
b  <- coef(m_logit)
xb <- b["(Intercept)"] + b["pareduc"] * 12 + b["wordsum"] * c(3, 6, 9)
round(cbind(wordsum = c(3, 6, 9), log_odds = xb, odds = exp(xb),
            prob = plogis(xb)), 3)
```

Each step of three words adds the same $3 \times 0.391 = 1.17$ to the
log-odds and multiplies the odds by the same $e^{1.17} = 3.23$; the probability
moves by a different amount each time.

## Start with predicted probabilities

How likely is a degree if your parents finished high school (12 years) rather
than college (16)? The observed-value approach: set `pareduc` to 12 for
*everyone*, keep every other covariate at its actual value, average the fitted
probabilities; repeat at 16.

```{r}
#| label: predprob
## avg_predictions() does exactly that loop, with a delta-method interval
avg_predictions(m_logit, variables = list(pareduc = c(8, 12, 16, 20)))
```

```{r}
#| label: contrast
## the contrast we report: high school to college, 18 points
avg_comparisons(m_logit, variables = list(pareduc = c(12, 16)))
```

```{r}
#| label: fig-predprob
#| fig-cap: "Average predicted probability of holding a BA by parents' education (observed-value approach), with 95% intervals."
plot_predictions(m_logit, by = "pareduc") +
  labs(x = "Parents' education (years)", y = "Pr(bachelor's or more)") +
  theme_minimal(base_size = 10)
```

Contrast with setting everything else to its mean, which describes one
hypothetical person (a 0.5 woman, a 0.13 Black respondent):

```{r}
#| label: predprob-mean
predictions(m_logit, newdata = datagrid(pareduc = c(12, 16)))
```

## Then marginal effects: one number per variable

```{r}
#| label: ame
## AME: each person's own slope dp/dx = beta * p(1 - p), averaged
avg_slopes(m_logit)
```

Compare with the marginal effect at the mean and with the LPM coefficients:

```{r}
#| label: ame-vs-mem
mem <- slopes(m_logit, newdata = "mean")
comparison <- data.frame(
  variable = c("pareduc", "wordsum", "female", "black"),
  AME  = avg_slopes(m_logit)$estimate,
  MEM  = mem$estimate,
  LPM  = coef(m_lpm)[-1]
)
round(comparison[, -1], 4) |> `rownames<-`(comparison$variable)
```

-   The AME and the LPM coefficients are close. This is typical, and it is the
    main reason people defend the LPM for descriptive and causal work.
-   The MEM answers a different question: the slope for one hypothetical person
    at the mean of everything, including the means of dummies.
-   Report the AME (Hanmer and Kalkan 2013).

## Odds ratios, at most as a footnote

```{r}
#| label: odds
round(exp(cbind(OR = coef(m_logit), confint.default(m_logit))), 3)
```

Each correct word multiplies the *odds* of a BA by 1.48. That is one number for
everyone, because the model is linear on the log-odds scale. It is not "48%
more likely": the probability change depends on where the person starts, which
is why the probabilities came first.

## Uncertainty: robust and clustered standard errors

The sandwich from Week 2 works on `glm` objects, and `marginaleffects` will
carry a chosen variance through to the AMEs.

```{r}
#| label: glm-robust
round(cbind(
  classical = sqrt(diag(vcov(m_logit))),
  HC0       = sqrt(diag(vcovHC(m_logit, type = "HC0"))),
  cluster   = sqrt(diag(vcovCL(m_logit, cluster = ~ psu)))
), 4)
```

```{r}
#| label: ame-robust
avg_slopes(m_logit, vcov = ~ psu)
```

## Probit: another S-curve

```{r}
#| label: probit
## coefficients differ by a scale factor; AMEs do not
round(cbind(logit = coef(m_logit), probit = coef(m_probit),
            ratio = coef(m_logit) / coef(m_probit)), 3)
round(cbind(AME_logit  = avg_slopes(m_logit)$estimate,
            AME_probit = avg_slopes(m_probit)$estimate), 4) |>
  `rownames<-`(avg_slopes(m_logit)$term)
```

The coefficients differ by a factor of about 1.6 to 1.7 because the two
S-curves have different steepness at the center. The AMEs agree to two
decimals. Default to logit; use probit when the next step needs a normal
(selection models, bivariate probit) or when the subfield expects it.

::: callout-important
Never compare raw logit or probit coefficients across nested models or across
groups. Adding a covariate rescales every coefficient, even ones whose effect is
unchanged (Mood 2010). Compare AMEs or predicted probabilities.
:::

# Interactions

Does parents' education matter more for women? Two questions: should the
product `pareduc * female` be in the model, and do the models with and
without it agree on the probability scale?

```{r}
#| label: interaction
## no product: one log-odds slope for pareduc, forced equal for men and women
m_np  <- glm(ba ~ pareduc + female + wordsum + black, data = gss, family = binomial)
## with product: the slope is beta_1 + beta_12 * female; the data choose beta_12
m_int <- glm(ba ~ pareduc * female + wordsum + black, data = gss, family = binomial)
round(coef(summary(m_int))[, 1:3], 3)
## LR test of the product
anova(m_np, m_int, test = "LRT")
```

The product is $0.021$ on the log-odds scale, $z = 0.79$. That is the wrong
scale for the question. The AME of parents' education, by gender, in both
models:

```{r}
#| label: interaction-me
## no product: the AMEs still differ by gender, through p(1 - p) alone
avg_slopes(m_np,  variables = "pareduc", by = "female")
## with product: they differ through p(1 - p) and through beta_12
avg_slopes(m_int, variables = "pareduc", by = "female")
```

```{r}
#| label: interaction-diff
## the interaction we report: the difference in AMEs, with an interval
avg_slopes(m_int, variables = "pareduc", by = "female",
           hypothesis = "b2 - b1 = 0")
```

-   **Include the product** whenever the question is whether $X_1$'s effect
    differs by $X_2$. Without it the answer is fixed in advance: zero on the
    log-odds scale, and on the probability scale whatever the S-curve gives
    (the no-product model already shows women with a larger AME, because
    they sit where the curve is steeper: Ai and Norton 2003).
-   **The two models agree** on probabilities only when $\hat\beta_{12}$ is
    near zero, as here. When it is not, the no-product model has the wrong
    probabilities in the cells.
-   **Report** the difference in AMEs with its interval, here $0.006$ per year,
    SE $0.004$. The answer is no.

# Separation

Construct a variable that perfectly predicts the outcome, and watch the MLE run
off to infinity.

```{r}
#| label: separation
gss2 <- gss |> mutate(perfect = as.integer(educ >= 18 & ba == 1))
m_sep <- glm(ba ~ pareduc + perfect, data = gss2, family = binomial)
round(coef(summary(m_sep))[, 1:2], 3)
```

An enormous coefficient with an enormous standard error is the signature. The
fix is Firth's penalized likelihood:

```{r}
#| label: firth
m_firth <- glm(ba ~ pareduc + perfect, data = gss2,
               family = binomial, method = "brglmFit")
round(coef(summary(m_firth))[, 1:2], 3)
```

Finite estimates, usable standard errors, and the variable is retained
(Zorn 2005).

# Outcomes with several categories

## Multinomial logit: a nominal outcome

Where a respondent lives, in four census regions. The categories have no
order, so one logit for every region against a base (the Midwest).

```{r}
#| label: region4
## collapse the nine GSS regions to four; Midwest first so it is the base
gss$region4 <- factor(case_when(
  gss$region %in% c("New England", "Middle Atlantic") ~ "Northeast",
  gss$region %in% c("E. North Central", "W. North Central") ~ "Midwest",
  gss$region %in% c("South Atlantic", "E. South Central", "W. South Central") ~ "South",
  TRUE ~ "West"), levels = c("Midwest", "Northeast", "South", "West"))
table(gss$region4)

m_mn <- multinom(region4 ~ educ + black + female + age, data = gss, trace = FALSE)
## one row per non-base region: exp(coef) multiplies the odds AGAINST THE MIDWEST
round(exp(coef(m_mn)), 2)
```

Read the `black` column: being Black multiplies the odds of South-versus-Midwest
by about 2.9 and of West-versus-Midwest by about 0.8. Those are odds against
the base, not probabilities; the four probabilities move together and add to
one:

```{r}
#| label: region4-pred
## observed-value predicted probabilities, one column per region
avg_predictions(m_mn, variables = "black")
## check: for each person the four probabilities sum to one
range(rowSums(fitted(m_mn)))
```

```{r}
#| label: region4-lr
## do the covariates matter at all? LR against the intercept-only model
m_mn0 <- multinom(region4 ~ 1, data = gss, trace = FALSE)
lr <- 2 * (as.numeric(logLik(m_mn)) - as.numeric(logLik(m_mn0)))
df <- length(coef(m_mn)) - length(coef(m_mn0))
c(LR = round(lr, 1), df = df)
pchisq(lr, df, lower.tail = FALSE) < 0.001
```

## Ordered logit: an ordered outcome

Highest degree, five ordered categories. One coefficient vector and four
cutpoints.

```{r}
#| label: ordered
m_ord <- polr(degree ~ pareduc + wordsum + female + black,
              data = gss, Hess = TRUE)
round(coef(summary(m_ord)), 4)
```

`polr()` does not print $p$-values; get them from the $t$-values, or better,
work on the probability scale.

```{r}
#| label: gates
## the cutpoints are gates: for one person, X'beta and the four cumulative
## probabilities Pr(Y <= j) = Lambda(kappa_j - X'beta); differences are the
## five category probabilities
b <- coef(m_ord); kappa <- m_ord$zeta
x  <- c(pareduc = 12, wordsum = 5, female = 0, black = 0)
xb <- sum(b * x)
cum <- plogis(kappa - xb)
round(c(xb = xb), 3)
round(rbind(cumulative = c(cum, 1), category = diff(c(0, cum, 1))), 3) |>
  `colnames<-`(levels(gss$degree))
```

```{r}
#| label: ordered-pred
## the same, averaged over everyone, at two vocabulary scores
avg_predictions(m_ord, variables = list(wordsum = c(3, 9)), type = "probs")
```

## Checking proportional odds

One $\beta$ for every cutpoint is an assumption: vocabulary pushes people past
the high-school line and the bachelor's line by the same log-odds. Check it
the direct way, with four separate binary logits, "above cut $j$" versus not.

```{r}
#| label: prop-odds
## one binary logit per cut; under proportional odds the four rows agree
cuts <- levels(gss$degree)[-5]
above <- sapply(1:4, function(k) {
  yk <- as.integer(as.integer(gss$degree) > k)   # 1 if above cut k
  coef(glm(yk ~ pareduc + wordsum + female + black, data = gss, family = binomial))[-1]
})
colnames(above) <- paste("above", cuts)
round(cbind(above, ordered = coef(m_ord)), 3)
```

The ordered model's `wordsum` coefficient is a compromise between the four;
proportional odds is a fair summary here.

```{r}
#| label: ordered-vs-multinom
## the formal check: a multinomial logit on degree frees all four rows
m_mn_deg <- multinom(degree ~ pareduc + wordsum + female + black, data = gss, trace = FALSE)
k_ord <- length(coef(m_ord)) + length(m_ord$zeta)   # 4 slopes + 4 cutpoints
k_mn  <- length(coef(m_mn_deg))                      # 4 rows x 5 columns
lr <- 2 * (as.numeric(logLik(m_mn_deg)) - as.numeric(logLik(m_ord)))
c(LR = round(lr, 1), df = k_mn - k_ord,
  p = round(pchisq(lr, k_mn - k_ord, lower.tail = FALSE), 3))

## the same comparison by AIC and BIC
rbind(ordered     = c(k = k_ord, AIC = AIC(m_ord),    BIC = BIC(m_ord)),
      multinomial = c(k = k_mn,  AIC = AIC(m_mn_deg), BIC = BIC(m_mn_deg)))
```

The LR rejects at $n = 3{,}509$, as everything does (Week 3). BIC charges
$12 \log 3509 = 98$ for the twelve extra parameters and gets back $27$, so it
keeps the ordered model; AIC charges $24$ and is close to indifferent. Judge
the table first, then the test. If one covariate fails badly, free that
covariate only (`ordinal::clm` with a `nominal =` term).

# Count outcomes

Outcome: `childs`, the number of children.

```{r}
#| label: poisson
f_c <- childs ~ educ + female + black + exper + evermar
m_pois <- glm(f_c, data = gss, family = poisson(link = "log"))
## exp(coef) multiplies the expected count
round(cbind(coef(summary(m_pois))[, 1:2], exp_b = exp(coef(m_pois))), 3)
```

## Is it overdispersed?

The Poisson makes one strong claim: $\Var(Y_i \mid X_i) = \theta_i$. Check it
three ways.

```{r}
#| label: overdispersion
## raw: variance against mean
c(mean = mean(gss$childs), var = var(gss$childs))

## Pearson dispersion: under the Poisson each (y - theta)^2 / theta averages 1
n <- nrow(gss); k <- length(coef(m_pois))
phi <- sum((gss$childs - fitted(m_pois))^2 / fitted(m_pois)) / (n - k)
c(phi = phi)

## formal test, using only the Poisson fit
dispersiontest(m_pois, trafo = 1)
```

The counts scatter 15% more than the Poisson allows. The claim is false, so
the Poisson SE is wrong. Three repairs, each replacing the claim with a
different one:

```{r}
#| label: remedies
m_qp <- glm(f_c, data = gss, family = quasipoisson)   # Var = phi * theta
m_nb <- glm.nb(f_c, data = gss)                       # Var = theta + theta^2 / alpha

round(cbind(
  poisson = sqrt(diag(vcov(m_pois))),
  quasi   = sqrt(diag(vcov(m_qp))),
  negbin  = sqrt(diag(vcov(m_nb)))[names(coef(m_pois))],
  robust  = sqrt(diag(vcovHC(m_pois, type = "HC0")))   # no claim at all
), 4)

## the quasi-Poisson SE is the Poisson SE times sqrt(phi); check on evermar
c(by_hand = sqrt(diag(vcov(m_pois)))["evermar"] * sqrt(phi),
  quasi   = sqrt(diag(vcov(m_qp)))["evermar"])
```

The coefficients barely move (check `coef(m_nb)` against `coef(m_pois)`); the
standard errors do. The robust SE for `evermar` is much larger than the other
two because the excess scatter is not uniform: it sits among the never
married, exactly the group this dummy compares.

```{r}
#| label: nb-theta
## the negative binomial's alpha (theta in MASS): large means close to Poisson
c(alpha = m_nb$theta, se_alpha = m_nb$SE.theta)
```

## Excess zeros, and the hurdle model

```{r}
#| label: zeros
## share of zeros: observed, and averaged over the fitted distributions
c(observed = mean(gss$childs == 0),
  poisson  = mean(dpois(0, fitted(m_pois))),
  negbin   = mean(dnbinom(0, mu = fitted(m_nb), size = m_nb$theta)))
```

Thirty percent are childless; neither model has a way to make zeros special.
The hurdle model does: a logit for zero versus positive, then a truncated
Poisson for the positives, the same covariates in both parts.

```{r}
#| label: hurdle
m_hur <- hurdle(f_c, data = gss, dist = "poisson")
## the two parts tell different stories
round(cbind(any_child = exp(coef(m_hur, model = "zero")),
            how_many  = exp(coef(m_hur, model = "count"))), 2)
## the zero share is matched exactly: the logit part is fitted to it
mean(predict(m_hur, type = "prob")[, 1])
```

Marriage multiplies the odds of *any* child by about 9 but the count among
parents by about 1.35; women are likelier to have any child and have slightly
fewer. One Poisson averaged these two processes.

## Comparing the count models by AIC and BIC

Poisson and negative binomial are nested (the Poisson is the negative binomial
with $\alpha \to \infty$), so an LR test applies. The hurdle is not nested in
either, so use AIC and BIC, which need only a log-likelihood and a parameter
count.

```{r}
#| label: count-compare
## LR: negative binomial against Poisson, one parameter (alpha)
2 * (as.numeric(logLik(m_nb)) - as.numeric(logLik(m_pois)))

## AIC = -2 loglik + 2k;  BIC = -2 loglik + k log n
mods <- list(Poisson = m_pois, `Neg. binomial` = m_nb, Hurdle = m_hur)
tab <- t(sapply(mods, function(m) {
  ll <- logLik(m)
  c(logLik = as.numeric(ll), k = attr(ll, "df"),
    AIC = -2 * as.numeric(ll) + 2 * attr(ll, "df"),
    BIC = -2 * as.numeric(ll) + attr(ll, "df") * log(n))
}))
round(tab, 1)
## check against the built-in functions
round(c(AIC(m_hur), BIC(m_hur)), 1)
```

Read the Poisson-versus-hurdle line the Week 3 way. The hurdle adds six
parameters (a second intercept and five more coefficients). BIC charges
$6 \log 3509 = 49$ for them, AIC charges $12$; the log-likelihood rises by
about $155$, so $-2\,\ell$ falls by about $310$. Both criteria prefer the
hurdle by a wide margin, and BIC's heavier penalty does not change the verdict.
The negative binomial is the interesting line: one extra parameter, a gain of
$2.4$ in log-likelihood. AIC prefers it to the Poisson by about $3$; BIC
prefers the Poisson by about $3$. That is Week 3's disagreement in the wild:
the evidence for $\alpha$ is real (the LR is $4.8$) but not strong enough to
pay BIC's $\log 3509 = 8.2$ per parameter. Both criteria agree that the
zeros, not the variance, are where the Poisson is wrong.

::: callout-note
AIC and BIC compare models fitted to the *same rows* with the *same outcome*.
They say nothing about whether either model is right; the predictive check
below does.
:::

## Does the model reproduce the data?

```{r}
#| label: fig-ppc
#| fig-cap: "Observed distribution of the number of children against the distributions implied by the Poisson, negative binomial, and hurdle fits."
## for each count k, average Pr(Y = k) over everyone's fitted distribution
ks <- 0:8
dist <- bind_rows(
  tibble::tibble(k = ks, source = "observed",
                 p = as.numeric(table(factor(gss$childs, levels = ks))) / n),
  tibble::tibble(k = ks, source = "Poisson",
                 p = sapply(ks, function(j) mean(dpois(j, fitted(m_pois))))),
  tibble::tibble(k = ks, source = "Neg. binomial",
                 p = sapply(ks, function(j) mean(dnbinom(j, mu = fitted(m_nb), size = m_nb$theta)))),
  tibble::tibble(k = ks, source = "Hurdle",
                 p = colMeans(predict(m_hur, type = "prob")[, 1:9]))
) |> mutate(source = factor(source, levels = c("observed", "Poisson", "Neg. binomial", "Hurdle")))

ggplot(dist, aes(k, p, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("observed" = "grey30", "Poisson" = "navy",
                               "Neg. binomial" = "firebrick", "Hurdle" = "darkgreen")) +
  labs(x = "number of children", y = "proportion", fill = NULL) +
  theme_minimal(base_size = 10)
```

This predictive check is the single most informative count diagnostic, and the
one people most often skip. The hurdle matches the zeros by construction; no
count model matches the bump at two children.

::: callout-note
**Which count model.** Start with Poisson and robust SEs. Run the
overdispersion test; if it rejects and you need a likelihood (prediction, AIC,
an LR test), fit the negative binomial. Look at the observed and predicted
distributions. Choose hurdle or zero-inflated on substance: hurdle when
crossing from zero to one is a different *process*, zero-inflated when some
zeros are a different *kind* of person. The two are not nested; AIC or BIC
between them.
:::

# Model fit

```{r}
#| label: fit-compare
AIC(m_np, m_int)
BIC(m_np, m_int)

## McFadden's pseudo-R2
m_null <- glm(ba ~ 1, data = gss, family = binomial)
1 - as.numeric(logLik(m_logit)) / as.numeric(logLik(m_null))
```

```{r}
#| label: fig-roc
#| fig-cap: "ROC curve for the logit model of holding a bachelor's degree."
roc_obj <- roc(gss$ba, fitted(m_logit), quiet = TRUE)
plot(roc_obj, col = "navy", lwd = 2, legacy.axes = TRUE)
text(0.6, 0.2, paste("AUC =", round(auc(roc_obj), 3)))
```

```{r}
#| label: accuracy
## classification accuracy against the null rule of "always predict the majority"
pred <- as.integer(fitted(m_logit) > 0.5)
c(model = mean(pred == gss$ba),
  always_majority = max(mean(gss$ba), 1 - mean(gss$ba)))
```

# Multilevel models (Thursday's session)

The GSS extract has no natural grouping, so this section uses the High School
and Beyond data that ship with `nlme`: 7,185 students in 160 schools, a math
achievement score, student SES, and school sector.

```{r}
#| label: hsb
data(MathAchieve, package = "nlme"); data(MathAchSchool, package = "nlme")
hsb <- merge(as.data.frame(MathAchieve),
             as.data.frame(MathAchSchool)[, c("School", "Sector")], by = "School")
hsb$School <- factor(as.character(hsb$School))   # plain factor, not ordered
c(students = nrow(hsb), schools = length(unique(hsb$School)))
range(table(hsb$School))
```

## The null model and the intraclass correlation

```{r}
#| label: null-model
## no predictors: a grand mean, a school variance, a student variance
m0 <- lmer(MathAch ~ 1 + (1 | School), data = hsb, REML = FALSE)
vc <- as.data.frame(VarCorr(m0))
tau2 <- vc$vcov[1]; sigma2 <- vc$vcov[2]
c(grand_mean = fixef(m0), tau2 = tau2, sigma2 = sigma2, ICC = tau2 / (tau2 + sigma2))

## is tau2 zero? LR against plain OLS, one parameter pinned at zero
ols0 <- lm(MathAch ~ 1, data = hsb)
2 * (as.numeric(logLik(m0)) - as.numeric(logLik(ols0)))
```

Eighteen percent of the variance in math scores is between schools. The LR
statistic is enormous; schools matter.

## Shrinkage, school by school

```{r}
#| label: shrinkage
## own mean, model estimate, and the weight the model puts on the own mean
school_means <- aggregate(MathAch ~ School, data = hsb, mean)
names(school_means)[2] <- "own_mean"
school_means$n <- as.vector(table(hsb$School))
school_means$model <- fixef(m0) + ranef(m0)$School[school_means$School, 1]
school_means$weight <- school_means$n * tau2 / (school_means$n * tau2 + sigma2)

## check the formula on the smallest and largest school
sm <- school_means[c(which.min(school_means$n), which.max(school_means$n)), ]
sm$by_hand <- sm$weight * sm$own_mean + (1 - sm$weight) * fixef(m0)
data.frame(school = sm$School, round(sm[, -1], 2))
```

```{r}
#| label: fig-shrinkage
#| fig-cap: "Model estimates against raw school means. Points below the 45-degree line on the left and above it on the right are being pulled toward the grand mean; small schools move furthest."
plot(school_means$own_mean, school_means$model, cex = 0.3 + school_means$n / 40,
     col = adjustcolor("navy", 0.6), pch = 16,
     xlab = "school's own mean", ylab = "model estimate of the school mean")
abline(0, 1, col = "gray50"); abline(h = fixef(m0), lty = 3, col = "firebrick")
```

## Predictors at both levels

```{r}
#| label: two-levels
m_ols <- lm(MathAch ~ SES + Sector, data = hsb)
m_re  <- lmer(MathAch ~ SES + Sector + (1 | School), data = hsb, REML = FALSE)
m_fe  <- lm(MathAch ~ SES + Sector + School, data = hsb)

round(cbind(
  pooled      = coef(m_ols)[c("SES", "SectorCatholic")],
  pooled_clSE = sqrt(diag(vcovCL(m_ols, cluster = hsb$School)))[c("SES", "SectorCatholic")],
  random_int  = fixef(m_re)[c("SES", "SectorCatholic")],
  random_SE   = sqrt(diag(vcov(m_re)))[c("SES", "SectorCatholic")],
  fixed_eff   = coef(m_fe)[c("SES", "SectorCatholic")]), 3)
as.data.frame(VarCorr(m_re))$vcov
```

Sector is `NA` under fixed effects: it is constant within school, so the
school dummies absorb it. The random-intercept model estimates it, with a
standard error that treats the 160 schools, not the 7,185 students, as the
units. The between-school variance fell from `r round(tau2, 2)` to about 3.6:
SES and sector explain more than half of it.

## A random-intercept logit

```{r}
#| label: glmer
hsb$top <- as.integer(hsb$MathAch > quantile(hsb$MathAch, 0.75))
g1 <- glmer(top ~ SES + Sector + (1 | School), data = hsb,
            family = binomial, nAGQ = 10)   # nAGQ: accuracy of the integral over u_j
round(summary(g1)$coefficients, 3)
as.data.frame(VarCorr(g1))$vcov
```

The same logit as this morning, with a school shift on the log-odds scale
whose variance is estimated. Coefficients are within-school; for
population-average probabilities, average predictions over the distribution
of `u_j`.

# Exercises

1.  **Logit versus LPM for a causal contrast.** Estimate the "effect" of being
    ever-married (`evermar`) on holding a BA, controlling for `pareduc`,
    `wordsum`, `female`, `black`. Compare the LPM coefficient with the logit AME.
    How close are they? Which would you report, and to whom?

2.  **The rescaling problem.** Fit the logit for `ba` with (a) `pareduc` only and
    (b) `pareduc + wordsum + female + black`. Compare the `pareduc` *coefficient*
    across the two, then compare the `pareduc` *AME*. Explain the difference
    using Mood (2010).

3.  **Simulation-based intervals.** Take the logit model and get an interval for
    the difference in predicted probability between `pareduc = 12` and
    `pareduc = 16` by simulating 10,000 draws of $\tilde\beta$ from
    $\mathcal N(\hat\beta, \hat{V}(\hat\beta))$
    (King, Tomz, and Wittenberg 2000). Compare to the delta-method interval from
    `avg_comparisons()`.

4.  **Offsets.** Suppose you wanted to model children *per year since age 18*
    rather than children. Add `offset(log(pmax(age - 18, 1)))` to the Poisson and
    interpret the coefficients. What changed?

5.  **Hurdle versus zero-inflated.** Fit
    `pscl::zeroinfl(childs ~ educ + female + black + exper + evermar, dist = "negbin")`
    and add it to the AIC/BIC table. The two are not nested: which does each
    criterion prefer, and by how much? Then say which story fits the outcome:
    is a childless respondent someone who has not crossed a hurdle, or a
    different kind of person who could never have a positive count?

6.  **A block test in the multinomial.** Drop `educ` from the region model
    and test it with an LR test (how many degrees of freedom?) and with BIC.
    Do the two agree? Explain in one sentence why they might not at $n = 3509$.

7.  **Random slopes.** Extend the random-intercept model to
    `MathAch ~ SES + Sector + (1 + SES | School)`, so that the SES slope varies
    across schools. Report the variance of the slope and its correlation with
    the intercept, test the extension against the random-intercept model with
    an LR test, and say in one sentence what a school with a steep SES slope
    looks like.

# Session information

```{r}
#| label: session
sessionInfo()
```
