---
title: "\\textbf{Lab 4: Limited Dependent Variable Models}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 17"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Fit and read logit and probit models, and see how little the choice matters.
2.  Interpret on the probability scale: average marginal effects and predicted
    probabilities, with intervals.
3.  Handle interactions correctly in a nonlinear model.
4.  Diagnose and repair separation.
5.  Fit count models, detect overdispersion, and choose a remedy.
6.  Fit an ordered logit and test proportional odds.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(marginaleffects)
library(sandwich); library(lmtest)
library(MASS)      # glm.nb, polr
library(AER)       # dispersiontest
library(brglm2)    # Firth penalized likelihood
library(pROC)

gss <- readRDS("../Data/gss_earnings.rds")
```

# Binary outcomes

Outcome: `ba`, whether the respondent holds a bachelor's degree or more.

```{r}
#| label: fit-binary
f <- ba ~ pareduc + wordsum + female + black + exper

m_lpm    <- lm(f, data = gss)
m_logit  <- glm(f, data = gss, family = binomial(link = "logit"))
m_probit <- glm(f, data = gss, family = binomial(link = "probit"))

round(cbind(LPM = coef(m_lpm),
            logit = coef(m_logit),
            probit = coef(m_probit),
            ratio = coef(m_logit) / coef(m_probit)), 4)
```

The logit-to-probit ratio hovers around 1.7, as the theory predicts. The two are
*different* models, different link functions, different assumed error
distributions, but they are close enough empirically that the choice between
them almost never matters for the quantities we report.

```{r}
#| label: lpm-range
## how often does the LPM predict outside [0,1]?
mean(fitted(m_lpm) < 0 | fitted(m_lpm) > 1)
range(fitted(m_lpm))
```

::: callout-note
The linear probability model is not automatically wrong, but check this
number. If a substantial share of fitted values leaves $[0,1]$, the linear
approximation is being asked to do something it cannot.
:::

## Odds ratios

```{r}
#| label: odds
round(exp(cbind(OR = coef(m_logit), confint.default(m_logit))), 3)
```

Read the `wordsum` row as: each additional correct word multiplies the *odds* of
holding a BA by about 1.5. Whether that corresponds to a big or small change in
the *probability* depends entirely on where you start, which is why we now
move to the probability scale.

## Average marginal effects

```{r}
#| label: ame
avg_slopes(m_logit)
```

Compare with the marginal effect at the mean, and with the LPM coefficients:

```{r}
#| label: ame-vs-mem
mem <- slopes(m_logit, newdata = "mean")
comparison <- data.frame(
  variable = c("pareduc", "wordsum", "female", "black", "exper"),
  AME  = avg_slopes(m_logit)$estimate,
  MEM  = mem$estimate,
  LPM  = coef(m_lpm)[-1]
)
round(comparison[, -1], 4) |> `rownames<-`(comparison$variable)
```

Three observations worth taking seriously:

-   The AME and the LPM coefficients are very close. This is typical, and it is
    the main reason people defend the LPM for descriptive and causal work.
-   The MEM differs from the AME. It answers a different question: the effect for
    one hypothetical person at the mean of everything, including means of dummy
    variables.
-   Following Hanmer and Kalkan (2013), report the **AME**.

## Predicted probabilities, the observed-value way

```{r}
#| label: predprob
avg_predictions(m_logit, variables = list(wordsum = c(3, 6, 9)))
```

This sets `wordsum` to each value for *every* respondent, leaving their other
covariates alone, computes predicted probabilities, and averages. Contrast with
setting everything else to its mean:

```{r}
#| label: predprob-mean
predictions(m_logit, newdata = datagrid(wordsum = c(3, 6, 9)))
```

```{r}
#| label: fig-predprob
#| fig-cap: "Average predicted probability of holding a BA by vocabulary score (observed-value approach), with 95% intervals."
plot_predictions(m_logit, by = "wordsum") +
  labs(x = "Vocabulary score (0-10)", y = "Pr(bachelor's or more)") +
  theme_minimal(base_size = 10)
```

And the contrast we probably want to report:

```{r}
#| label: contrast
avg_comparisons(m_logit, variables = list(wordsum = c(3, 9)))
```

## Robust and clustered standard errors

Everything from Week 2 carries over, the sandwich works on `glm` objects.

```{r}
#| label: glm-robust
round(cbind(
  classical = sqrt(diag(vcov(m_logit))),
  HC0       = sqrt(diag(vcovHC(m_logit, type = "HC0"))),
  cluster   = sqrt(diag(vcovCL(m_logit, cluster = ~ psu)))
), 4)
```

```{r}
#| label: ame-robust
avg_slopes(m_logit, vcov = ~ psu)
```

# Interactions

Does the vocabulary gradient differ by gender?

```{r}
#| label: interaction
m_int <- glm(ba ~ pareduc + wordsum * female + black + exper,
             data = gss, family = binomial)
round(coef(summary(m_int))[, 1:2], 4)
```

The product-term coefficient is on the **log-odds** scale. What we actually want
is whether the *marginal effect on the probability* differs by gender:

```{r}
#| label: interaction-me
avg_slopes(m_int, variables = "wordsum", by = "female")
```

```{r}
#| label: interaction-diff
avg_comparisons(m_int, variables = "wordsum", by = "female",
                hypothesis = "b2 - b1 = 0")
```

::: callout-important
Note the two can disagree in sign or significance (Ai and Norton 2003). In a
nonlinear model the interaction on the probability scale depends on where each
person sits on the curve, not only on the product term. **Always report the
quantity you actually mean**, with an interval.
:::

# Separation

Construct a variable that perfectly predicts the outcome, and watch the MLE run
off to infinity.

```{r}
#| label: separation
gss2 <- gss |> mutate(perfect = as.integer(educ >= 18 & ba == 1))
m_sep <- glm(ba ~ pareduc + perfect, data = gss2, family = binomial)
round(coef(summary(m_sep))[, 1:2], 3)
```

The `perfect` coefficient is enormous with an enormous standard error, the
classic signature. The fix is Firth's penalized likelihood:

```{r}
#| label: firth
m_firth <- glm(ba ~ pareduc + perfect, data = gss2,
               family = binomial, method = "brglmFit")
round(coef(summary(m_firth))[, 1:2], 3)
```

Finite estimates, usable standard errors, and the variable is retained
(Zorn 2005).

# Count outcomes

Outcome: `childs`, the number of children.

```{r}
#| label: poisson
m_pois <- glm(childs ~ educ + female + black + exper + evermar,
              data = gss, family = poisson(link = "log"))
round(coef(summary(m_pois))[, 1:2], 4)
```

Coefficients are semi-elasticities: $e^{\beta}$ is the multiplicative change in
the expected count.

```{r}
#| label: pois-irr
round(exp(coef(m_pois)), 3)
```

## Is it overdispersed?

```{r}
#| label: overdispersion
c(mean = mean(gss$childs), var = var(gss$childs))

dispersiontest(m_pois, trafo = 1)
```

```{r}
#| label: remedies
m_qp <- glm(childs ~ educ + female + black + exper + evermar,
            data = gss, family = quasipoisson)
m_nb <- glm.nb(childs ~ educ + female + black + exper + evermar, data = gss)

round(cbind(
  poisson_se   = sqrt(diag(vcov(m_pois))),
  quasi_se     = sqrt(diag(vcov(m_qp))),
  negbin_se    = sqrt(diag(vcov(m_nb)))[names(coef(m_pois))],
  poisson_robust = sqrt(diag(vcovHC(m_pois, type = "HC0")))
), 4)
```

Note that the coefficients barely move but the standard errors do. That is the
signature of overdispersion: the mean is fine, the variance assumption is not.

```{r}
#| label: nb-theta
c(theta = m_nb$theta, se_theta = m_nb$SE.theta)
AIC(m_pois, m_nb)
```

A finite $\hat\theta$ well away from infinity, and a much lower AIC, both say the
negative binomial is doing real work.

## Does the model reproduce the data?

```{r}
#| label: fig-ppc
#| fig-cap: "Observed distribution of the number of children against the distributions implied by the Poisson and negative binomial fits."
set.seed(1)
n <- nrow(gss)
sim <- bind_rows(
  tibble::tibble(k = gss$childs,                                source = "observed"),
  tibble::tibble(k = rpois(n, fitted(m_pois)),                  source = "Poisson"),
  tibble::tibble(k = rnbinom(n, mu = fitted(m_nb), size = m_nb$theta),
                                                                source = "Neg. binomial")
) |>
  count(source, k) |>
  group_by(source) |> mutate(p = n / sum(n)) |> ungroup() |>   # normalize on FULL n
  filter(k <= 8)                                              # then trim for display

ggplot(sim, aes(k, p, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("observed" = "grey30", "Poisson" = "navy",
                               "Neg. binomial" = "firebrick")) +
  labs(x = "number of children", y = "proportion", fill = NULL) +
  theme_minimal(base_size = 10)
```

This *predictive check* is the single most informative count diagnostic, and the
one people most often skip.

# Ordered outcomes

```{r}
#| label: ordered
m_ord <- polr(degree ~ pareduc + wordsum + female + black,
              data = gss, Hess = TRUE)
round(coef(summary(m_ord)), 4)
```

`polr()` does not print $p$-values; get them from the $t$-values, or better, work
on the probability scale:

```{r}
#| label: ordered-pred
avg_predictions(m_ord, variables = list(wordsum = c(3, 9)), type = "probs")
```

## Testing proportional odds

The model imposes one $\beta$ across all cutpoints. Compare it to a multinomial
logit, which frees the coefficients:

The proportional-odds model is *nested inside* a generalized ordered logit that
lets each cutpoint have its own coefficients, so we can test the restriction by
a likelihood-ratio test against that model.

```{r}
#| label: prop-odds
library(VGAM)
## proportional odds imposed
m_po  <- vglm(degree ~ pareduc + wordsum + female + black,
              family = cumulative(parallel = TRUE), data = gss)
## proportional odds relaxed
m_npo <- vglm(degree ~ pareduc + wordsum + female + black,
              family = cumulative(parallel = FALSE), data = gss)

lr <- 2 * (logLik(m_npo) - logLik(m_po))
df <- length(coef(m_npo)) - length(coef(m_po))
c(LR = lr, df = df, p = pchisq(lr, df, lower.tail = FALSE))
```

A small $p$-value says the proportional-odds restriction is rejected. With $n$
this large that is common; look at *how much* the free coefficients differ
before abandoning the parsimony. (An ordered logit and a *multinomial* logit are
not nested, so their deviances cannot be compared this way.)

# Model fit

```{r}
#| label: fit-compare
AIC(m_logit, m_int)
BIC(m_logit, m_int)

## McFadden's pseudo-R2
m_null <- glm(ba ~ 1, data = gss, family = binomial)
1 - as.numeric(logLik(m_logit)) / as.numeric(logLik(m_null))
```

```{r}
#| label: fig-roc
#| fig-cap: "ROC curve for the logit model of holding a bachelor's degree."
roc_obj <- roc(gss$ba, fitted(m_logit), quiet = TRUE)
plot(roc_obj, col = "navy", lwd = 2, legacy.axes = TRUE)
text(0.6, 0.2, paste("AUC =", round(auc(roc_obj), 3)))
```

```{r}
#| label: accuracy
## classification accuracy against the null rule of "always predict the majority"
pred <- as.integer(fitted(m_logit) > 0.5)
c(model = mean(pred == gss$ba),
  always_majority = max(mean(gss$ba), 1 - mean(gss$ba)))
```

# Exercises

1.  **Logit versus LPM for a causal contrast.** Estimate the "effect" of being
    ever-married (`evermar`) on holding a BA, controlling for `pareduc`,
    `wordsum`, `female`, `black`. Compare the LPM coefficient with the logit AME.
    How close are they? Which would you report, and to whom?

2.  **The rescaling problem.** Fit the logit for `ba` with (a) `pareduc` only and
    (b) `pareduc + wordsum + female + black`. Compare the `pareduc` *coefficient*
    across the two, then compare the `pareduc` *AME*. Explain the difference
    using Mood (2010).

3.  **Simulation-based intervals.** Take the logit model and get an interval for
    the difference in predicted probability between `wordsum = 3` and
    `wordsum = 9` by simulating 10,000 draws of $\tilde\beta$ from
    $\mathcal N(\hat\beta, \hat{V}(\hat\beta))$
    (King, Tomz, and Wittenberg 2000). Compare to the delta-method interval from
    `avg_comparisons()`.

4.  **Offsets.** Suppose you wanted to model children *per year since age 18*
    rather than children. Add `offset(log(pmax(age - 18, 1)))` to the Poisson and
    interpret the coefficients. What changed?

5.  **Zero inflation.** Fit `pscl::zeroinfl(childs ~ ... | evermar + age)` and
    compare with the negative binomial by AIC and by the predictive check above.
    Does the zero-inflation story make substantive sense here?

6.  **Separation in the wild.** Fit a logit for `ba` including
    `factor(region) * black`. Does anything blow up? If so, apply Firth and
    report what changes.

# Session information

```{r}
#| label: session
sessionInfo()
```
