---
title: "\\textbf{Lab 4: Limited Dependent Variable Models}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 17"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Fit a logit, then read it on the probability scale: predicted probabilities
    first, average marginal effects second, odds ratios as a footnote.
2.  Handle an interaction in a nonlinear model: when to include the product,
    and how to report the interaction.
3.  Fit a multinomial logit on a nominal outcome and an ordered logit on an
    ordered one.
4.  Fit count models: Poisson, the three overdispersion repairs, the hurdle
    model; compare them by AIC and BIC.
5.  Fit a random-intercept model, read the two variances and the intraclass
    correlation, and compare it with fixed effects.

```{r}
#| label: packages
library(lme4)      # lmer, glmer
library(dplyr)
library(ggplot2)
library(marginaleffects)
library(sandwich); library(lmtest)
library(MASS)      # glm.nb, polr
library(AER)       # dispersiontest
library(nnet)      # multinom
library(pscl)      # hurdle, zeroinfl

gss <- readRDS("../Data/gss_earnings.rds")
```

# Binary outcomes

Outcome: `ba`, whether the respondent holds a bachelor's degree or more. The
model is Tuesday's: parents' education, vocabulary score, female, black.

```{r}
#| label: fit-binary
f <- ba ~ pareduc + wordsum + female + black

m_logit  <- glm(f, data = gss, family = binomial(link = "logit"))
m_probit <- glm(f, data = gss, family = binomial(link = "probit"))
m_lpm    <- lm(f, data = gss)

## the logit table from the slides: coefficient, SE, z, exp(coefficient)
round(cbind(coef(summary(m_logit))[, 1:3], exp_b = exp(coef(m_logit))), 3)
```

## Three scales, one coefficient

The `wordsum` coefficient is $0.391$ on the log-odds scale, $e^{0.391} = 1.48$
on the odds scale, and *not one number* on the probability scale. The first
two are exact and uninterpretable to most readers; the third is what the
reader wants and depends on where the person starts.

```{r}
#| label: three-scales
## one person at a time: log-odds, odds, probability for wordsum = 3, 6, 9
## with the other covariates at pareduc = 12, female = 0, black = 0
b  <- coef(m_logit)
xb <- b["(Intercept)"] + b["pareduc"] * 12 + b["wordsum"] * c(3, 6, 9)
round(cbind(wordsum = c(3, 6, 9), log_odds = xb, odds = exp(xb),
            prob = plogis(xb)), 3)
```

Each step of three words adds the same $3 \times 0.391 = 1.17$ to the
log-odds and multiplies the odds by the same $e^{1.17} = 3.23$; the probability
moves by a different amount each time.

## Start with predicted probabilities

How likely is a degree if your parents finished high school (12 years) rather
than college (16)? The observed-value approach: set `pareduc` to 12 for
*everyone*, keep every other covariate at its actual value, average the fitted
probabilities; repeat at 16.

```{r}
#| label: predprob
## avg_predictions() does exactly that loop, with a delta-method interval
avg_predictions(m_logit, variables = list(pareduc = c(8, 12, 16, 20)))
```

```{r}
#| label: contrast
## the contrast we report: high school to college, 18 points
avg_comparisons(m_logit, variables = list(pareduc = c(12, 16)))
```

```{r}
#| label: fig-predprob
#| fig-cap: "Average predicted probability of holding a BA by parents' education (observed-value approach), with 95% intervals."
## the counterfactual grid sets pareduc to each value for EVERYONE, then averages;
## plain `by = "pareduc"` would instead average fitted values among the people
## who actually have that pareduc, which mixes in their other covariates
plot_predictions(m_logit, by = "pareduc",
                 newdata = datagrid(pareduc = 8:20, grid_type = "counterfactual")) +
  labs(x = "Parents' education (years)", y = "Pr(bachelor's or more)") +
  theme_minimal(base_size = 10)
```

::: callout-note
`plot_predictions(m, by = "pareduc")` without the counterfactual grid is a
different quantity: the average fitted probability *among respondents whose
parents have that much education*. At 20 years it gives 0.75 rather than 0.69, and at 8 years 0.16
rather than 0.20, because those respondents also differ in vocabulary and the
other covariates. The counterfactual version isolates parents' education; the
subgroup version describes who those people are.
:::

Contrast with setting everything else to its mean, which describes one
hypothetical person (a 0.48 woman, a 0.15 Black respondent). Build that grid
explicitly: `datagrid()`'s default is *not* the mean for every variable. It uses
the mean for continuous covariates but the mode for binary and integer-valued
ones, which here would give a man, not Black, with a vocabulary score of 6.

```{r}
#| label: predprob-mean
## one row per pareduc value, every other covariate at its sample mean
at_means <- datagrid(model = m_logit, pareduc = c(12, 16), wordsum = mean(gss$wordsum),
                     female = mean(gss$female), black = mean(gss$black))
predictions(m_logit, newdata = at_means)
```

## Then marginal effects: one number per variable

```{r}
#| label: ame
## AME: each person's own slope dp/dx = beta * p(1 - p), averaged
avg_slopes(m_logit)
```

Compare with the marginal effect at the mean and with the LPM coefficients:

```{r}
#| label: ame-vs-mem
## MEM: the slope for one person at the sample means (same caution as above:
## newdata = "mean" would use modes for the binary and integer variables)
mem <- slopes(m_logit, newdata = datagrid(model = m_logit, pareduc = mean(gss$pareduc),
                                          wordsum = mean(gss$wordsum), female = mean(gss$female),
                                          black = mean(gss$black)))
ame <- avg_slopes(m_logit)
## match by variable name: avg_slopes() returns terms in alphabetical order,
## coef() in formula order
comparison <- data.frame(
  AME = ame$estimate,
  MEM = mem$estimate[match(ame$term, mem$term)],
  LPM = coef(m_lpm)[ame$term],
  row.names = ame$term
)
round(comparison, 4)
```

-   The AME and the LPM coefficients are close. This is typical, and it is the
    main reason people defend the LPM for descriptive and causal work.
-   The MEM answers a different question: the slope for one hypothetical person
    at the mean of everything, including the means of dummies.
-   Report the AME (Hanmer and Kalkan 2013).

## Odds ratios, at most as a footnote

```{r}
#| label: odds
round(exp(cbind(OR = coef(m_logit), confint.default(m_logit))), 3)
```

Each correct word multiplies the *odds* of a BA by 1.48. That is one number for
everyone, because the model is linear on the log-odds scale. It is not "48%
more likely": the probability change depends on where the person starts, which
is why the probabilities came first.

## Uncertainty: robust and clustered standard errors

Every quantity in this section, a predicted probability, a contrast, an AME,
is a function $q(\hat\beta)$ of the coefficients. Its standard error comes from
the delta method: linearize $q$ around $\hat\beta$ and use
$\widehat{\mathrm{Var}}(q) = \nabla q'\,\hat V(\hat\beta)\,\nabla q$. The
gradient $\nabla q$ is fixed by the model; the only choice is which
$\hat V(\hat\beta)$ to plug in. The `vcov` argument of every `marginaleffects`
function makes that choice, so a robust or clustered standard error for an AME
is the same sandwich as in Week 2, pushed through the same gradient.

Three variance matrices for the coefficients first:

```{r}
#| label: glm-robust
## classical: inverse information (Week 3). HC0: the sandwich, each person's own
## squared score. HC1: HC0 times n/(n - k). cluster: scores summed within a GSS
## sampling unit (psu) before squaring, so people in the same psu may be correlated
round(cbind(
  classical = sqrt(diag(vcov(m_logit))),
  HC0       = sqrt(diag(vcovHC(m_logit, type = "HC0"))),
  HC1       = sqrt(diag(vcovHC(m_logit, type = "HC1"))),
  cluster   = sqrt(diag(vcovCL(m_logit, cluster = ~ psu)))
), 4)
```

The same four choices carried through to the AMEs. `vcov = TRUE` is the
classical matrix, a string names an HC type, a one-sided formula names the
cluster variable:

```{r}
#| label: ame-robust
se_of <- function(v) {
  a <- as.data.frame(avg_slopes(m_logit, vcov = v))
  setNames(a$std.error, a$term)
}
round(rbind(classical = se_of(TRUE), HC0 = se_of("HC0"), HC1 = se_of("HC1"),
            cluster_psu = se_of(~ psu)), 4)
```

Predicted probabilities take the same argument:

```{r}
#| label: pred-robust
avg_predictions(m_logit, variables = list(pareduc = c(12, 16)), vcov = ~ psu)
```

Two things to notice. The robust and classical standard errors are within a
few percent of each other here: for a logit, a large gap between them is a
warning that the model is misspecified rather than a fix, because the logit's
variance is not a free parameter the way the Poisson's is. And clustering by
`psu` matters little for this outcome because the GSS sampling units are small
(839 of them, a median of four respondents each); with respondents nested in
states or years it would matter more. A full design-based analysis would also
use the strata and weights, through the `survey` package.

## Probit: another S-curve

```{r}
#| label: probit
## coefficients differ by a scale factor; AMEs do not
round(cbind(logit = coef(m_logit), probit = coef(m_probit),
            ratio = coef(m_logit) / coef(m_probit)), 3)
round(cbind(AME_logit  = avg_slopes(m_logit)$estimate,
            AME_probit = avg_slopes(m_probit)$estimate), 4) |>
  `rownames<-`(avg_slopes(m_logit)$term)
```

The coefficients differ by a factor of about 1.6 to 1.7 because the two
S-curves have different steepness at the center. The AMEs agree to two
decimals. Default to logit; use probit when the next step needs a normal
(selection models, bivariate probit) or when the subfield expects it.

::: callout-important
Never compare raw logit or probit coefficients across nested models or across
groups. Adding a covariate rescales every coefficient, even ones whose effect is
unchanged (Mood 2010). Compare AMEs or predicted probabilities.
:::

# Interactions

Does parents' education matter more for women? Two questions: should the
product `pareduc * female` be in the model, and do the models with and
without it agree on the probability scale?

```{r}
#| label: interaction
## no product: one log-odds slope for pareduc, forced equal for men and women
m_np  <- glm(ba ~ pareduc + female + wordsum + black, data = gss, family = binomial)
## with product: the slope is beta_1 + beta_12 * female; the data choose beta_12
m_int <- glm(ba ~ pareduc * female + wordsum + black, data = gss, family = binomial)
round(coef(summary(m_int))[, 1:3], 3)
## LR test of the product
anova(m_np, m_int, test = "LRT")
```

The product is $0.021$ on the log-odds scale, $z = 0.79$. That is the wrong
scale for the question. The AME of parents' education, by gender, in both
models:

```{r}
#| label: interaction-me
## no product: the AMEs still differ by gender, through p(1 - p) alone
avg_slopes(m_np,  variables = "pareduc", by = "female")
## with product: they differ through p(1 - p) and through beta_12
avg_slopes(m_int, variables = "pareduc", by = "female")
```

```{r}
#| label: interaction-diff
## the interaction we report: the difference in AMEs, with an interval
avg_slopes(m_int, variables = "pareduc", by = "female",
           hypothesis = "b2 - b1 = 0")
```

-   **Include the product** whenever the question is whether $X_1$'s effect
    differs by $X_2$. Without it the answer is fixed in advance: zero on the
    log-odds scale, and on the probability scale whatever the S-curve gives
    (the no-product model already shows women with a larger AME, because
    they sit where the curve is steeper: Ai and Norton 2003).
-   **The two models agree** on probabilities only when $\hat\beta_{12}$ is
    near zero, as here. When it is not, the no-product model has the wrong
    probabilities in the cells.
-   **Report** the difference in AMEs with its interval, here $0.006$ per year,
    SE $0.004$. The answer is no.

# Outcomes with several categories

## Multinomial logit: a nominal outcome

Where a respondent lives, in four census regions. The categories have no
order, so one logit for every region against a base (the Midwest).

```{r}
#| label: region4
## collapse the nine GSS regions to four; Midwest first so it is the base
gss$region4 <- factor(case_when(
  gss$region %in% c("New England", "Middle Atlantic") ~ "Northeast",
  gss$region %in% c("E. North Central", "W. North Central") ~ "Midwest",
  gss$region %in% c("South Atlantic", "E. South Central", "W. South Central") ~ "South",
  TRUE ~ "West"), levels = c("Midwest", "Northeast", "South", "West"))
table(gss$region4)

m_mn <- multinom(region4 ~ educ + black + female + age, data = gss, trace = FALSE)
## one row per non-base region: exp(coef) multiplies the odds AGAINST THE MIDWEST
round(exp(coef(m_mn)), 2)
```

Read the `black` column: being Black multiplies the odds of South-versus-Midwest
by about 2.9 and of West-versus-Midwest by about 0.8. Those are odds against
the base, not probabilities; the four probabilities move together and add to
one:

```{r}
#| label: region4-pred
## observed-value predicted probabilities, one column per region
avg_predictions(m_mn, variables = "black")
## check: for each person the four probabilities sum to one
range(rowSums(fitted(m_mn)))
```

```{r}
#| label: region4-lr
## do the covariates matter at all? LR against the intercept-only model
m_mn0 <- multinom(region4 ~ 1, data = gss, trace = FALSE)
lr <- 2 * (as.numeric(logLik(m_mn)) - as.numeric(logLik(m_mn0)))
df <- length(coef(m_mn)) - length(coef(m_mn0))
c(LR = round(lr, 1), df = df)
pchisq(lr, df, lower.tail = FALSE) < 0.001
```

## Ordered logit: an ordered outcome

Highest degree, five ordered categories. One coefficient vector and four
cutpoints.

```{r}
#| label: ordered
m_ord <- polr(degree ~ pareduc + wordsum + female + black,
              data = gss, Hess = TRUE)
round(coef(summary(m_ord)), 4)
```

`polr()` does not print $p$-values; get them from the $t$-values, or better,
work on the probability scale.

```{r}
#| label: ordered-pred
## observed-value predictions: set wordsum to 3 for everyone, then 9, keep the
## other covariates as observed, average the five probabilities
avg_predictions(m_ord, variables = list(wordsum = c(3, 9)), type = "probs")
```

Four more words move about a quarter of the population out of "high school
only," mostly into the two degree categories. The middle category barely
moves: one coefficient pushes probability *through* a narrow band without
piling it up there, so the sign of the effect on a middle category is not the
sign of $\hat\beta$.

```{r}
#| label: ordered-ame
## AMEs: one per covariate per category, the average change in each category's
## probability per unit of the covariate (or for the dummy switching 0 to 1)
avg_slopes(m_ord, type = "probs")
```

Read the `group` column: each covariate has five AMEs, one per category, and
the five sum to zero because the probabilities sum to one. This is the
standard way to report an ordered logit: category probabilities at chosen
values, and the AMEs by category, rather than the cutpoints.

# Count outcomes

Outcome: `childs`, the number of children.

```{r}
#| label: poisson
f_c <- childs ~ educ + female + black + exper + evermar
m_pois <- glm(f_c, data = gss, family = poisson(link = "log"))
## exp(coef) multiplies the expected count
round(cbind(coef(summary(m_pois))[, 1:2], exp_b = exp(coef(m_pois))), 3)
```

## Is it overdispersed?

The Poisson makes one strong claim: $\Var(Y_i \mid X_i) = \theta_i$. Check it
three ways.

```{r}
#| label: overdispersion
## raw: variance against mean
c(mean = mean(gss$childs), var = var(gss$childs))

## Pearson dispersion: under the Poisson each (y - theta)^2 / theta averages 1
n <- nrow(gss); k <- length(coef(m_pois))
phi <- sum((gss$childs - fitted(m_pois))^2 / fitted(m_pois)) / (n - k)
c(phi = phi)

## formal test, using only the Poisson fit. trafo says what shape the extra
## variance takes: trafo = 1 tests Var = theta + alpha * theta (a constant
## multiple of the mean, the quasi-Poisson shape); trafo = 2 tests
## Var = theta + alpha * theta^2 (the negative-binomial shape). Either way
## H0 is alpha = 0 and the printed alpha is the estimated excess
dispersiontest(m_pois, trafo = 1)
```

The counts scatter 15% more than the Poisson allows. The claim is false, so
the Poisson SE is wrong. Three repairs, each replacing the claim with a
different one:

```{r}
#| label: remedies
m_qp <- glm(f_c, data = gss, family = quasipoisson)   # Var = phi * theta
m_nb <- glm.nb(f_c, data = gss)                       # Var = theta + theta^2 / alpha

## the coefficients barely move across the three likelihoods ...
round(cbind(poisson = coef(m_pois), quasi = coef(m_qp), negbin = coef(m_nb)), 3)

## ... the standard errors do
round(cbind(
  poisson = sqrt(diag(vcov(m_pois))),
  quasi   = sqrt(diag(vcov(m_qp))),
  negbin  = sqrt(diag(vcov(m_nb)))[names(coef(m_pois))],
  robust  = sqrt(diag(vcovHC(m_pois, type = "HC0")))   # no claim at all
), 4)

## the quasi-Poisson SE is the Poisson SE times sqrt(phi); check on evermar
c(by_hand = sqrt(diag(vcov(m_pois)))["evermar"] * sqrt(phi),
  quasi   = sqrt(diag(vcov(m_qp)))["evermar"])
```

Quasi-Poisson keeps the Poisson coefficients exactly, since it changes only
the variance; the negative binomial refits and lands within a hundredth of
them. The standard errors are what move. The robust SE for `evermar` is much
larger than the other two because the excess scatter is not uniform: it sits
among the never married, exactly the group this dummy compares.

## Excess zeros, and the hurdle model

```{r}
#| label: zeros
## share of zeros: observed, and averaged over the fitted distributions
c(observed = mean(gss$childs == 0),
  poisson  = mean(dpois(0, fitted(m_pois))),
  negbin   = mean(dnbinom(0, mu = fitted(m_nb), size = m_nb$theta)))
```

Thirty percent are childless; neither model has a way to make zeros special.
The hurdle model does: a logit for zero versus positive, then a truncated
Poisson for the positives, the same covariates in both parts.

```{r}
#| label: hurdle
m_hur <- hurdle(f_c, data = gss, dist = "poisson")
## the two parts tell different stories
round(cbind(any_child = exp(coef(m_hur, model = "zero")),
            how_many  = exp(coef(m_hur, model = "count"))), 2)
## the zero share is matched exactly: the logit part is fitted to it
mean(predict(m_hur, type = "prob")[, 1])
```

Marriage multiplies the odds of *any* child by about 9 but the count among
parents by about 1.35; women are likelier to have any child and have slightly
fewer. One Poisson averaged these two processes.

The zero-inflated model tells the other story: a share of people are
"structural zeros" who will never have a child, and everyone else follows a
Poisson that can also produce zeros. Same syntax, different meaning of the
zero part.

```{r}
#| label: zip
m_zip <- zeroinfl(f_c, data = gss, dist = "poisson")
## zero part: a logit for being a STRUCTURAL zero (not for having any child, as
## in the hurdle), so the signs flip relative to the hurdle's zero part
round(cbind(structural_zero = exp(coef(m_zip, model = "zero")),
            count_part      = exp(coef(m_zip, model = "count"))), 2)
## two kinds of zero: the structural share, and the total share the model implies
c(structural = mean(predict(m_zip, type = "zero")),
  total      = mean(predict(m_zip, type = "prob")[, 1]),
  observed   = mean(gss$childs == 0))
```

Read the zero part with care: an odds ratio of 0.003 for `evermar` says the
ever-married are almost never structural zeros; the structural zeros are the
never-married, and the more educated. About 16% of respondents are estimated
to be structural zeros; the rest of the 30% childless are "zeros for now" from
the count part. Unlike the hurdle, the zero share is not matched exactly.

## Comparing the count models by AIC and BIC

Poisson and negative binomial are nested (the Poisson is the negative binomial
with $\alpha \to \infty$), so an LR test applies. The hurdle and the
zero-inflated model are not nested in either, or in each other, so use AIC and
BIC, which need only a log-likelihood and a parameter count.

```{r}
#| label: count-compare
## LR: negative binomial against Poisson, one parameter (alpha)
2 * (as.numeric(logLik(m_nb)) - as.numeric(logLik(m_pois)))

## AIC = -2 loglik + 2k;  BIC = -2 loglik + k log n
mods <- list(Poisson = m_pois, `Neg. binomial` = m_nb, Hurdle = m_hur, `Zero-inflated` = m_zip)
tab <- t(sapply(mods, function(m) {
  ll <- logLik(m)
  c(logLik = as.numeric(ll), k = attr(ll, "df"),
    AIC = -2 * as.numeric(ll) + 2 * attr(ll, "df"),
    BIC = -2 * as.numeric(ll) + attr(ll, "df") * log(n))
}))
round(tab, 1)
## check against the built-in functions
round(c(AIC(m_hur), BIC(m_hur)), 1)
```

Read the Poisson-versus-hurdle line the Week 3 way. The hurdle adds six
parameters (a second intercept and five more coefficients). BIC charges
$6 \log 3509 = 49$ for them, AIC charges $12$; the log-likelihood rises by
about $155$, so $-2\,\ell$ falls by about $310$. Both criteria prefer the
hurdle by a wide margin, and BIC's heavier penalty does not change the verdict.
The zero-inflated model, with the same twelve parameters, fits better still, by
about 56 on either criterion. Fit alone would pick it; whether a permanently
childless class is the right story is a substantive question the criteria
cannot answer, and the hurdle's "any child, then how many" reading is the one
demographers usually defend. The negative binomial is the interesting line: one extra parameter, a gain of
$2.4$ in log-likelihood. AIC prefers it to the Poisson by about $3$; BIC
prefers the Poisson by about $3$. That is Week 3's disagreement in the wild:
the evidence for $\alpha$ is real (the LR is $4.8$) but not strong enough to
pay BIC's $\log 3509 = 8.2$ per parameter. Both criteria agree that the
zeros, not the variance, are where the Poisson is wrong.

## Does the model reproduce the data?

```{r}
#| label: fig-ppc
#| fig-cap: "Observed distribution of the number of children against the distributions implied by the Poisson, negative binomial, and hurdle fits."
## for each count k, average Pr(Y = k) over everyone's fitted distribution
ks <- 0:8
dist <- bind_rows(
  tibble::tibble(k = ks, source = "observed",
                 p = as.numeric(table(factor(gss$childs, levels = ks))) / n),
  tibble::tibble(k = ks, source = "Poisson",
                 p = sapply(ks, function(j) mean(dpois(j, fitted(m_pois))))),
  tibble::tibble(k = ks, source = "Neg. binomial",
                 p = sapply(ks, function(j) mean(dnbinom(j, mu = fitted(m_nb), size = m_nb$theta)))),
  tibble::tibble(k = ks, source = "Hurdle",
                 p = colMeans(predict(m_hur, type = "prob")[, 1:9]))
) |> mutate(source = factor(source, levels = c("observed", "Poisson", "Neg. binomial", "Hurdle")))

ggplot(dist, aes(k, p, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("observed" = "grey30", "Poisson" = "navy",
                               "Neg. binomial" = "firebrick", "Hurdle" = "darkgreen")) +
  labs(x = "number of children", y = "proportion", fill = NULL) +
  theme_minimal(base_size = 10)
```

This predictive check is the single most informative count diagnostic, and the
one people most often skip. The hurdle matches the zeros by construction; no
count model matches the bump at two children.

::: callout-note
**Which count model.** Start with Poisson and robust SEs. Run the
overdispersion test; if it rejects and you need a likelihood (prediction, AIC,
an LR test), fit the negative binomial. Look at the observed and predicted
distributions. Choose hurdle or zero-inflated on substance: hurdle when
crossing from zero to one is a different *process*, zero-inflated when some
zeros are a different *kind* of person. The two are not nested; AIC or BIC
between them.
:::

# Multilevel models

The GSS extract has no natural grouping, so this section uses the High School
and Beyond data that ship with `nlme`: 7,185 students in 160 schools, a math
achievement score, student SES, and school sector.

```{r}
#| label: hsb
data(MathAchieve, package = "nlme"); data(MathAchSchool, package = "nlme")
hsb <- merge(as.data.frame(MathAchieve),
             as.data.frame(MathAchSchool)[, c("School", "Sector")], by = "School")
hsb$School <- factor(as.character(hsb$School))   # plain factor, not ordered
c(students = nrow(hsb), schools = length(unique(hsb$School)))
range(table(hsb$School))
```

## Reading the `lmer` formula

The random-intercept model is

$$
\text{MathAch}_{ij} = \beta_0 + \beta_1 \text{SES}_{ij} + \beta_2 \text{Catholic}_j
  + u_j + e_{ij}, \qquad u_j \sim N(0, \tau^2),\; e_{ij} \sim N(0, \sigma^2).
$$

In `lmer` the formula is `MathAch ~ SES + Sector + (1 | School)`. The part
outside the parentheses is the fixed part, an ordinary regression formula for
the $\beta$s. The part in parentheses is the random part, and it has two sides.
Right of the bar is the grouping variable: which rows share a $u_j$. Left of
the bar is the design of the random part: which columns get a coefficient
that varies by group. `1` is the intercept column, so `(1 | School)` gives
every school its own intercept shift, the $u_j$ that multiplies the column of
ones; the model estimates its variance $\tau^2$, not the 160 shifts.

```{r}
#| label: formula-syntax
#| eval: false
(1 | School)          # random intercept: u_j, one variance tau^2
(1 + SES | School)    # random intercept AND a random slope on SES:
                      #   u_j + v_j * SES_ij, two variances and their covariance
(0 + SES | School)    # random slope only, common intercept
(1 | School) + (1 | District)   # two grouping levels, one u for each
```

A random slope says the SES gradient itself differs across schools; exercise
7 fits it. Everything in this lab keeps `(1 | School)`.

## Predictors at both levels

```{r}
#| label: two-levels
m_ols <- lm(MathAch ~ SES + Sector, data = hsb)
## REML = FALSE: plain maximum likelihood, so logLik() is the likelihood from
## Week 3 and models with different fixed effects can be compared by LR, AIC,
## BIC. The default REML = TRUE corrects the variances for the degrees of
## freedom spent on the betas (like RSS/(n - k) in OLS) but its likelihood is
## not comparable across fixed effects. On HSB the two differ in the third decimal
m_re  <- lmer(MathAch ~ SES + Sector + (1 | School), data = hsb, REML = FALSE)
## fixed effects: school dummies first. Sector is constant within school, so
## it is collinear with the dummies; lm() drops the LAST collinear column, and
## with Sector listed before School it would keep Sector and drop a school
## dummy instead, printing a sector "effect" measured against whichever school
## was dropped. Listing School first makes lm() report the honest NA.
m_fe  <- lm(MathAch ~ School + SES + Sector, data = hsb)

round(cbind(
  pooled      = coef(m_ols)[c("SES", "SectorCatholic")],
  pooled_clSE = sqrt(diag(vcovCL(m_ols, cluster = hsb$School)))[c("SES", "SectorCatholic")],
  random_int  = fixef(m_re)[c("SES", "SectorCatholic")],
  random_SE   = sqrt(diag(vcov(m_re)))[c("SES", "SectorCatholic")],
  fixed_eff   = coef(m_fe)[c("SES", "SectorCatholic")]), 3)
```

Sector is `NA` under fixed effects: it is constant within school, so the
school dummies absorb it. The random-intercept model estimates it, with a
standard error that treats the 160 schools, not the 7,185 students, as the
units. The SES slope sits between the pooled and the fixed-effects one: the
random-intercept model uses both within-school and between-school variation.

## The two variances and the intraclass correlation

```{r}
#| label: variances
## tau2 = between-school variance of the intercepts, sigma2 = within-school
vc <- as.data.frame(VarCorr(m_re))
tau2 <- vc$vcov[1]; sigma2 <- vc$vcov[2]
c(tau2 = tau2, sigma2 = sigma2, ICC = tau2 / (tau2 + sigma2))

## is tau2 zero? LR against OLS with the same x, one parameter pinned at zero
2 * (as.numeric(logLik(m_re)) - as.numeric(logLik(m_ols)))
```

Of the variation SES and sector leave unexplained, 9% is between schools and
91% within. The LR statistic is enormous; schools still differ after the
controls.

## A random-intercept logit

```{r}
#| label: glmer
#| warning: false
hsb$top <- as.integer(hsb$MathAch > quantile(hsb$MathAch, 0.75))
g1 <- glmer(top ~ SES + Sector + (1 | School), data = hsb,
            family = binomial, nAGQ = 10)   # nAGQ: accuracy of the integral over u_j
round(summary(g1)$coefficients, 3)

## observed-value predictions: set every student's SES to each value, keep their
## own sector and their own school's predicted u_j (the default, re.form = NULL),
## average the 7,185 probabilities. re.form = NA would set every u_j to zero.
## The standard error carries the uncertainty in the coefficients only
avg_predictions(g1, variables = list(SES = seq(-2, 2, by = 0.5)))
```

The same logit as this morning, with a school shift on the log-odds scale
whose variance is estimated. A coefficient is the log-odds change holding the
school's `u_j` fixed. The predictions put the coefficient on the probability
scale: moving every student from SES $-1$ to SES $+1$ (the SD of SES is
0.78) raises the share in the top quartile from 0.14 to 0.37.

The figure from the slides shows what the school shift does to the curve.
For a public school, the S-curve depends on which school: $u_j = 0$ is the
typical school, $\pm\hat\tau$ are schools one SD above and below. The
population-average curve integrates the probability over $u_j \sim
N(0, \hat\tau^2)$; averaging S-curves with different intercepts spreads the
transition out, so it is flatter than the $u_j = 0$ curve, only slightly here
because $\hat\tau = 0.58$ is small.

```{r}
#| label: fig-glmer
#| fig-cap: "Pr(top quartile in math) by SES, public schools: the curve for a typical school, for schools one SD above and below, and the average over the fitted distribution of school shifts."
b <- fixef(g1)
tau <- sqrt(as.data.frame(VarCorr(g1))$vcov[1])   # SD of the school shifts
ses <- seq(-2, 2, by = 0.05)
eta <- b["(Intercept)"] + b["SES"] * ses            # public school: Sector dummy = 0

## population average: integrate plogis(eta + u) over u ~ N(0, tau^2) on a grid
u <- seq(-4 * tau, 4 * tau, length.out = 401)
w <- dnorm(u, 0, tau); w <- w / sum(w)
p_avg <- sapply(eta, function(e) sum(plogis(e + u) * w))

curves <- data.frame(
  SES = rep(ses, 4),
  p = c(plogis(eta + tau), plogis(eta), plogis(eta - tau), p_avg),
  school = rep(c("u_j = +tau", "u_j = 0", "u_j = -tau", "average over schools"),
               each = length(ses)))
curves$school <- factor(curves$school, levels = unique(curves$school))
ggplot(curves, aes(SES, p, colour = school, linetype = school)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("#00539B", "#00539B", "#00539B", "#C84E00")) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted", "solid")) +
  labs(x = "Student SES", y = "Pr(top quartile in math)",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
```

The table above averages over the students as they are, in both sectors and
in their own schools; the orange curve averages over the school distribution
for public schools only. Both are population-average quantities; the
coefficient $0.674$ describes the blue curves, one school at a time.

# Exercises

1.  **Logit versus LPM for a causal contrast.** Estimate the "effect" of being
    ever-married (`evermar`) on holding a BA, controlling for `pareduc`,
    `wordsum`, `female`, `black`. Compare the LPM coefficient with the logit AME.
    How close are they? Which would you report, and to whom?

2.  **The rescaling problem.** Fit the logit for `ba` with (a) `pareduc` only and
    (b) `pareduc + wordsum + female + black`. Compare the `pareduc` *coefficient*
    across the two, then compare the `pareduc` *AME*. Explain the difference
    using Mood (2010).

3.  **Simulation-based intervals.** Take the logit model and get an interval for
    the difference in predicted probability between `pareduc = 12` and
    `pareduc = 16` by simulating 10,000 draws of $\tilde\beta$ from
    $\mathcal N(\hat\beta, \hat{V}(\hat\beta))$
    (King, Tomz, and Wittenberg 2000). Compare to the delta-method interval from
    `avg_comparisons()`.

4.  **Offsets.** Suppose you wanted to model children *per year since age 18*
    rather than children. Add `offset(log(pmax(age - 18, 1)))` to the Poisson and
    interpret the coefficients. What changed?

5.  **Hurdle versus zero-inflated.** Fit
    `pscl::zeroinfl(childs ~ educ + female + black + exper + evermar, dist = "negbin")`
    and add it to the AIC/BIC table. The two are not nested: which does each
    criterion prefer, and by how much? Then say which story fits the outcome:
    is a childless respondent someone who has not crossed a hurdle, or a
    different kind of person who could never have a positive count?

6.  **A block test in the multinomial.** Drop `educ` from the region model
    and test it with an LR test (how many degrees of freedom?) and with BIC.
    Do the two agree? Explain in one sentence why they might not at $n = 3509$.

7.  **Random slopes.** Extend the random-intercept model to
    `MathAch ~ SES + Sector + (1 + SES | School)`, so that the SES slope varies
    across schools. Report the variance of the slope and its correlation with
    the intercept, test the extension against the random-intercept model with
    an LR test, and say in one sentence what a school with a steep SES slope
    looks like.

# Session information

```{r}
#| label: session
sessionInfo()
```
