---
title: "\\textbf{Lab 9: Estimating and Stress-Testing an RD}"
subtitle: "SOCIOL 723 --- Social Statistics II --- Thursday, October 22"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Draw the picture first, because in an RD the picture *is* the result.
2.  Estimate with local linear regression and robust bias-corrected intervals.
3.  Choose a bandwidth, and see how much the answer depends on it.
4.  Run every validity check an RD paper should report.
5.  See in simulation what manipulation of the running variable does.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(rdrobust)
library(rddensity)

data("rdrobust_RDsenate", package = "rdrobust")
sen <- rdrobust_RDsenate |> filter(!is.na(vote))
c(n = nrow(sen), min_margin = min(sen$margin), max_margin = max(sen$margin))
```

# The design

**Question.** Does winning a Senate seat make a party more likely to win that
seat again six years later — the *incumbency advantage*?

**Running variable.** `margin`: the Democratic vote-share margin of victory in
the previous election. **Cutoff.** Zero. **Treatment.** Democratic incumbency.
**Outcome.** `vote`: the Democratic vote share in the following election.

The identification claim is that elections decided by a fraction of a percentage
point are essentially coin flips, so parties that barely won and parties that
barely lost are comparable in everything except incumbency.

# The picture comes first

```{r}
#| label: fig-rdplot
#| fig-cap: "Binned means of the following election's Democratic vote share against the previous margin of victory, with fourth-order polynomials fitted separately on each side."
rdplot(y = sen$vote, x = sen$margin, c = 0,
       x.label = "Democratic margin of victory, previous election",
       y.label = "Democratic vote share, next election",
       title = "")
```

::: callout-important
Look at the jump at zero before you compute anything. If you cannot see it here,
no amount of bandwidth tuning should convince you it is there. Conversely, a
jump this visible does not need a high-order polynomial to find it.
:::

```{r}
#| label: fig-rdplot-linear
#| fig-cap: "The same data with local linear fits inside a 15-point window. This is closer to what the estimator actually uses."
rdplot(y = sen$vote[abs(sen$margin) < 15], x = sen$margin[abs(sen$margin) < 15],
       c = 0, p = 1, nbins = c(20, 20),
       x.label = "margin", y.label = "next vote share", title = "")
```

# Estimation

```{r}
#| label: rd-main
rd <- rdrobust(y = sen$vote, x = sen$margin, c = 0)
summary(rd)
```

Three numbers to read from that output.

-   **Conventional**: the local linear estimate at the MSE-optimal bandwidth,
    with a naive standard error.
-   **Bias-Corrected**: the point estimate after subtracting an estimate of the
    leading bias term.
-   **Robust**: the confidence interval that accounts for the *uncertainty in
    that bias correction*. **This is the one to report** (Calonico, Cattaneo, and
    Titiunik 2014).

```{r}
#| label: rd-eff-n
## how many observations are actually doing the work?
c(bandwidth_left = rd$bws[1, 1], bandwidth_right = rd$bws[1, 2],
  n_left = rd$N_h[1], n_right = rd$N_h[2], n_total = sum(rd$N))
```

Note the gap between the nominal sample size and the effective one. An RD with
1,300 observations may be estimated from 400.

## Doing it by hand

`rdrobust` is a black box until you have written the estimator once.

```{r}
#| label: rd-manual
h <- rd$bws[1, 1]                       # use the same bandwidth

sub <- sen |> filter(abs(margin) <= h) |>
  mutate(D = as.integer(margin >= 0),
         kw = 1 - abs(margin) / h)      # triangular kernel

## fully interacted local linear regression: separate slopes each side
fit_manual <- lm(vote ~ D * margin, data = sub, weights = kw)

c(manual = coef(fit_manual)["D"], rdrobust_conventional = rd$coef[1])
```

The coefficient on the treatment indicator in a kernel-weighted, fully
interacted local linear regression *is* the RD estimate. Everything `rdrobust`
adds is bandwidth selection and bias correction.

## Bandwidth sensitivity

```{r}
#| label: fig-bwsens
#| fig-cap: "The RD estimate across bandwidths from 3 to 40 points, with robust 95% intervals. A credible result is flat over a wide range."
bws <- seq(3, 40, by = 1)
sens <- lapply(bws, function(b) {
  r <- rdrobust(y = sen$vote, x = sen$margin, c = 0, h = b)
  tibble::tibble(h = b, est = r$coef[3],          # bias-corrected point estimate
                 lo = r$ci[3, 1], hi = r$ci[3, 2]) # matching robust interval
}) |> bind_rows()

ggplot(sens, aes(h, est)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "navy", alpha = .15) +
  geom_line(colour = "navy", linewidth = .8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = rd$bws[1, 1], colour = "firebrick", linetype = 3) +
  labs(x = "bandwidth", y = "RD estimate") +
  theme_minimal(base_size = 10)
```

The dotted red line marks the MSE-optimal bandwidth. The estimate is stable
across a wide range — this is what a robust RD looks like.

## Why not a global polynomial?

```{r}
#| label: global-poly
poly_est <- sapply(1:8, function(p) {
  m <- lm(vote ~ poly(margin, p, raw = TRUE) * I(margin >= 0), data = sen)
  coef(m)["I(margin >= 0)TRUE"]
})
round(setNames(poly_est, paste0("degree_", 1:8)), 3)
```

The answer wanders with the polynomial degree, and every one of these estimates
uses observations 80 points from the cutoff to fit the value *at* the cutoff.
Gelman and Imbens (2019) explain why the implied weights are indefensible.

# Validity checks

## Is the density of the running variable continuous?

If parties could precisely control their vote margin, we would see bunching just
above zero.

```{r}
#| label: density-test
dens <- rddensity(X = sen$margin, c = 0)
summary(dens)
```

```{r}
#| label: fig-density
#| fig-cap: "Estimated density of the running variable on each side of the cutoff with confidence bands. Overlapping bands at zero are what we want to see."
rdplotdensity(dens, sen$margin, plotN = 100)
```

Non-rejection is consistent with the design, but it is not a validation of it:
the test has limited power, and it detects only manipulation that shows up as
bunching in the density. In settings where actors *can* manipulate the running
variable (income thresholds, test retakes) treat it as one weak signal among
several.

## Placebo cutoffs

The estimator should find nothing at thresholds where nothing happens.

```{r}
#| label: placebo
placebos <- c(-20, -15, -10, -5, 5, 10, 15, 20)
plc <- sapply(placebos, function(cc) {
  d <- if (cc < 0) sen |> filter(margin < 0) else sen |> filter(margin >= 0)
  r <- rdrobust(y = d$vote, x = d$margin, c = cc)
  c(estimate = r$coef[1], p_robust = r$pv[3])
})
round(t(plc), 3) |> `rownames<-`(paste("cutoff", placebos))
```

Note the restriction to one side of the true cutoff: otherwise a placebo test
would pick up the genuine discontinuity at zero.

## Donut hole

Drop the observations nearest the cutoff, where manipulation would be
concentrated.

```{r}
#| label: donut
donut <- sapply(c(0, 0.25, 0.5, 1), function(dd) {
  d <- sen |> filter(abs(margin) > dd)
  r <- rdrobust(y = d$vote, x = d$margin, c = 0)
  c(estimate = r$coef[1], ci_lo = r$ci[3, 1], ci_hi = r$ci[3, 2])
})
round(t(donut), 3) |> `rownames<-`(paste("hole =", c(0, 0.25, 0.5, 1)))
```

# What manipulation actually does

Simulate a world where the treatment has **no effect at all**, but where units
just below the cutoff can push themselves above it.

```{r}
#| label: manip-sim
set.seed(723)
n <- 4000
u <- rnorm(n)                              # ability: raises both X and Y
x_true <- rnorm(n) + 0.8 * u

## units with high u who are just below the cutoff push themselves over
push <- (x_true > -0.3 & x_true < 0) & (u > 0.5)
x <- ifelse(push, abs(x_true) + 0.01, x_true)

d <- as.integer(x >= 0)
y <- 0 * d + u + rnorm(n)                  # TRUE EFFECT IS ZERO

r_bad <- rdrobust(y = y, x = x, c = 0)
c(estimate = r_bad$coef[1], p_robust = r_bad$pv[3], truth = 0)
```

```{r}
#| label: manip-density
dt <- rddensity(X = x, c = 0)
c(t_statistic = dt$test$t_jk, p_value = dt$test$p_jk)
```

The RD finds a large "effect" that does not exist — and **the density test
catches it**. That is why the test belongs in every RD paper — while remembering
that manipulation which does *not* produce bunching would slip past it.

```{r}
#| label: fig-manip
#| fig-cap: "Histogram of the running variable under manipulation. The excess mass just above zero is the signature."
tibble::tibble(x) |> filter(abs(x) < 1.5) |>
  ggplot(aes(x)) +
  geom_histogram(bins = 60, fill = "navy", alpha = .7) +
  geom_vline(xintercept = 0, colour = "firebrick", linewidth = .8) +
  labs(x = "running variable", y = "count") +
  theme_minimal(base_size = 10)
```

# Fuzzy RD

When crossing the cutoff changes the *probability* of treatment, `rdrobust`
takes a `fuzzy` argument and returns the ratio estimate.

```{r}
#| label: fuzzy-sim
set.seed(1)
n <- 3000
xx <- runif(n, -1, 1)
takeup <- rbinom(n, 1, ifelse(xx >= 0, 0.75, 0.20))   # jump of 0.55
yy <- 1.5 * takeup + 0.5 * xx + rnorm(n)               # TRUE effect = 1.5

## sharp (intention-to-treat): attenuated toward zero
r_itt   <- rdrobust(y = yy, x = xx, c = 0)
## fuzzy: scaled up by the first stage
r_fuzzy <- rdrobust(y = yy, x = xx, c = 0, fuzzy = takeup)

c(ITT = r_itt$coef[1], fuzzy = r_fuzzy$coef[1], truth = 1.5)
```

The sharp estimate is the *intention-to-treat* effect of crossing the cutoff. The
fuzzy estimate divides it by the jump in take-up, exactly as in Week 7 — and it
inherits the LATE interpretation. Here the estimand is the effect for compliers
*at the cutoff*.

# Exercises

1.  **Report an RD properly.** Write a short results paragraph for the Senate
    application containing: the robust estimate and interval, the bandwidth and
    effective $n$, the density test, one placebo, and one sentence stating the
    estimand. Under 150 words.

2.  **Kernel choice.** Re-estimate with `kernel = "uniform"` and
    `kernel = "epanechnikov"`. How much does the answer move relative to its
    standard error? Which choice matters more, kernel or bandwidth?

3.  **Local polynomial order.** Compare `p = 1`, `p = 2`, and `p = 3` in
    `rdrobust`. Note both the estimate and the selected bandwidth. Why does the
    optimal bandwidth widen with $p$?

4.  **Covariates.** `rdrobust` accepts a `covs` argument. Add one and see what
    happens to the standard error and the point estimate. Under what condition is
    adding covariates to an RD legitimate?

5.  **Detecting your own manipulation.** In `manip-sim`, vary the strength of the
    push (the `u > 0.5` threshold). At what point does the density test start to
    detect it? What does that tell you about the test's power?

6.  **A real design.** Find an RD in the sociological literature. Identify the
    running variable, the cutoff, and the estimand in one sentence each. Then
    list which of the six checks from lecture the paper reports.

# Session information

```{r}
#| label: session
sessionInfo()
```
