---
title: "\\textbf{Lab 2 Exercise Solutions}"
subtitle: "SOCIOL 723, Social Statistics II"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.2, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

```{r}
#| label: packages
library(dplyr); library(ggplot2); library(sandwich)
gss <- readRDS("../Data/gss_earnings.rds")
m <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)
```

# Exercise 1: Coverage of the classical interval

Same design as the lab, with the heteroskedasticity turned up from
$0.5\exp(0.8x)$ to $0.5\exp(1.5x)$.

```{r}
#| label: ex1
set.seed(723)
sim_once <- function(n = 200, expo = 1.5) {
  x <- rnorm(n)
  e <- rnorm(n, sd = 0.5 * exp(expo * x))
  mm <- lm(I(1 + 0.5 * x + e) ~ x)
  c(b  = coef(mm)[2],
    sc = sqrt(diag(vcov(mm)))[2],
    s3 = sqrt(diag(vcovHC(mm, type = "HC3")))[2])
}
reps <- t(replicate(2000, sim_once()))
covers <- function(b, se) mean(abs(b - 0.5) / se < 1.96)
c(classical = covers(reps[, 1], reps[, 2]),
  hc3       = covers(reps[, 1], reps[, 3]))
```

Classical coverage collapses far below its nominal 95%: the error variance
now explodes where $x$ is large, exactly where the meat's weights $d_i^2$
are biggest, so the pooled $s^2$ understates the variance badly. HC3 loses
some ground too --- with heteroskedasticity this violent, the meat is
estimated from a handful of extreme observations --- but it stays far closer
to nominal. The lesson survives: the classical interval's failure is
structural, the robust interval's is finite-sample.

# Exercise 2: HC3 versus HC0 in small samples

```{r}
#| label: ex2
set.seed(723)
one_n <- function(n) {
  reps <- t(replicate(2000, {
    x <- rnorm(n)
    e <- rnorm(n, sd = 0.5 * exp(0.8 * x))
    mm <- lm(I(1 + 0.5 * x + e) ~ x)
    c(b  = coef(mm)[2],
      s0 = sqrt(diag(vcovHC(mm, type = "HC0")))[2],
      s3 = sqrt(diag(vcovHC(mm, type = "HC3")))[2])
  }))
  c(n = n, hc0 = covers(reps[, 1], reps[, 2]),
    hc3 = covers(reps[, 1], reps[, 3]))
}
round(t(sapply(c(30, 60, 120, 500), one_n)), 3)
```

At $n = 30$ HC0 under-covers noticeably while HC3's leverage rescaling buys
several points of coverage; by $n = 120$ the gap has nearly closed, and at
$n = 500$ the two are indistinguishable. That is the general pattern behind
the lecture's advice: the HC variants are a small-sample story, and all of
them coincide as $n$ grows.

# Exercise 3: Choosing the cluster level

A state-level minimum-wage law is one treatment draw shared by everyone in
the state, so the lottery runs over states: **cluster by state**. This is
coarser than the GSS's own sampling clusters (PSUs sit inside states), and
with nested lotteries the coarsest one covers both --- same-PSU pairs are
also same-state pairs, so state clustering keeps the sampling covariances
too.

If the law varies by state *and* year, the tempting answer --- cluster on
state--year cells --- is too fine: a state's errors are strongly correlated
*across* years (the Bertrand--Duflo--Mullainathan lesson), and state-year
clustering assumes those covariances away. The safe answer remains
clustering by state, which allows arbitrary correlation over time within
each state.

# Exercise 4: A statistic with no easy formula

```{r}
#| label: ex4
set.seed(723)
turning <- function(fit) -coef(fit)["exper"] / (2 * coef(fit)["I(exper^2)"])
turning(m)

B <- 5000
tp_b <- replicate(B, {
  idx <- sample(nrow(gss), replace = TRUE)
  turning(lm(lnearn ~ educ + exper + I(exper^2) + female + black,
             data = gss[idx, ]))
})

c(estimate = turning(m), se_boot = sd(tp_b))
c(normal_approx = turning(m) + c(-1.96, 1.96) * sd(tp_b))
quantile(tp_b, c(.025, .975))
```

```{r}
#| label: fig-ex4
#| fig-cap: "Bootstrap distribution of the experience-profile turning point, $B = 5{,}000$."
tibble::tibble(tp = tp_b) |>
  ggplot(aes(tp)) +
  geom_histogram(bins = 60, fill = "navy", alpha = .65) +
  geom_vline(xintercept = turning(m), colour = "firebrick") +
  labs(x = "bootstrap replicate of the turning point (years)", y = "count") +
  theme_minimal(base_size = 10)
```

The turning point lands around 36 years of experience. The histogram is
close to bell-shaped --- the denominator $2\hat\beta_{\text{exper}^2}$ is
many standard errors from zero, so the ratio behaves well --- and
accordingly the normal-approximation and percentile intervals nearly agree.
When they agree, either is defensible; the percentile interval is still the
safer habit, because it would have followed the skew had there been any.
