---
title: "\\textbf{Lab 2: Simulation and Robust Inference}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, September 3"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator*{\argmin}{arg\,min}
        \DeclareMathOperator*{\argmax}{arg\,max}
        \DeclareMathOperator{\trace}{tr}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\En}{\mathbb{E}_n}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  See a sampling distribution by *building* one, through simulation.
2.  Watch classical standard errors fail under heteroskedasticity, and watch robust standard errors fix them.
3.  Compute HC0--HC3 and cluster-robust standard errors in `R`, by hand and with packages.
4.  Bootstrap, from first principles and with `boot`, including the cluster bootstrap.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(sandwich)      # vcovHC(), vcovCL()
library(lmtest)        # coeftest()
library(boot)          # bootstrap
library(fixest)        # fast fixed effects with clustered SEs
```

```{r}
#| label: data
gss <- readRDS("../Data/gss_earnings.rds")
c(n = nrow(gss), psu = n_distinct(gss$psu),
  stratum = n_distinct(gss$stratum), region = n_distinct(gss$region))
```

# Building a sampling distribution

The fastest way to understand a standard error is to construct one.

```{r}
#| label: sampling-dist
set.seed(723)

## a population we control completely
sim_once <- function(n = 200, beta1 = 0.5, hetero = FALSE) {
  x <- rnorm(n, mean = 0, sd = 1)
  sd_i <- if (hetero) 0.5 * exp(0.8 * x) else 1     # error sd
  e <- rnorm(n, sd = sd_i)
  y <- 1 + beta1 * x + e
  m <- lm(y ~ x)
  c(b = coef(m)[2],
    se_classical = sqrt(diag(vcov(m)))[2],
    se_hc3 = sqrt(diag(vcovHC(m, type = "HC3")))[2])
}

reps_homo <- t(replicate(4000, sim_once(hetero = FALSE)))
head(round(reps_homo, 4), 3)
```

```{r}
#| label: fig-sampdist
#| fig-cap: "Sampling distribution of $\\hat\\beta_1$ over 4,000 samples of size 200. The true value is 0.5."
tibble::tibble(b = reps_homo[, 1]) |>
  ggplot(aes(b)) +
  geom_histogram(bins = 50, fill = "navy", alpha = .65) +
  geom_vline(xintercept = 0.5, colour = "firebrick", linewidth = .7) +
  labs(x = expression(hat(beta)[1]), y = "count") +
  theme_minimal(base_size = 10)
```

The *true* standard error is the standard deviation of this histogram. Compare it to what the two formulas report on average:

```{r}
#| label: se-compare-homo
c(truth = sd(reps_homo[, "b.x"]),
  mean_classical = mean(reps_homo[, "se_classical.x"]),
  mean_hc3 = mean(reps_homo[, "se_hc3.x"]))
```

Under homoskedasticity both are right; the robust one is very slightly noisier. That is the small price we pay for insurance. Now turn heteroskedasticity on:

```{r}
#| label: se-compare-hetero
reps_het <- t(replicate(4000, sim_once(hetero = TRUE)))
c(truth = sd(reps_het[, "b.x"]),
  mean_classical = mean(reps_het[, "se_classical.x"]),
  mean_hc3 = mean(reps_het[, "se_hc3.x"]))
```

The classical formula is badly wrong; HC3 tracks the truth.

# Robust standard errors on real data

## By hand

The scalar formula from lecture, for the bivariate case:

$$\hat{V}_{\text{HC0}}(\hat\beta_1)
= \frac{\sum_i d_i^2 \hat e_i^2}{\left(\sum_i d_i^2\right)^2},
\qquad d_i = X_i - \bar X .$$

```{r}
#| label: hc0-by-hand
m_biv <- lm(lnearn ~ educ, data = gss)
d  <- gss$educ - mean(gss$educ)
eh <- residuals(m_biv)

se_hc0_manual <- sqrt(sum(d^2 * eh^2) / (sum(d^2))^2)
se_hc0_pkg    <- sqrt(diag(vcovHC(m_biv, type = "HC0")))["educ"]
se_classical  <- sqrt(diag(vcov(m_biv)))["educ"]

round(c(classical = se_classical, hc0_manual = se_hc0_manual,
        hc0_sandwich = se_hc0_pkg), 6)
```

The general matrix version, also by hand:

```{r}
#| label: hc0-matrix
m <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)
X <- model.matrix(m); e <- residuals(m)
bread <- solve(crossprod(X))
meat  <- crossprod(X * e)          # = sum_i e_i^2 x_i x_i'
V_hc0 <- bread %*% meat %*% bread
round(cbind(manual = sqrt(diag(V_hc0)),
            sandwich = sqrt(diag(vcovHC(m, type = "HC0")))), 6)
```

## HC0 through HC3

```{r}
#| label: hc-family
ses <- sapply(c("const", "HC0", "HC1", "HC2", "HC3"),
              function(ty) sqrt(diag(vcovHC(m, type = ty))))
round(ses, 5)
```

```{r}
#| label: coeftest
coeftest(m, vcov = vcovHC(m, type = "HC3"))
```

With $n = `r nrow(gss)`$ the four robust variants are nearly identical. Rerun this on a subsample of 100 and the gaps open up:

```{r}
#| label: small-n
set.seed(1)
small <- gss[sample(nrow(gss), 100), ]
m_small <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = small)
round(sapply(c("const", "HC0", "HC1", "HC2", "HC3"),
             function(ty) sqrt(diag(vcovHC(m_small, type = ty)))["educ"]), 5)
```

## The convenience tool: `stargazer`

For a paper-style table, hand `stargazer` the model once per column and a
list of standard errors; it recomputes the stars from whatever SEs you
supply.

```{r}
#| label: stargazer
library(stargazer)
se_hc1 <- sqrt(diag(vcovHC(m, type = "HC1")))
se_hc3 <- sqrt(diag(vcovHC(m, type = "HC3")))
stargazer(m, m, m,
          se = list(NULL, se_hc1, se_hc3),
          column.labels = c("Classical", "HC1", "HC3"),
          keep = c("educ", "female", "black"),
          keep.stat = c("n", "rsq"),
          digits = 4,
          type = "text", no.space = TRUE)
```

Same coefficients in every column, only the standard errors move: the whole
lecture in one table. For a paper, switch `type = "text"` to
`type = "latex"` and always say in the table note which variance estimator
each column uses.

# Clustering

The GSS is a multi-stage area probability sample: it draws primary sampling units (`psu`) within strata (`stratum`), then households within PSUs. Respondents in the same PSU live in the same neighbourhood, so their errors are correlated.

## By hand: sum within clusters first, then square

Lecture's boxed scalar formula, plus the finite-sample scaling software
applies by default (the cluster analogue of HC1's degrees-of-freedom
correction; Stata's default too):

$$\hat V_{\text{CR1}}(\hat\beta_1)
= \underbrace{\frac{G}{G-1}\cdot\frac{n-1}{n-k}}_{c}\ 
  \frac{\sum_{g} S_g^{\,2}}{\big(\sum_i d_i^2\big)^2},
\qquad S_g = \sum_{i \in g} d_i \hat e_i .$$

```{r}
#| label: cr-by-hand
Sg <- tapply(d * eh, gss$psu, sum)          # one number per cluster
G <- n_distinct(gss$psu); n <- nrow(gss); k <- 2
c1 <- G / (G - 1) * (n - 1) / (n - k)
se_cr_manual <- sqrt(c1 * sum(Sg^2) / (sum(d^2))^2)
se_cr_pkg    <- sqrt(vcovCL(m_biv, cluster = ~ psu)["educ", "educ"])
round(c(scaling = c1, cr_manual = se_cr_manual,
        cr_sandwich = se_cr_pkg, hc0 = se_hc0_manual), 6)
```

Sum within clusters first, then square. With `r n_distinct(gss$psu)`
clusters the scaling $c$ is negligible; it matters when clusters are few.
Up to that factor, one observation per cluster collapses this to the robust
formula: the two are one idea.

## Which level for the GSS?

The GSS is a multi-stage sample: PSUs are the *sampled* units, and no
regressor in this model varies at any group level. So the design says
**cluster on `psu`**, full stop. Coarser levels would allow extra dependence
the design gives no reason to expect, at the price of a noisier variance
estimate.

```{r}
#| label: stargazer-cluster
se_psu <- sqrt(diag(vcovCL(m, cluster = ~ psu)))
stargazer(m, m, m,
          se = list(NULL, se_hc3, se_psu),
          column.labels = c("Classical", "HC3", "Cluster PSU"),
          keep = c("educ", "female", "black"),
          keep.stat = c("n", "rsq"),
          type = "text", no.space = TRUE)
```

Clustering moves the standard errors only modestly here: GSS PSUs hold few
respondents each, and `educ` varies mostly *within* PSUs, so the surviving
cross terms are few and small.

# The bootstrap

## By hand

```{r}
#| label: boot-manual
set.seed(723)
B <- 2000
boot_b <- replicate(B, {
  idx <- sample(nrow(gss), replace = TRUE)
  coef(lm(lnearn ~ educ + exper + I(exper^2) + female + black,
          data = gss[idx, ]))["educ"]
})

c(estimate = coef(m)["educ"],
  se_boot  = sd(boot_b),
  se_hc3   = sqrt(diag(vcovHC(m, type = "HC3")))["educ"])

quantile(boot_b, c(.025, .975))       # percentile interval
```

The pairs bootstrap resamples $(Y_i, X_i)$ jointly, so it makes no assumption about $\sigma_i^2$. Asymptotically it targets the same sandwich variance that the HC estimators do, which is why it lands in the same neighbourhood, the small gaps among HC0/HC3/bootstrap here are finite-sample, not conceptual.

```{r}
#| label: fig-boot
#| fig-cap: "Bootstrap distribution of the coefficient on years of schooling, $B = 2{,}000$."
tibble::tibble(b = boot_b) |>
  ggplot(aes(b)) +
  geom_histogram(bins = 50, fill = "navy", alpha = .65) +
  geom_vline(xintercept = coef(m)["educ"], colour = "firebrick") +
  labs(x = "bootstrap replicate of the coefficient on educ", y = "count") +
  theme_minimal(base_size = 10)
```

The histogram is bell-shaped here, so the normal-approximation and percentile intervals will nearly coincide; when the histogram is skewed or lumpy, they disagree, and the shape is the reason to prefer the percentile side.

## With the `boot` package

```{r}
#| label: boot-pkg
b_fun <- function(data, idx) {
  unname(coef(lm(lnearn ~ educ + exper + I(exper^2) + female + black,
                 data = data[idx, ]))["educ"])
}
set.seed(723)
bo <- boot(gss, b_fun, R = 2000)
boot.ci(bo, type = c("norm", "perc"))
```

## The cluster bootstrap

Resample **clusters**, not individuals:

```{r}
#| label: boot-cluster
set.seed(723)
psus <- unique(gss$psu)
boot_cl <- replicate(1000, {
  draw <- sample(psus, length(psus), replace = TRUE)
  dat  <- do.call(rbind, lapply(draw, function(p) gss[gss$psu == p, ]))
  coef(lm(lnearn ~ educ + exper + I(exper^2) + female + black,
          data = dat))["educ"]
})
c(cluster_boot = sd(boot_cl),
  analytic_CR1 = sqrt(diag(vcovCL(m, cluster = ~ psu)))["educ"])
```

::: callout-warning
If you had resampled *individuals* here, the bootstrap would have reproduced the too-small non-clustered standard error just as faithfully as the analytic formula does. The bootstrap does not know your design; you have to tell it.
:::

# A convenient shortcut: `fixest`

For everyday work, `fixest::feols()` computes clustered standard errors on the fly and is very fast.

```{r}
#| label: fixest
fe <- feols(lnearn ~ educ + exper + I(exper^2) + female + black | year,
            data = gss, cluster = ~ psu)
summary(fe)
```

The `| year` term absorbs year fixed effects (FWL under the hood, as in Lab 1), and `cluster = ~ psu` supplies the cluster-robust variance.

# Exercises

1.  **Coverage of the classical interval.** Repeat the coverage simulation with error standard deviation $0.5\exp(1.5 x_i)$ instead of $0.5\exp(0.8 x_i)$. How bad does classical coverage get? Does HC3 hold up?

2.  **HC3 versus HC0 in small samples.** For $n \in \{30, 60, 120, 500\}$, simulate 2,000 heteroskedastic datasets and record the coverage of intervals built from HC0 and from HC3. At what $n$ does the difference stop mattering?

4.  **Choosing the cluster level.** Suppose you wanted to estimate the effect of a state-level minimum-wage law on individual earnings, using individual GSS data. At what level would you cluster, and why? What if the law varied by state *and* year?

5.  **A statistic with no easy formula.** Estimate the experience profile's turning point $-\hat\beta_{\text{exper}} / (2\hat\beta_{\text{exper}^2})$ and bootstrap it ($B = 5{,}000$). Plot the histogram of the replicates: is it bell-shaped? Report both the normal-approximation and the percentile interval, and say which you would report and why.

# Session information

```{r}
#| label: session
sessionInfo()
```