---
title: "\\textbf{Lab 7: Two-Stage Least Squares and Its Diagnostics}"
subtitle: "SOCIOL 723 --- Social Statistics II --- Thursday, October 8"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  Build the IV estimator by hand and see that it is a ratio of two reduced-form
    quantities.
2.  Run 2SLS properly, and understand why running the two stages manually gives
    the wrong standard errors.
3.  Read every diagnostic an IV paper should report.
4.  Watch weak instruments destroy 2SLS in simulation.
5.  Characterize the compliers.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(ivreg)
library(fixest)
library(sandwich); library(lmtest)

data("SchoolingReturns", package = "ivreg")
card <- SchoolingReturns |>
  mutate(lwage = log(wage), near4 = as.integer(nearcollege4 != "none"))
c(n = nrow(card), share_near_4yr = mean(card$near4))
table(card$nearcollege4)
```

# The design

The question is the causal return to a year of schooling. Schooling is
endogenous: unmeasured ability raises both schooling and wages, so OLS is biased
upward; measurement error pushes the other way.

The instrument, following Card (1995), is **growing up near a four-year
college**. The argument for relevance is that proximity lowers the cost of
attending. The argument for exclusion is that, conditional on region and urban
status, where your parents happened to live does not affect your wage except
through the schooling it made cheaper.

::: callout-warning
That exclusion argument is contestable, and it has been contested — college
towns differ from other places in labour markets, school quality, and family
background. Part of today's job is learning to interrogate it rather than accept
it.
:::

# By hand: the Wald estimator

With a binary instrument, IV is the ratio of two differences in means.

```{r}
#| label: wald
rf <- mean(card$lwage[card$near4 == 1]) - mean(card$lwage[card$near4 == 0])
fs <- mean(card$education[card$near4 == 1]) - mean(card$education[card$near4 == 0])

c(reduced_form = rf, first_stage = fs, wald_ratio = rf / fs)
```

```{r}
#| label: wald-reg
## the same numbers as regressions
c(reduced_form = coef(lm(lwage ~ near4, data = card))["near4"],
  first_stage  = coef(lm(education ~ near4, data = card))["near4"],
  iv           = coef(ivreg(lwage ~ education | near4, data = card))["education"])
```

::: callout-note
Notice the two ingredients. The reduced form is small (about 0.16 log points) and
so is the first stage (about 0.83 years). Dividing one small number by another is
exactly the operation that makes IV fragile — and once we add covariates, both
shrink further.
:::

# 2SLS with covariates

```{r}
#| label: models
f_ols <- lwage ~ education + experience + I(experience^2) + ethnicity +
  smsa + south

m_ols <- lm(f_ols, data = card)

m_iv <- ivreg(lwage ~ education + experience + I(experience^2) + ethnicity +
                smsa + south |
                near4 + age + I(age^2) + ethnicity + smsa + south,
              data = card)

round(cbind(OLS = coef(m_ols)[1:3], IV = coef(m_iv)[1:3]), 4)
```

Note the instrument set: because `experience` is a function of `education`
($\text{exper} = \text{age} - \text{educ} - 6$), it is endogenous too, and we
instrument it with `age`. Getting this bookkeeping right matters — every
endogenous regressor needs its own instrument.

```{r}
#| label: iv-summary
summary(m_iv, vcov = sandwich, diagnostics = TRUE)
```

The IV estimate is *larger* than OLS. That is the standard finding with this
instrument and it is worth pausing on: if ability bias were the only problem, IV
should come out *smaller*. Candidate explanations are measurement error in
self-reported schooling, and the fact that the LATE applies to compliers whose
return may genuinely be higher.

## Why not run the two stages by hand?

To make the comparison clean, use a specification with a **single** endogenous
regressor: instrument `education` with `near4`, and treat `age` (rather than
`experience`) as the exogenous age control so nothing else is endogenous.

```{r}
#| label: simple-iv
m_iv1 <- ivreg(lwage ~ education + age + I(age^2) + ethnicity + smsa + south |
                 near4 + age + I(age^2) + ethnicity + smsa + south, data = card)

stage1 <- lm(education ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
card$edu_hat <- fitted(stage1)
stage2 <- lm(lwage ~ edu_hat + age + I(age^2) + ethnicity + smsa + south,
             data = card)

round(cbind(
  coefficient = c(manual = coef(stage2)["edu_hat"], ivreg = coef(m_iv1)["education"]),
  std_error = c(manual = sqrt(diag(vcov(stage2)))["edu_hat"],
                ivreg = sqrt(diag(vcov(m_iv1)))["education"])
), 5)
```

The **coefficients agree** but the **standard errors do not**. The manual second
stage treats $\hat D$ as if it were data rather than an estimate, so it
understates the uncertainty. Always use `ivreg()` or `feols()`.

## The same thing in `fixest`

```{r}
#| label: fixest-iv
m_fe <- feols(lwage ~ age + I(age^2) + ethnicity + smsa + south |
                education ~ near4,
              data = card, vcov = "hetero")
summary(m_fe)
```

```{r}
#| label: fixest-stage1
fitstat(m_fe, c("ivf", "ivwald", "wh"))
```

# Diagnostics

## The first stage

```{r}
#| label: firststage
fs_mod <- lm(education ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
round(coef(summary(fs_mod))["near4", ], 4)

## robust F on the excluded instrument
car::linearHypothesis(fs_mod, "near4 = 0", vcov. = vcovHC(fs_mod, "HC1"))
```

Growing up near a four-year college is associated with about a third of a year
more schooling once covariates are included, and the robust $F$ is right at the
traditional threshold of 10.

That is **not reassuring**. Recall from lecture that the threshold of 10 is now
known to be far too lax: Lee, McCrary, Moreira, and Porter (2022) show that a
conventional 5% $t$-test on the 2SLS estimate has correct size only when the
first-stage $F$ exceeds roughly **104**. At $F \approx 10$ we are deep in the
weak-instrument regime, and the conventional standard error is badly
optimistic.

```{r}
#| label: tF
## the Lee et al. tF adjustment factor, interpolated from their Table 3
## Lee et al. (2022, Table 3) give ADJUSTED CRITICAL VALUES, not SE multipliers.
## The implied interval-widening factor is (critical value) / 1.96.
tF_crit <- function(F) {
  Fgrid <- c(4, 6, 8, 10, 12, 15, 20, 30, 40, 60, 100, 104.7, 200)
  crit  <- c(Inf, 18.66, 5.72, 3.43, 3.02, 2.71, 2.54, 2.31, 2.20, 2.11, 2.01,
             1.96, 1.96)
  approx(Fgrid, crit, xout = F, rule = 2)$y
}
Fstat <- car::linearHypothesis(fs_mod, "near4 = 0",
                               vcov. = vcovHC(fs_mod, "HC1"))$F[2]
se_iv <- sqrt(diag(vcovHC(m_iv1, "HC1")))["education"]
b_iv  <- coef(m_iv1)["education"]
crit  <- tF_crit(Fstat)

c(F = Fstat, adjusted_crit_value = crit, widening_factor = crit / 1.96,
  estimate = b_iv, conventional_se = se_iv,
  conventional_t = b_iv / se_iv,
  significant_at_5pct = abs(b_iv / se_iv) > crit)
```

At $F \approx 10$ the honest critical value is about 3.4 rather than 1.96 --- so
the interval is roughly 1.75 times wider, and a $t$-statistic that looked
comfortably significant may not clear the bar. This is the single most important
diagnostic in the lab.

Note also that `tF` is designed for the **just-identified, single-endogenous-
regressor** case, which is why we built `m_iv1` above.

## The reduced form

```{r}
#| label: reduced-form
rf_mod <- lm(lwage ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
round(coef(summary(rf_mod))["near4", ], 4)
```

::: callout-tip
**Always report the reduced form.** It is the one regression in an IV design
that requires no instrument-strength assumption at all. Here the adjusted
reduced-form coefficient is about 0.032 with a standard error of 0.016 — a
$t$ of roughly 2. That is the entire evidentiary basis for the IV estimate, and
it is thin. If the instrument has no detectable effect on the outcome, the IV
estimate is a small number divided by another small number, and its apparent
precision is an artefact.
:::

## Overidentification

```{r}
#| label: overid
m_over <- ivreg(lwage ~ education + experience + I(experience^2) + ethnicity +
                  smsa + south |
                  near4 + nearcollege2 + age + I(age^2) + ethnicity + smsa + south,
                data = card)
summary(m_over, diagnostics = TRUE)$diagnostics
```

Read the Sargan row carefully. It also assumes homoskedasticity, unlike
everything else in this lab --- the heteroskedasticity-robust analogue is
Hansen's $J$. And non-rejection is weak evidence either way: it is entirely
consistent with both instruments being invalid in the same direction.

# Weak instruments, by simulation

The cleanest way to see the danger is to build a world where you know the answer.

```{r}
#| label: weak-sim
set.seed(723)

sim_iv <- function(n = 500, pi = 0.5, rho = 0.6, beta = 1) {
  z <- rnorm(n)
  u <- rnorm(n)                               # confounder
  d <- pi * z + rho * u + rnorm(n)            # pi = instrument strength
  y <- beta * d + u + rnorm(n)                # TRUE effect is 1
  m <- ivreg(y ~ d | z)
  fs <- summary(lm(d ~ z))$fstatistic[1]
  c(iv = coef(m)["d"], ols = coef(lm(y ~ d))["d"], F = fs)
}

strengths <- c(strong = 0.5, moderate = 0.15, weak = 0.05)
res <- lapply(strengths, function(p) t(replicate(600, sim_iv(pi = p))))

sapply(res, function(r) c(
  median_F = median(r[, "F.value"]),
  mean_IV  = mean(r[, "iv.d"]),
  median_IV = median(r[, "iv.d"]),
  mean_OLS = mean(r[, "ols.d"])
)) |> round(3)
```

```{r}
#| label: fig-weak
#| fig-cap: "Sampling distribution of the 2SLS estimate at three instrument strengths, truncated to [-5, 5] for display. The truth is 1 (dashed). As the instrument weakens, 2SLS collapses toward OLS and develops very heavy tails -- see the share truncated, reported below."
bind_rows(lapply(names(res), function(nm)
  tibble::tibble(est = res[[nm]][, "iv.d"], strength = nm))) |>
  filter(abs(est) < 5) |>
  ggplot(aes(est)) +
  geom_histogram(bins = 60, fill = "navy", alpha = .7) +
  geom_vline(xintercept = 1, linetype = 2) +
  facet_wrap(~ factor(strength, levels = c("strong", "moderate", "weak")),
             scales = "free_y") +
  labs(x = expression(hat(beta)[2*SLS]), y = "count") +
  theme_minimal(base_size = 9)
```

```{r}
#| label: tail-share
sapply(res, function(r) mean(abs(r[, "iv.d"]) >= 5))   # share outside the plot
```

Three lessons:

-   With a **strong** instrument, 2SLS is centred on the truth and roughly normal.
-   With a **weak** instrument, the distribution is badly skewed and shifted
    toward the OLS probability limit. IV is biased *toward the thing it was
    supposed to fix*.
-   The tails are heavy enough that the mean is not even a useful summary.
    Conventional standard errors are hopeless here.

```{r}
#| label: coverage-weak
## coverage of nominal 95% intervals
cov_check <- function(pi) {
  out <- replicate(600, {
    z <- rnorm(500); u <- rnorm(500)
    d <- pi * z + 0.6 * u + rnorm(500); y <- 1 * d + u + rnorm(500)
    m <- ivreg(y ~ d | z)
    ci <- confint(m)["d", ]
    ci[1] < 1 & 1 < ci[2]
  })
  mean(out)
}
sapply(strengths, cov_check)
```

A nominal 95% interval that covers 60% of the time is not a 95% interval.

# Who are the compliers?

The LATE applies to people whose schooling was changed by college proximity. We
cannot name them, but we can describe them.

```{r}
#| label: complier-share
## complier share = first stage on a binary treatment
card2 <- card |> mutate(some_college = as.integer(education >= 13))
fs_bin <- lm(some_college ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card2)
round(coef(summary(fs_bin))["near4", ], 4)
```

About 5 percent of the sample are compliers on this definition — that is, roughly
5 percent attend some college *because* they grew up near one.

```{r}
#| label: complier-profile
## Abadie's kappa weights let us describe complier characteristics
pz <- fitted(glm(near4 ~ age + I(age^2) + ethnicity + smsa + south,
                 data = card2, family = binomial))
kappa <- 1 - card2$some_college * (1 - card2$near4) / (1 - pz) -
  (1 - card2$some_college) * card2$near4 / pz

profile <- function(x) c(all = mean(x), compliers = sum(kappa * x) / sum(kappa))
round(rbind(
  father_educ = profile(card2$feducation |> tidyr::replace_na(mean(card2$feducation, na.rm = TRUE))),
  south = profile(as.integer(card2$south == "yes")),
  smsa  = profile(as.integer(card2$smsa == "yes")),
  black = profile(as.integer(card2$ethnicity == "afam"))
), 3)
```

::: callout-important
The complier profile is the interpretive heart of an IV paper. If the compliers
are disproportionately from low-education families, then the LATE is a return to
college **for people at the margin of attending** — which is often the
policy-relevant group, but is emphatically not the ATE.
:::

# Exercises

1.  **The bookkeeping.** Refit the IV model treating `experience` as exogenous
    (i.e. not instrumenting it). How much does the education coefficient move?
    Explain why `experience` cannot be exogenous if `education` is not.

2.  **Reduced form first.** Suppose a referee reports only the 2SLS estimate and
    its $t$-statistic. Compute the reduced form and first stage and write two
    sentences a referee could use to evaluate whether the design is credible.

3.  **A placebo.** Growing up near a college should not affect variables
    determined before college attendance. Regress `feducation`, `meducation`, and
    `library14` on `near4` with the same controls. Do any of them move? What
    would it mean for exclusion if they did?

4.  **Anderson–Rubin.** Write a function that, for a grid of $\beta_0$ values,
    regresses $Y - \beta_0 D$ on the instrument and controls and records the
    $p$-value on the instrument. Plot it, and read off the 95% AR confidence set.
    Compare with the conventional 2SLS interval.

5.  **Bias amplification with weak instruments.** In `sim_iv()`, add a small
    direct effect of $Z$ on $Y$ (violating exclusion by, say, 0.05). Show that
    the resulting bias in 2SLS grows as the instrument weakens, and compare it
    with the OLS bias.

6.  **Many weak instruments.** Add 20 pure-noise instruments to the strong
    design. What happens to the first-stage $F$, and to the 2SLS estimate?
    Explain in terms of overfitting the first stage.

# Session information

```{r}
#| label: session
sessionInfo()
```
