---
title: "\\textbf{Lab 7: Two-Stage Least Squares and Its Diagnostics}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, October 8"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\},breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,breaksymbolleft={},breaksymbolright={}}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.3, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999, width = 76)
```

# Goals

1.  Build the IV estimator by hand and see that it is a ratio of two reduced-form
    quantities.
2.  Run 2SLS properly, and understand why running the two stages manually gives
    the wrong standard errors.
3.  Read the diagnostics an IV paper should report: the first-stage $F$, the
    reduced form, and the Anderson--Rubin test.
4.  See, in simulation, what a weak instrument does to the 2SLS output R prints.
5.  Characterize the compliers.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(ivreg)                  ## ivreg(): 2SLS
library(fixest)                 ## feols(): 2SLS with robust SEs and fixed effects
library(sandwich); library(lmtest)
library(ivDiag)                 ## AR_test(): Anderson-Rubin test and interval
library(ivdesc)                 ## ivdesc(): complier profiles

## Card's (1995) NLS Young Men extract; log wage and a binary instrument
data("SchoolingReturns", package = "ivreg")
card <- SchoolingReturns |>
  mutate(lwage = log(wage),
         near4 = as.integer(nearcollege4 != "none"),   ## grew up near a 4-year college
         age2  = age^2,
         black = as.integer(ethnicity == "afam"),
         smsa_yes  = as.integer(smsa == "yes"),
         south_yes = as.integer(south == "yes"))
c(n = nrow(card), share_near_4yr = mean(card$near4))
table(card$nearcollege4)
```

# The design

The question is the causal return to a year of schooling. Schooling is
endogenous: unmeasured ability raises both schooling and wages, so OLS is biased
upward; measurement error pushes the other way.

The instrument, following Card (1995), is **growing up near a four-year
college**. The argument for relevance is that proximity lowers the cost of
attending. The argument for exclusion is that, conditional on region and urban
status, where your parents happened to live does not affect your wage except
through the schooling it made cheaper.

::: callout-warning
That exclusion argument is contestable, and it has been contested — college
towns differ from other places in labour markets, school quality, and family
background. Part of today's job is learning to interrogate it rather than accept
it.
:::

# By hand: the Wald estimator

With a binary instrument, IV is the ratio of two differences in means.

```{r}
#| label: wald
## reduced form: the instrument's effect on the outcome
rf <- mean(card$lwage[card$near4 == 1]) - mean(card$lwage[card$near4 == 0])
## first stage: the instrument's effect on the treatment
fs <- mean(card$education[card$near4 == 1]) - mean(card$education[card$near4 == 0])

c(reduced_form = rf, first_stage = fs, wald_ratio = rf / fs)
```

```{r}
#| label: wald-reg
## the same numbers as regressions
c(reduced_form = coef(lm(lwage ~ near4, data = card))["near4"],
  first_stage  = coef(lm(education ~ near4, data = card))["near4"],
  iv           = coef(ivreg(lwage ~ education | near4, data = card))["education"])
```

::: callout-note
Notice the two ingredients. The reduced form is small (about 0.16 log points) and
so is the first stage (about 0.83 years). Dividing one small number by another is
exactly the operation that makes IV fragile — and once we add covariates, both
shrink further.
:::

# 2SLS with covariates

The controls, used throughout the lab, are age and its square, race, living in a
metropolitan area, and living in the South. Age is fixed at birth, so it is
exogenous. We do **not** use potential experience: it is computed from schooling
($\text{exper} = \text{age} - \text{educ} - 6$), so it would be endogenous too
and would need an instrument of its own. With age, the model has one endogenous
regressor (`education`) and one excluded instrument (`near4`): it is just
identified.

```{r}
#| label: models
## OLS with the controls
m_ols <- lm(lwage ~ education + age + I(age^2) + ethnicity + smsa + south,
            data = card)

## 2SLS: outcome ~ regressors | instruments
## the exogenous controls appear on BOTH sides of the bar;
## near4 replaces education on the right
m_iv <- ivreg(lwage ~ education + age + I(age^2) + ethnicity + smsa + south |
                near4 + age + I(age^2) + ethnicity + smsa + south,
              data = card)

round(cbind(OLS = coef(m_ols)[1:2], IV = coef(m_iv)[1:2]), 4)
```

```{r}
#| label: iv-summary
## heteroskedasticity-robust standard errors; diagnostics are read in section 5
summary(m_iv, vcov = sandwich, diagnostics = FALSE)
```

The IV estimate is *larger* than OLS. That is the standard finding with this
instrument and it is worth pausing on: if ability bias were the only problem, IV
should come out *smaller*. Candidate explanations are measurement error in
self-reported schooling, and the fact that the LATE applies to compliers whose
return may genuinely be higher.

## Why not run the two stages by hand?

```{r}
#| label: by-hand
## stage 1: education on the instrument and the controls
stage1 <- lm(education ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
card$edu_hat <- fitted(stage1)
## stage 2: the outcome on the fitted education and the same controls
stage2 <- lm(lwage ~ edu_hat + age + I(age^2) + ethnicity + smsa + south,
             data = card)

## same coefficient, different standard error
round(cbind(
  coefficient = c(manual = coef(stage2)["edu_hat"], ivreg = coef(m_iv)["education"]),
  std_error = c(manual = sqrt(diag(vcov(stage2)))["edu_hat"],
                ivreg = sqrt(diag(vcov(m_iv)))["education"])
), 5)
```

The **coefficients agree** but the **standard errors do not**. The manual second
stage computes its residuals with $\hat D$ instead of the actual $D$, so it
estimates the wrong error variance. Always use `ivreg()` or `feols()`.

## The same thing in `fixest`

```{r}
#| label: fixest-iv
## feols syntax: outcome ~ controls | endogenous ~ instrument
m_fe <- feols(lwage ~ age + I(age^2) + ethnicity + smsa + south |
                education ~ near4,
              data = card, vcov = "hetero")
## fitstat = ~ ivwald: show only the robust first-stage test under the table
print(m_fe, fitstat = ~ ivwald)
```

```{r}
#| label: fixest-stage1
## two versions of the first-stage test of the instrument
fitstat(m_fe, c("ivf", "ivwald"))
```

How to read the two lines:

-   **`F-test (1st stage)`** (`ivf`) tests whether the excluded instrument's
    coefficient in the first stage is zero, using the classical formula that
    assumes homoskedastic errors.
-   **`Wald (1st stage)`** (`ivwald`) tests the same hypothesis with the
    model's own variance estimate, here heteroskedasticity-robust. With one
    instrument it is the squared robust $t$ of `near4` in the first stage.
    Report this one when your standard errors are robust or clustered.

## Several endogenous regressors

When a model has more than one endogenous regressor, it needs at least as many
excluded instruments. In `feols()`, list the endogenous regressors on the left of
`~` in the last part of the formula and the instruments on the right. The order
in either list does not matter: 2SLS does not pair an instrument with a
regressor. Every endogenous regressor is regressed on *all* the instruments (and
the controls), one first stage each, and the second stage uses all the fitted
values together. What matters is that the instruments move the regressors in
different directions. Card's original specification is the classic example: it controls
for potential experience, which is endogenous because it is built from
schooling, and instruments experience and its square with age and its square.

```{r}
#| label: two-endog
## syntax: outcome ~ exogenous controls | endogenous1 + endogenous2 ~ z1 + z2
card <- card |> mutate(exper2 = experience^2)
m_two <- feols(lwage ~ ethnicity + smsa + south |
                 education + experience + exper2 ~ near4 + age + age2,
               data = card, vcov = "hetero")
print(m_two, fitstat = ~ ivwald)   ## second stage
summary(m_two, stage = 1)     ## one first stage per endogenous regressor
```

Notice in the first stages that `near4`, `age`, and `age2` all appear in each
one: the education first stage is not "education on `near4`".

With several endogenous regressors, a strong first-stage $F$ for each one
separately does not guarantee the instruments can tell them apart; papers report
the Sanderson--Windmeijer conditional $F$ instead.
The rest of this lab stays with one endogenous regressor.

# Diagnostics

## The first stage

```{r}
#| label: firststage
## the first stage with the same controls as the second stage
fs_mod <- lm(education ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
round(coef(summary(fs_mod))["near4", ], 4)

## robust F on the excluded instrument: the weak-instrument test
car::linearHypothesis(fs_mod, "near4 = 0", vcov. = vcovHC(fs_mod, "HC1"))
```

Growing up near a four-year college is associated with about a third of a year
more schooling once covariates are included, and the robust $F$ is right at the
traditional threshold of 10.

::: callout-note
**Reading the weak-instrument test.** `ivreg`'s `summary(..., diagnostics =
TRUE)` labels it "Weak instruments"; `fixest` labels it "F-test (1st stage)" or
"Wald (1st stage)". All three are the same idea: an $F$ test that the excluded
instruments have zero coefficients in the first stage. With one instrument,
$F = t^2$.

-   **Read the $F$, not the $p$-value.** The $p$-value tests whether the
    instrument is *irrelevant*, not whether it is *strong enough*. Here
    $p = 0.001$, and the instrument is still weak.
-   The rule of thumb is $F > 10$ (Staiger and Stock 1997). The conventional
    $t$-test has its stated size only when $F > 104$ (Lee et al. 2022).

`summary(..., diagnostics = TRUE)` also prints two other tests. **Wu--Hausman**
asks whether the 2SLS and OLS estimates differ by more than chance (a test of
endogeneity). **Sargan** asks whether several instruments agree with each other;
it needs more instruments than endogenous regressors, so it is `NA` in a
just-identified model. Neither is part of this course, and we do not use them.
:::

That $F$ is **not reassuring**. Recall from lecture that the threshold of 10 is
now known to be far too lax: Lee, McCrary, Moreira, and Porter (2022) show that a
conventional 5% $t$-test on the 2SLS estimate has correct size only when the
first-stage $F$ exceeds roughly **104**. At $F \approx 10$ the conventional
standard error cannot be trusted, which is why we test with Anderson--Rubin
below.

## The reduced form

```{r}
#| label: reduced-form
## the outcome on the instrument and the controls, robust SE
rf_mod <- lm(lwage ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card)
round(coeftest(rf_mod, vcov = vcovHC(rf_mod, "HC1"))["near4", ], 4)
```

::: callout-tip
**Always report the reduced form.** It is the one regression in an IV design
that requires no instrument-strength assumption at all. Here the adjusted
reduced-form coefficient is about 0.032 with a standard error of 0.016 — a
$t$ of roughly 2. That is the entire evidentiary basis for the IV estimate, and
it is thin. If the instrument has no detectable effect on the outcome, the IV
estimate is a small number divided by another small number, and its apparent
precision is an artefact.
:::

## Anderson--Rubin

The lecture's four steps: guess $\beta_0$, regress $Y - \beta_0 D$ on the
instrument and the controls, test the instrument's coefficient, and keep every
guess that is not rejected. `ivDiag::AR_test()` runs this test over a grid of
guesses. It has no standard-error option and prints no standard error: each test
on the grid uses heteroskedasticity-robust standard errors internally (clustered
ones if you add `cl = "your_cluster_variable"`), and the output is the test of
$\beta = 0$ and the interval.

```{r}
#| label: ar
## controls must be numeric columns, built in the packages chunk
ar <- AR_test(data = card, Y = "lwage", D = "education", Z = "near4",
              controls = c("age", "age2", "black", "smsa_yes", "south_yes"))
ar$Fstat        ## the AR test of beta = 0
ar$ci           ## the 95% AR confidence set

## the usual 95% CI, estimate +- 1.96 robust SE, for comparison
b_iv  <- coef(m_iv)["education"]
se_iv <- sqrt(diag(vcovHC(m_iv, type = "HC1")))["education"]
round(c(estimate = unname(b_iv), lower = unname(b_iv - 1.96 * se_iv),
        upper = unname(b_iv + 1.96 * se_iv)), 4)
```

How to read it:

-   **The test of $\beta = 0$.** At $\beta_0 = 0$, $Y - \beta_0 D$ is just $Y$,
    so the AR test of "no effect" is the reduced-form test: $F = 4.14$ is
    essentially the squared robust $t$ of the reduced form above ($2.03^2$).
    It needs no assumption about instrument strength.
-   **The interval.** The 95% AR set is $[0.004, 0.256]$: the returns to a year
    of schooling, from 0.4% to 25.6%, that the data cannot reject at the 5%
    level. With a first-stage $F$ of about 10, the instrument is weak by modern
    standards, so this is the interval to trust and report.
-   **Compared with the usual 95% CI.** The usual interval, $[-0.003, 0.190]$, is
    symmetric around the estimate and just includes zero; at this $F$ its 95%
    is not guaranteed. The AR set excludes zero and stretches further on the
    high side.

`ivDiag::ivDiag()` reports OLS, 2SLS, several first-stage $F$ statistics, and the
AR test in one call; it bootstraps by default, so it takes a minute.

# Weak instruments, by simulation

The cleanest way to see the danger is to build a world where you know the answer.
We use the lecture's design: $n = 500$, true effect 1, and strong endogeneity
(the errors in $D$ and $Y$ correlate at 0.9), so OLS converges to 1.9.

```{r}
#| label: weak-sim
set.seed(723)

one_draw <- function(pi, n = 500, beta = 1) {
  z <- rnorm(n); v <- rnorm(n)
  e <- 0.9 * v + sqrt(1 - 0.81) * rnorm(n)   ## error in Y, correlated with v
  d <- pi * z + v                            ## pi = instrument strength
  y <- beta * d + e                          ## TRUE effect is 1
  m <- ivreg(y ~ d | z)
  ## what the printed output shows: estimate, SE, first-stage F
  ## plus whether the AR test (at the truth) rejects the truth
  c(est = unname(coef(m)["d"]), se = sqrt(vcov(m)["d", "d"]),
    F   = unname(summary(lm(d ~ z))$fstatistic[1]),
    ar_rejects_truth = summary(lm(I(y - beta * d) ~ z))$coefficients["z", "t value"]^2 >
      qchisq(0.95, 1),
    ols = unname(coef(lm(y ~ d))["d"]))
}

strengths <- c(strong = 0.45, moderate = 0.11, weak = 0.06)
res <- lapply(strengths, function(p) as.data.frame(t(replicate(1000, one_draw(p)))))

## the sampling distribution, summarized
sapply(res, function(r) c(
  median_F  = median(r$F),
  median_IV = median(r$est),
  IQR_IV    = IQR(r$est),
  median_SE = median(r$se),
  mean_OLS  = mean(r$ols)
)) |> round(3)
```

```{r}
#| label: fig-weak
#| fig-cap: "Sampling distribution of the 2SLS estimate at three instrument strengths, truncated to [-3, 5] for display. The truth is 1 (dashed), the OLS limit 1.9 (dotted)."
## stack the three simulations into one data frame for ggplot
bind_rows(lapply(names(res), function(nm)
  tibble::tibble(est = res[[nm]]$est, strength = nm))) |>
  filter(est > -3, est < 5) |>
  ggplot(aes(est)) +
  geom_histogram(bins = 60, fill = "navy", alpha = .7) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_vline(xintercept = 1.9, linetype = 3) +
  facet_wrap(~ factor(strength, levels = c("strong", "moderate", "weak")),
             scales = "free_y") +
  labs(x = expression(hat(beta)[2*SLS]), y = "count") +
  theme_minimal(base_size = 9)
```

The strong instrument gives a tight, normal distribution centred on 1. As the
instrument weakens, the distribution spreads out, grows heavy tails, and leans
toward the OLS limit.

## What the printed output does

A researcher never sees the histogram. They see one draw: an estimate, a
standard error, and the interval $\hat\beta \pm 1.96\,\widehat{SE}$. How often
is that interval right?

```{r}
#| label: read-output
read_out <- function(r) {
  covers <- abs(r$est - 1) <= 1.96 * r$se   ## interval contains the truth
  signif <- abs(r$est / r$se) > 1.96        ## "significant" at 5%
  c(CI_covers_truth         = mean(covers),
    share_significant       = mean(signif),
    CI_covers_when_signif   = mean(covers[signif]),
    median_IV_when_signif   = median(r$est[signif]),
    AR_covers_truth         = mean(!r$ar_rejects_truth))
}
sapply(res, read_out) |> round(2)
```

Read the weak column:

-   **Most weak draws look honest.** The typical draw has a large standard
    error, so its interval is wide and covers 1, and the estimate is not
    significant (58% of weak draws). That is why overall coverage of the usual CI is still
    87%: a weak instrument usually prints a noisy, insignificant estimate.
-   **The draws that look best are the worst.** Among the significant results,
    the ones a paper would report, the interval covers the truth only 69% of
    the time (84% at $F \approx 6$), and their median estimate is 1.41, not 1.
    The significant draws are the ones that drifted toward OLS, and the printed
    output cannot tell you which draw you have.
-   **AR covers 95% in every column**, because it never divides by the first
    stage. This is why the diagnostics section tests with AR rather than reading
    the $t$-statistic.

# Who are the compliers?

The LATE applies to people whose schooling was changed by college proximity. We
cannot name them, but we can describe them.

```{r}
#| label: complier-share
## complier share = first stage on a binary treatment
card2 <- card |> mutate(some_college = as.integer(education >= 13))
fs_bin <- lm(some_college ~ near4 + age + I(age^2) + ethnicity + smsa + south,
             data = card2)
round(coef(summary(fs_bin))["near4", ], 4)
```

About 6 percent of the sample are compliers on this definition — that is, roughly
6 percent attend some college *because* they grew up near one.

```{r}
#| label: complier-profile
## Abadie's kappa weights let us describe complier characteristics;
## pz, the probability of growing up near a college given the controls,
## makes the weights valid when near4 is random only given the controls
pz <- fitted(glm(near4 ~ age + I(age^2) + ethnicity + smsa + south,
                 data = card2, family = binomial))
kappa <- 1 - card2$some_college * (1 - card2$near4) / (1 - pz) -
  (1 - card2$some_college) * card2$near4 / pz

## kappa-weighted mean of x = the compliers' mean of x
profile <- function(x) c(all = mean(x), compliers = sum(kappa * x) / sum(kappa))
round(rbind(
  father_educ = profile(card2$feducation |> tidyr::replace_na(mean(card2$feducation, na.rm = TRUE))),
  south = profile(card2$south_yes),
  smsa  = profile(card2$smsa_yes),
  black = profile(card2$black)
), 3)
```

::: callout-important
The complier profile is the interpretive heart of an IV paper. If the compliers
are disproportionately from low-education families, then the LATE is a return to
college **for people at the margin of attending** — which is often the
policy-relevant group, but is emphatically not the ATE.
:::

## The same profile with `ivdesc`

`ivdesc()` (Marbach and Hangartner 2020) computes the mean of a covariate for
compliers (`co`), never-takers (`nt`), and always-takers (`at`), with standard
errors, from a binary treatment and a binary instrument.

```{r}
#| label: ivdesc
## X = covariate to profile, D = binary treatment, Z = binary instrument
with(card2, ivdesc(X = black, D = some_college, Z = near4, boot = FALSE))
```

`ivdesc()` assumes the instrument is randomly assigned **without** conditioning
on covariates. Its balance test at the bottom asks whether the covariate differs
by $Z$, and here it rejects: Black respondents are less likely to have grown up
near a college. So this table is the profile for a design in which `near4` was a
lottery, and the covariate-adjusted $\kappa$ above is the right one for Card's
design. The same goes for its complier share, 12 percent here against 6 percent
with the controls. Use `ivdesc()` when $Z$ is randomized (a lottery, an experiment's
encouragement), and the hand-built $\kappa$ when it is random only given
controls.

# Exercises

1.  **Reduced form first.** Suppose a referee reports only the 2SLS estimate and
    its $t$-statistic. Compute the reduced form and first stage and write two
    sentences a referee could use to evaluate whether the design is credible.

2.  **A placebo.** Growing up near a college should not affect variables
    determined before college attendance. Regress `feducation`, `meducation`, and
    `library14` on `near4` with the same controls. Do any of them move? What
    would it mean for exclusion if they did?

3.  **Anderson–Rubin by hand.** Write a function that, for a grid of $\beta_0$
    values, regresses $Y - \beta_0 D$ on the instrument and controls and records
    the robust $p$-value on the instrument. Plot it, read off the 95% AR
    confidence set, and check it against `ar$ci`.

4.  **Bias amplification with weak instruments.** In `one_draw()`, add a small
    direct effect of $Z$ on $Y$ (violating exclusion by, say, 0.05). Show that
    the resulting bias in 2SLS grows as the instrument weakens, and compare it
    with the OLS bias.

# Session information

```{r}
#| label: session
## R version and the packages this lab loaded
c(R = R.version.string, sapply(sessionInfo()$otherPkgs, \(p) as.character(p$Version)))
```
