---
title: "\\textbf{Lab 13: Double Machine Learning and Causal Forests}"
subtitle: "SOCIOL 723 --- Social Statistics II --- Thursday, November 19"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.4, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

1.  **Watch naive ML fail** on a problem where we know the answer.
2.  Fix it with orthogonalization, and fix it again with cross-fitting.
3.  Run DML with `DoubleML`, and check that the answer does not depend on the
    split.
4.  Estimate heterogeneous effects with a causal forest, and test whether the
    heterogeneity is real.

```{r}
#| label: packages
library(dplyr)
library(ggplot2)
library(glmnet)
library(ranger)
library(DoubleML); library(mlr3); library(mlr3learners)
library(grf)
lgr::get_logger("mlr3")$set_threshold("warn")
set.seed(723)
```

# Regularization bias, in a world we control

The partially linear model $Y = \theta D + g(W) + \epsilon$,
$D = m(W) + v$, with $\theta = 0.5$ known.

```{r}
#| label: dgp
## approximately sparse coefficients: a few large, a long decaying tail
make_data <- function(n = 500, p = 200, theta = 0.5) {
  W <- matrix(rnorm(n * p), n, p); colnames(W) <- paste0("w", seq_len(p))
  b <- 1 / (seq_len(p))^2
  m <- as.vector(W %*% b)          # E[D | W]
  g <- as.vector(W %*% b)          # E[Y | W] beyond theta*D
  D <- m + rnorm(n)
  Y <- theta * D + g + rnorm(n)
  list(Y = Y, D = D, W = W)
}
```

Note the structure: the *same* covariates drive both $D$ and $Y$, so $W$ is
genuinely confounding. With $p = 200$ and $n = 500$ OLS is estimable, but it is
so high-variance and poorly conditioned as to be useless --- which is why we
regularize.

```{r}
#| label: three-estimators
estimators <- function(dat) {
  Y <- dat$Y; D <- dat$D; W <- dat$W; n <- length(Y)

  ## 1. NAIVE: lasso Y on (D, W), leaving D unpenalized, and read off D
  pf    <- c(0, rep(1, ncol(W)))
  f_all <- cv.glmnet(cbind(D, W), Y, penalty.factor = pf)
  naive <- as.vector(coef(f_all, s = "lambda.min"))[2]

  ## 2. ORTHOGONAL, no cross-fitting: residualize both on W, same data
  ly  <- cv.glmnet(W, Y);  lm_ <- cv.glmnet(W, D)
  ry0 <- Y - as.vector(predict(ly,  W, s = "lambda.min"))
  rd0 <- D - as.vector(predict(lm_, W, s = "lambda.min"))
  ortho <- unname(coef(lm(ry0 ~ rd0))[2])

  ## 3. ORTHOGONAL + CROSS-FITTING
  K <- 5; folds <- sample(rep(1:K, length.out = n))
  ry <- rd <- numeric(n)
  for (k in 1:K) {
    tr <- folds != k; te <- folds == k
    a <- cv.glmnet(W[tr, ], Y[tr]); b <- cv.glmnet(W[tr, ], D[tr])
    ry[te] <- Y[te] - as.vector(predict(a, W[te, ], s = "lambda.min"))
    rd[te] <- D[te] - as.vector(predict(b, W[te, ], s = "lambda.min"))
  }
  dml <- unname(coef(lm(ry ~ rd))[2])

  c(naive = naive, orthogonal = ortho, dml = dml)
}

round(estimators(make_data()), 4)
```

```{r}
#| label: repeat-sim
reps <- t(replicate(30, estimators(make_data())))
round(rbind(mean = colMeans(reps), bias = colMeans(reps) - 0.5,
            sd = apply(reps, 2, sd),
            rmse = sqrt(colMeans((reps - 0.5)^2))), 4)
```

```{r}
#| label: fig-dml
#| fig-cap: "Sampling distributions over 30 simulations. The truth is 0.5 (dashed). The naive lasso is biased upward; residualizing both sides removes almost all of it."
as.data.frame(reps) |>
  tidyr::pivot_longer(everything()) |>
  mutate(name = factor(name, levels = c("naive", "orthogonal", "dml"),
                       labels = c("naive", "orthogonal", "orthogonal + cross-fit"))) |>
  ggplot(aes(value, fill = name)) +
  geom_density(alpha = .5, colour = NA) +
  geom_vline(xintercept = 0.5, linetype = 2) +
  scale_fill_manual(values = c("firebrick", "goldenrod", "navy")) +
  labs(x = expression(hat(theta)), y = NULL, fill = NULL) +
  theme_minimal(base_size = 10)
```

::: callout-important
Read the bias column. The naive estimator's error does not come from a bad
learner — the lasso predicts $Y$ well. It comes from the fact that *any*
regularized estimate is shrunk, and that shrinkage transmits directly to
$\hat\theta$ unless the moment is orthogonal.

Note also that here **orthogonalization does nearly all the work** and
cross-fitting adds little: a cross-validated lasso does not overfit badly at this
$n$. Cross-fitting matters much more with flexible learners such as forests and
boosting, which fit the training sample far more aggressively — Exercise 2 makes
that comparison.
:::

# DML with the `DoubleML` package

The 401(k) data: does *eligibility* for a 401(k) plan raise net financial
assets? Eligibility is not random — it is offered by employers — so we condition
on income, age, family size, education, and other plan participation.

```{r}
#| label: dml-data
## cached locally so the handout renders without a network connection
d401 <- readRDS("../Data/pension401k.rds")
dim(d401)

dml_data <- DoubleMLData$new(d401, y_col = "net_tfa", d_cols = "e401",
                             x_cols = c("age", "inc", "educ", "fsize", "marr",
                                        "twoearn", "db", "pira", "hown"))
dml_data
```

```{r}
#| label: dml-plr
lrn_g <- lrn("regr.ranger", num.trees = 400, min.node.size = 10)
lrn_m <- lrn("regr.ranger", num.trees = 400, min.node.size = 10)

set.seed(1)
plr <- DoubleMLPLR$new(dml_data, ml_l = lrn_g, ml_m = lrn_m,
                       n_folds = 5, n_rep = 1)
plr$fit()
plr$summary()
```

```{r}
#| label: dml-compare
## a naive OLS benchmark
ols <- lm(net_tfa ~ e401 + age + inc + educ + fsize + marr + twoearn + db +
            pira + hown, data = d401)
c(OLS = coef(ols)["e401"], DML_PLR = plr$coef)
```

## The interactive model (AIPW with ML)

The partially linear model assumes a constant effect. The *interactive*
regression model does not — it is the AIPW score from Week 6 with cross-fitted
ML nuisances.

```{r}
#| label: dml-irm
lrn_g2 <- lrn("regr.ranger", num.trees = 400, min.node.size = 10)
lrn_m2 <- lrn("classif.ranger", num.trees = 400, min.node.size = 10,
              predict_type = "prob")

set.seed(1)
irm <- DoubleMLIRM$new(dml_data, ml_g = lrn_g2, ml_m = lrn_m2,
                       n_folds = 5, score = "ATE")
irm$fit()
irm$summary()
```

## Does the answer depend on the split?

```{r}
#| label: dml-reps
set.seed(2)
plr_rep <- DoubleMLPLR$new(dml_data, ml_l = lrn_g, ml_m = lrn_m,
                           n_folds = 5, n_rep = 10)
plr_rep$fit()
c(median_estimate = plr_rep$coef, se = plr_rep$se)
```

`n_rep = 10` repeats the whole cross-fitting procedure with different partitions
and reports the median. **If the estimate moves a lot across partitions, report
that** — it means the nuisance estimates are unstable relative to the signal.

# Heterogeneous treatment effects

```{r}
#| label: cf
X <- as.matrix(d401[, c("age", "inc", "educ", "fsize", "marr", "twoearn",
                        "db", "pira", "hown")])
Y <- d401$net_tfa
D <- d401$e401

cf <- causal_forest(X, Y, D, num.trees = 2000, seed = 1)
average_treatment_effect(cf, target.sample = "all")
```

```{r}
#| label: calibration
test_calibration(cf)
```

::: callout-important
Read `test_calibration()` carefully. `mean.forest.prediction` near 1 says the
average effect is well estimated. **`differential.forest.prediction` is the
one that matters**: a coefficient near 1 with a small $p$-value is evidence of
*real* heterogeneity; a coefficient near 0 says your CATE surface is noise, and
you should not interpret it.
:::

```{r}
#| label: fig-cate
#| fig-cap: "Estimated conditional average treatment effects against income."
tau <- predict(cf)$predictions
tibble::tibble(tau, inc = d401$inc) |>
  filter(inc < quantile(inc, .98)) |>
  ggplot(aes(inc, tau)) +
  geom_point(alpha = .1, size = .4, colour = "navy") +
  geom_smooth(method = "loess", se = TRUE, colour = "firebrick") +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "income", y = expression(hat(tau)(x))) +
  theme_minimal(base_size = 10)
```

## Sorted group ATEs, honestly

Rank units by $\hat\tau$, form quintiles, and estimate a **doubly robust** ATE
within each — using the forest's own out-of-bag predictions so the ranking and
the estimate do not use the same information.

```{r}
#| label: gates
dr <- get_scores(cf)                       # doubly robust scores, one per unit
grp <- cut(tau, breaks = quantile(tau, seq(0, 1, .2)), include.lowest = TRUE,
           labels = paste0("Q", 1:5))

gates <- tibble::tibble(dr, grp) |>
  group_by(grp) |>
  summarise(ate = mean(dr), se = sd(dr) / sqrt(n()), n = n(), .groups = "drop") |>
  mutate(lo = ate - 1.96 * se, hi = ate + 1.96 * se)
knitr::kable(gates, digits = 0)
```

```{r}
#| label: fig-gates
#| fig-cap: "Doubly robust group average treatment effects by predicted-effect quintile."
ggplot(gates, aes(grp, ate)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "navy") +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "quintile of predicted effect", y = "doubly robust ATE") +
  theme_minimal(base_size = 10)
```

```{r}
#| label: gate-test
## is the top quintile different from the bottom?
q1 <- dr[grp == "Q1"]; q5 <- dr[grp == "Q5"]
d_hat <- mean(q5) - mean(q1)
se_d  <- sqrt(var(q5)/length(q5) + var(q1)/length(q1))
c(difference = d_hat, se = se_d, t = d_hat / se_d)
```

```{r}
#| label: varimp
vi <- variable_importance(cf)
tibble::tibble(variable = colnames(X), importance = as.vector(vi)) |>
  arrange(desc(importance)) |> head(6) |> knitr::kable(digits = 3)
```

This is importance for **effect heterogeneity**, not for predicting the outcome —
a different and more useful quantity. It is still descriptive: it says where the
forest found variation, not why.

# Exercises

1.  **How slow can the learner be?** In `estimators()`, replace `cv.glmnet` with
    a deliberately crude learner --- a single shallow tree via
    `ranger(x = as.data.frame(W), y = Y, num.trees = 1, max.depth = 2)`. Does DML
    still recover 0.5? At what point does the rate condition plausibly fail?

2.  **Where cross-fitting earns its keep.** Repeat the three-estimator comparison
    with a random forest in place of the lasso (`ranger` on
    `as.data.frame(W)`, remembering that `ranger` needs column names). Forests fit
    their training sample far more aggressively than a cross-validated lasso, so
    the gap between the "orthogonal" and "orthogonal + cross-fit" columns should
    now be visible. Report both.

3.  **Overlap.** In the 401(k) IRM, extract the estimated propensity scores and
    plot them by treatment status. Are there units with $\hat e$ near 0 or 1?
    Re-run trimming to $[0.05, 0.95]$ and report how the ATE moves.

4.  **Learner choice.** Re-run the PLR with `regr.cv_glmnet` and with
    `regr.ranger`. How much does $\hat\theta$ move relative to its standard
    error? What does stability across learners buy you, and what does it not?

5.  **DML for IV.** `fetch_401k(instrument = TRUE)` gives eligibility (`e401`) as
    an instrument for participation (`p401`). Fit `DoubleMLPLIV` and compare with
    the PLR estimate. Which estimand is each targeting?

6.  **Honest heterogeneity.** The group ATEs above already use out-of-bag
    predictions and doubly robust scores, which removes most --- but not all ---
    of the optimism, because the same units informed the forest's structure. Split
    the sample in half: fit the causal forest on the first half, predict
    $\hat\tau$ on the second, and estimate group ATEs in the second half only.
    How much does the top-minus-bottom contrast shrink?

# Session information

```{r}
#| label: session
sessionInfo()
```
