---
title: "\\textbf{Lab 1: OLS by Hand, in Scalar Form}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, August 27"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator*{\argmin}{arg\,min}
        \DeclareMathOperator{\trace}{tr}
        \DeclareMathOperator{\rank}{rank}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\En}{\mathbb{E}_n}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.6, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

By the end of this lab you should be able to:

1.  Organize an analysis in RStudio and Quarto so that it is reproducible.
2.  Compute $\hat\beta_0$ and $\hat\beta_1$ from the scalar formulas with nothing but `sum()`, and verify that they match `lm()`.
3.  Verify the mechanical residual properties, and compute $s^2$ and the classical standard error by hand.
4.  Use **simulation** to see what the assumption $\E[\epsilon_i \mid X_i] = 0$ does and does not deliver.
5.  Verify the Frisch--Waugh--Lovell theorem and the omitted-variable-bias formula in real GSS data.

Everything below is scalar arithmetic: sums, means, and `lm()`. No matrix computations appear anywhere in this lab.

# Getting set up

## Packages

```{r}
#| label: packages
library(dplyr)      # data manipulation
library(ggplot2)    # graphics
library(broom)      # tidy() / augment() for model objects
library(modelsummary) # regression tables
```

If any of these are missing, install them once with `install.packages(c("dplyr", "ggplot2", "broom", "modelsummary"))`.

## Working reproducibly

Three habits, which I will assume for every problem set:

-   **Use an RStudio Project.** `File > New Project`. Never use `setwd()`; paths should be relative to the project root.
-   **Write in Quarto (`.qmd`) or RMarkdown (`.Rmd`).** Your submission should be a rendered PDF produced from a source file that I could run myself.
-   **Set a seed** before anything random: `set.seed(1)`.

A useful discipline: restart `R` (`Session > Restart R`, or `Cmd/Ctrl+Shift+F10`) and re-render from a clean session before you submit. If it does not run from scratch, it is not reproducible.

# The data

We use an extract from the **General Social Survey** (NORC) cumulative file, pooling the 2010, 2012, 2014, 2016, 2018 and 2022 waves. The sample is restricted to respondents working full time, aged 25--64, with positive personal income and non-missing values on the analysis variables.

```{r}
#| label: load-data
gss <- readRDS("../Data/gss_earnings.rds")
dim(gss)
str(gss, give.attr = FALSE)
```

| Variable | Meaning |
|:----------------|:------------------------------------------------------|
| `lnearn` | log of personal income, constant dollars (`realrinc`) |
| `educ` | years of schooling completed |
| `exper` | potential experience, $\max(\text{age} - \text{educ} - 6,\, 0)$ |
| `female`, `black`, `otherace` | indicator variables |
| `wordsum` | score on a 10-word vocabulary test (proxy for verbal ability) |
| `pareduc` | highest education of either parent, in years |
| `prestige`, `sei` | occupational prestige and socioeconomic index |
| `hours` | usual hours worked last week |
| `region` | Census division (9 categories) |
| `stratum`, `psu` | GSS sampling stratum and primary sampling unit (used in Lab 2) |
| `wt` | post-stratification sampling weight |

::: callout-note
The script that builds this file from the raw cumulative file is in `Data/build_gss_extract.R`. You do not need to run it.
:::

A first look:

```{r}
#| label: describe
gss |>
  summarise(across(c(lnearn, educ, exper, wordsum, pareduc, hours),
                   list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) |>
  t() |> round(2)
```

```{r}
#| label: fig-cef
#| fig-cap: "The CEF of log earnings given years of schooling, GSS 2010--2022."
gss |>
  group_by(educ) |>
  summarise(m = mean(lnearn), n = n()) |>
  filter(n >= 20) |>
  ggplot(aes(educ, m)) +
  geom_point(aes(size = n), colour = "navy", alpha = .7) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = .6) +
  labs(x = "Years of schooling", y = "Mean log earnings", size = "n") +
  theme_minimal(base_size = 10)
```

The points are the **conditional expectation function** estimated cell by cell (this is what a *saturated* regression would fit). The line is the **best linear approximation** to it. Notice that the CEF is not exactly linear; it is convex above roughly 12 years, but the line summarizes it reasonably well.

# OLS by hand, in scalar form

## The slope and intercept from sums

The whole estimator is two lines of arithmetic: $\hat\beta_1 = \Cov_n(X,Y)/\Var_n(X)$ and $\hat\beta_0 = \bar Y - \hat\beta_1\bar X$.

```{r}
#| label: ols-scalar
x <- gss$educ
y <- gss$lnearn
n <- length(y)

b1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
b0 <- mean(y) - b1 * mean(x)
round(c(b0 = b0, b1 = b1), 4)
```

Compare with `lm()`:

```{r}
#| label: ols-scalar-check
fit_biv <- lm(lnearn ~ educ, data = gss)
round(coef(fit_biv), 4)
```

Identical, and identical to the numbers on the lecture slides. `lm()` is doing exactly this arithmetic (in a numerically fancier way), nothing more.

## The slope is a weighted average of unit slopes

Lecture claimed $\hat\beta_1 = \sum_i w_i\,(Y_i - \bar Y)/(X_i - \bar X)$ with $w_i = (X_i - \bar X)^2 / \sum_j (X_j - \bar X)^2$. Verify it, and look at who carries the weight:

```{r}
#| label: weighted-slopes
w  <- (x - mean(x))^2 / sum((x - mean(x))^2)
bi <- ifelse(x == mean(x), 0, (y - mean(y)) / (x - mean(x)))
c(weighted_average = sum(w * bi), b1 = b1)

## total weight carried by each value of educ
tapply(w, x, sum) |> round(3)
```

Respondents far from the mean of schooling (about `r round(mean(gss$educ), 1)` years) carry almost all of the weight; respondents at the mean carry essentially none. Know which observations are shouting.

## Residual mechanics

```{r}
#| label: orthogonality
e <- y - b0 - b1 * x
round(c(sum_e = sum(e), sum_xe = sum(x * e), cor_xe = cor(x, e)), 8)
```

These hold to machine precision in *every* sample, no matter how wrong the model is. They are consequences of the normal equations, not evidence of anything.

## $s^2$ and the classical standard error

```{r}
#| label: ols-se
s2    <- sum(e^2) / (n - 2)
se_b1 <- sqrt(s2 / sum((x - mean(x))^2))
round(c(s2 = s2, se_b1 = se_b1), 5)
round(summary(fit_biv)$coefficients["educ", ], 5)
```

The standard error `R` prints is exactly $\sqrt{s^2 / \sum_i (X_i - \bar X)^2}$, i.e. it *assumes homoskedasticity*. Next week we replace it.

# Simulation: what the error assumption does

With real data we never know the truth. In a simulation we do, which makes it the honest way to see what an assumption buys.

## When the average of $Y$ really is a line

```{r}
#| label: sim-linear
set.seed(1)
n_sim <- 2000
xs <- sample(8:20, n_sim, replace = TRUE)          # schooling-like regressor
ys <- 8.2 + 0.12 * xs + rnorm(n_sim, 0, 0.7)       # E[eps | x] = 0 by design
coef(lm(ys ~ xs)) |> round(4)
```

We wrote the truth ($\beta_0 = 8.2$, $\beta_1 = 0.12$) and OLS finds it, up to sampling noise. Rerun with a different seed: the estimate moves a little. How much it moves from sample to sample is precisely next week's topic.

## When the CEF is not a line: uncorrelated is not mean-independent

Now make the population relationship curved, and fit a line anyway:

```{r}
#| label: sim-quadratic
#| fig-cap: "Residuals from fitting a line to a curved CEF. Their correlation with x is exactly zero, yet their average tracks the leftover curve: uncorrelated, but not mean-independent."
set.seed(2)
xq <- runif(n_sim, 8, 20)
yq <- 8 + 0.05 * xq + 0.015 * (xq - 14)^2 + rnorm(n_sim, 0, 0.4)
fq <- lm(yq ~ xq)

round(cor(xq, resid(fq)), 8)     # exactly zero, mechanically

plot(xq, resid(fq), pch = 16, cex = .35, col = "navy",
     xlab = "x", ylab = "residual")
lines(lowess(xq, resid(fq)), col = "firebrick", lwd = 2)
abline(h = 0, lty = 2)
```

This is the ladder from lecture, live: the residuals are *uncorrelated* with $x$ (bottom rung, guaranteed by the algebra) but their *average* clearly moves with $x$, so mean independence fails. The correlation could never have told you; the plot does. This is why you look at residual plots.

## Heteroskedasticity does not bias the slope

```{r}
#| label: sim-hetero
set.seed(3)
slopes <- replicate(500, {
  xh <- runif(500, 8, 20)
  yh <- 8.2 + 0.12 * xh + rnorm(500, 0, 0.1 * xh)   # spread grows with x
  coef(lm(yh ~ xh))[2]
})
round(c(mean_slope = mean(slopes), truth = 0.12), 4)
```

The error spread rises steeply with $x$, yet the slope estimates still centre on the truth: heteroskedasticity is a problem for *standard errors* (Week 2), not for the estimate itself. $\E[\epsilon\mid X] = 0$ still holds here, and that is the assumption doing the work.

# The multiple regression, and fit

With several regressors the same logic goes through; we let `lm()` carry the bookkeeping. (Thursday's session shows the notation that makes this compact.)

```{r}
#| label: fit-multiple
fit <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)
y   <- gss$lnearn
round(coef(fit), 4)
```

## The sum-of-squares decomposition

```{r}
#| label: ssq
yhat <- fitted(fit)
TSS <- sum((y - mean(y))^2)
ESS <- sum((yhat - mean(y))^2)
RSS <- sum(residuals(fit)^2)

c(TSS = TSS, ESS_plus_RSS = ESS + RSS)
c(R2_manual = ESS / TSS, R2_lm = summary(fit)$r.squared)
```

And the model-comparison reading of $R^2$: it is $PRE$ against the intercept-only model.

```{r}
#| label: pre
fit_c <- lm(lnearn ~ 1, data = gss)              # MODEL C
SSE_C <- sum(residuals(fit_c)^2)
SSE_A <- sum(residuals(fit)^2)                    # MODEL A
c(PRE = (SSE_C - SSE_A) / SSE_C, R2 = summary(fit)$r.squared)
```

# Frisch--Waugh--Lovell

We now verify FWL on the return to schooling: the treatment of interest is `educ`, and the controls are everything else in the regression.

```{r}
#| label: fwl
## Step 1: residualize the treatment on the controls
d_tilde <- residuals(lm(educ ~ exper + I(exper^2) + female + black, data = gss))

## Step 2: residualize the outcome on the controls
y_tilde <- residuals(lm(lnearn ~ exper + I(exper^2) + female + black, data = gss))

## Step 3: bivariate regression of the residuals
b_fwl <- coef(lm(y_tilde ~ d_tilde))["d_tilde"]

c(full_regression = coef(fit)["educ"], fwl = b_fwl)
```

Both forms of the regression-anatomy formula:

```{r}
#| label: anatomy
c(cov_ytilde_dtilde = cov(y_tilde, d_tilde) / var(d_tilde),
  cov_y_dtilde      = cov(gss$lnearn, d_tilde) / var(d_tilde))
```

In fact *all three* numerators agree, for the same reason:

```{r}
#| label: anatomy-three
c(both_residualized = cov(y_tilde, d_tilde) / var(d_tilde),
  only_regressor    = cov(gss$lnearn, d_tilde) / var(d_tilde),
  only_outcome      = cov(y_tilde, gss$educ) / var(d_tilde))
```

What you *cannot* change is the **denominator**:

```{r}
#| label: anatomy-wrong
c(correct = coef(lm(y_tilde ~ d_tilde))["d_tilde"],
  wrong   = coef(lm(y_tilde ~ gss$educ))[2])
```

::: callout-note
**Why does the second one fail?** Regressing $\tilde Y$ on raw `educ` divides by $\Var_n(\texttt{educ})$ instead of $\Var_n(\widetilde{\texttt{educ}})$. Since residualizing can only shrink the variance, the denominator is too large and the estimate is attenuated. The numerators are all equal because the part removed from $Y$ lies in the span of the controls and is orthogonal to $\tilde D$. Work this out on paper; it is Question 2 of Problem Set 1.
:::

The residuals are also identical, as the theorem promises:

```{r}
#| label: fwl-resid
all.equal(unname(residuals(lm(y_tilde ~ d_tilde))), unname(residuals(fit)))
```

## The added-variable plot

```{r}
#| label: fig-avp
#| fig-cap: "Added-variable plot for years of schooling. The slope of the fitted line *is* the multivariate coefficient on `educ`."
tibble::tibble(d_tilde, y_tilde) |>
  ggplot(aes(d_tilde, y_tilde)) +
  geom_point(alpha = .12, size = .5, colour = "navy") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = .7) +
  labs(x = "Years of schooling | controls", y = "Log earnings | controls") +
  theme_minimal(base_size = 10)
```

Note how much of the spread in `educ` survives residualization (`sd(d_tilde)` $=$ `r round(sd(d_tilde), 2)` versus `sd(educ)` $=$ `r round(sd(gss$educ), 2)`). When that ratio gets small, the coefficient is being estimated off a thin slice of the data, and its standard error will show it.

# Omitted variable bias, numerically

Short regression omits verbal ability and parental education; long regression includes them.

```{r}
#| label: ovb
long  <- lm(lnearn ~ educ + exper + I(exper^2) + female + black +
              wordsum + pareduc, data = gss)
short <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)

## auxiliary regressions: each omitted variable on the included ones
aux_w <- lm(wordsum ~ educ + exper + I(exper^2) + female + black, data = gss)
aux_p <- lm(pareduc ~ educ + exper + I(exper^2) + female + black, data = gss)

beta_short <- coef(short)["educ"]
beta_long  <- coef(long)["educ"]
bias_pred  <- coef(long)["wordsum"] * coef(aux_w)["educ"] +
              coef(long)["pareduc"] * coef(aux_p)["educ"]

round(c(short = beta_short,
        long = beta_long,
        long_plus_bias = beta_long + bias_pred,
        bias = bias_pred), 5)
```

The identity $\tilde\beta_1 = \beta_1 + \sum_j \beta_{2j}\delta_{1j}$ holds exactly, to machine precision. Note the direction: ability and family background are positively associated with schooling and positively associated with earnings, so omitting them **overstates** the return to schooling.

Column (1) is the bivariate regression; (2) adds the demographic controls; (3) adds vocabulary score and parental education.

```{r}
#| label: ovb-table
modelsummary(list("(1)" = lm(lnearn ~ educ, data = gss),
                  "(2)" = short,
                  "(3)" = long),
             coef_map = c("educ" = "Years of schooling",
                          "wordsum" = "Vocabulary score",
                          "pareduc" = "Parental education"),
             gof_map = c("nobs", "r.squared"),
             stars = TRUE, output = "markdown")
```

::: callout-important
Everything in this section is *algebra*. It tells you how the estimate moves when you add a control. It does **not** tell you whether the long regression recovers the causal return to schooling, for that you would need to argue that nothing else is omitted, which is a substantive claim, not a statistical one. That argument is what Weeks 5--11 are about.
:::

# Exercises

Work these during lab; they are not collected, but Problem Set 1 builds directly on them.

1.  **The dummy variable trap.** Run `lm(lnearn ~ factor(region), data = gss)` and look at how many coefficients you get. Now try to include *all nine* region indicators and an intercept. What does `lm()` report, and why? Explain in words: the nine dummies add up to the intercept column, so one of them carries no new information.

2.  **The** $n-2$, by simulation. In the bivariate model, verify that $\E[\sum_i \hat e_i^2] = \sigma^2(n-2)$: fix `xs` from the first simulation, generate $y = 8.2 + 0.12x + \epsilon$ with $\epsilon \sim N(0, 4)$, and average $\sum_i \hat e_i^2$ over 2,000 replications. Compare to $4 \times (n-2)$, and to the wrong guess $4 \times n$.

3.  **Sign the bias.** Suppose you omit `hours` from the long regression. Predict the sign of the bias in the `educ` coefficient *before* running anything, using the OVB formula and the two auxiliary regressions. Then check.

4.  **The FWL standard error.** Compute $s^2 / \sum_i \tilde d_i^{\,2}$ using `d_tilde` from the FWL section (with $s^2$ from the full regression), take the square root, and confirm it matches the `educ` standard error printed by `summary(fit)`.

5.  **Saturate the key covariate.** Fit `lm(lnearn ~ factor(educ) + exper + I(exper^2) + female + black, data = gss)`, which gives every year of schooling its own coefficient, and compare it to the linear-in-`educ` version with `anova()`. The $F$-test you get is exactly the model comparison from *Social Statistics I*: is the reduction in $SSE$ worth the extra parameters?

# Session information

```{r}
#| label: session
sessionInfo()
```
