---
title: "\\textbf{Lab 1: Matrix Algebra and OLS by Hand}"
subtitle: "SOCIOL 723, Social Statistics II, Thursday, August 27"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    number-sections: true
    toc: true
    toc-depth: 2
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator*{\argmin}{arg\,min}
        \DeclareMathOperator{\trace}{tr}
        \DeclareMathOperator{\rank}{rank}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\En}{\mathbb{E}_n}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.6, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

# Goals

By the end of this lab you should be able to:

1.  Organize an analysis in RStudio and Quarto so that it is reproducible.
2.  Do the linear algebra of the linear model in `R`: transposes, products,
    inverses, rank, and traces.
3.  Compute $\hat{\bm{\beta}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{y}$
    by hand and verify that it matches `lm()`.
4.  Construct the projection matrix $\bm{P}$ and the residual maker
    $\bm{M}$, and verify their properties numerically.
5.  Verify the Frisch--Waugh--Lovell theorem and the omitted-variable-bias
    formula in real data.

Everything below runs on base `R` plus a handful of packages. Nothing here
requires more linear algebra than we covered in lecture.

# Getting set up

## Packages

```{r}
#| label: packages
library(dplyr)      # data manipulation
library(ggplot2)    # graphics
library(broom)      # tidy() / augment() for model objects
library(modelsummary) # regression tables
```

If any of these are missing, install them once with
`install.packages(c("dplyr", "ggplot2", "broom", "modelsummary"))`.

## Working reproducibly

Three habits, which I will assume for every problem set:

-   **Use an RStudio Project.** `File > New Project`. Never use `setwd()`;
    paths should be relative to the project root.
-   **Write in Quarto (`.qmd`) or RMarkdown (`.Rmd`).** Your submission should be
    a rendered PDF produced from a source file that I could run myself.
-   **Set a seed** before anything random: `set.seed(1)`.

A useful discipline: restart `R` (`Session > Restart R`, or
`Cmd/Ctrl+Shift+F10`) and re-render from a clean session before you submit.
If it does not run from scratch, it is not reproducible.

# The data

We use an extract from the **General Social Survey** (NORC) cumulative file,
pooling the 2010, 2012, 2014, 2016, 2018 and 2022 waves. The sample is
restricted to respondents working full time, aged 25--64, with positive personal
income and non-missing values on the analysis variables.

```{r}
#| label: load-data
gss <- readRDS("../Data/gss_earnings.rds")
dim(gss)
str(gss, give.attr = FALSE)
```

| Variable | Meaning |
|:--------------|:--------------------------------------------------------|
| `lnearn` | log of personal income, constant dollars (`realrinc`) |
| `educ` | years of schooling completed |
| `exper` | potential experience, $\max(\text{age} - \text{educ} - 6,\, 0)$ |
| `female`, `black`, `otherace` | indicator variables |
| `wordsum` | score on a 10-word vocabulary test (proxy for verbal ability) |
| `pareduc` | highest education of either parent, in years |
| `prestige`, `sei` | occupational prestige and socioeconomic index |
| `hours` | usual hours worked last week |
| `region` | Census division (9 categories) |
| `stratum`, `psu` | GSS sampling stratum and primary sampling unit (used in Lab 2) |
| `wt` | post-stratification sampling weight |

::: callout-note
The script that builds this file from the raw cumulative file is in
`Data/build_gss_extract.R`. You do not need to run it.
:::

A first look:

```{r}
#| label: describe
gss |>
  summarise(across(c(lnearn, educ, exper, wordsum, pareduc, hours),
                   list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) |>
  t() |> round(2)
```

```{r}
#| label: fig-cef
#| fig-cap: "The CEF of log earnings given years of schooling, GSS 2010--2022."
gss |>
  group_by(educ) |>
  summarise(m = mean(lnearn), n = n()) |>
  filter(n >= 20) |>
  ggplot(aes(educ, m)) +
  geom_point(aes(size = n), colour = "navy", alpha = .7) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = .6) +
  labs(x = "Years of schooling", y = "Mean log earnings", size = "n") +
  theme_minimal(base_size = 10)
```

The points are the **conditional expectation function** estimated cell by cell
(this is what a *saturated* regression would fit). The line is the **best linear
approximation** to it. Notice that the CEF is not exactly linear; it is
convex above roughly 12 years, but the line summarizes it reasonably well.

# Matrix algebra in `R`

## Building blocks

```{r}
#| label: matrix-basics
A <- matrix(c(2, 1, 1, 3), nrow = 2)        # filled by column
b <- c(1, 2)

A
t(A)              # transpose
A %*% b           # matrix product (NOT A * b, which is elementwise)
solve(A)          # inverse
solve(A) %*% A    # should be the identity
diag(2)           # 2x2 identity
```

Two facts worth internalizing now, because they recur constantly:

```{r}
#| label: matrix-facts
crossprod(A)          # equals t(A) %*% A, but faster and more accurate
all.equal(crossprod(A), t(A) %*% A)

sum(diag(A))          # the trace
```

## Rank, invertibility, and collinearity

$(\bm{X}'\bm{X})^{-1}$ exists if and only if $\bm{X}$ has full
column rank. Here is what a rank failure looks like:

```{r}
#| label: rank
X_bad <- cbind(1, gss$educ, gss$educ * 2)   # third column is redundant
qr(X_bad)$rank                              # 2, not 3
try(solve(crossprod(X_bad)), silent = TRUE) |> class()
```

`lm()` does not error, it drops the offending column and reports `NA`:

```{r}
#| label: rank-lm
coef(lm(lnearn ~ educ + I(educ * 2), data = gss))
```

::: callout-warning
An `NA` coefficient is *never* a mystery: it means the column is an exact
linear combination of the others. The most common cause is the dummy variable
trap (including all $J$ category indicators plus an intercept).
:::

## Positive definiteness

$\bm{X}'\bm{X}$ is positive semi-definite by construction, and positive
definite when $\bm{X}$ has full column rank. We can check via eigenvalues:

```{r}
#| label: posdef
X <- model.matrix(~ educ + exper + I(exper^2) + female + black, data = gss)
ev <- eigen(crossprod(X), only.values = TRUE)$values
range(ev)
all(ev > 0)
```

The ratio of the largest to smallest eigenvalue is the **condition number**; a
very large value warns of near-collinearity (numerically fragile, and
substantively a warning that the data carry little independent information about
some coefficient).

```{r}
#| label: condnum
kappa(crossprod(X), exact = TRUE)
```

# OLS by hand

## The estimator

```{r}
#| label: ols-by-hand
y <- gss$lnearn
X <- model.matrix(~ educ + exper + I(exper^2) + female + black, data = gss)

n <- nrow(X); k <- ncol(X)
c(n = n, k = k)

XtX     <- crossprod(X)          # X'X   (k x k)
Xty     <- crossprod(X, y)       # X'y   (k x 1)
beta_hat <- solve(XtX, Xty)      # solve(A, b) is better than solve(A) %*% b
round(beta_hat, 4)
```

Compare with `lm()`:

```{r}
#| label: ols-check
fit <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)
all.equal(as.vector(beta_hat), unname(coef(fit)))
```

::: callout-tip
`solve(A, b)` solves the linear system directly instead of forming the inverse.
It is faster and numerically more stable. In production code you would go
further and use the QR decomposition, which is what `lm()` actually does:
`qr.solve(X, y)`.
:::

```{r}
#| label: qr
all.equal(as.vector(qr.solve(X, y)), as.vector(beta_hat))
```

## Residuals, $s^2$, and standard errors

```{r}
#| label: ols-se
e   <- y - X %*% beta_hat            # residuals
s2  <- sum(e^2) / (n - k)            # unbiased estimator of sigma^2
V   <- s2 * solve(XtX)               # classical variance matrix
se  <- sqrt(diag(V))

round(cbind(estimate = beta_hat, se = se,
            t = beta_hat / se), 4)
```

```{r}
#| label: ols-se-check
round(summary(fit)$coefficients, 4)
```

Identical. The default standard errors that `R` prints are exactly
$\sqrt{\operatorname{diag}\!\big(s^2(\bm{X}'\bm{X})^{-1}\big)}$, 
i.e. they *assume homoskedasticity*. Next week we replace them.

## The mechanical orthogonality conditions

```{r}
#| label: orthogonality
round(crossprod(X, e), 8)     # X'e = 0, up to floating point
round(sum(e), 8)              # residuals sum to zero (constant included)
round(cor(X %*% beta_hat, e), 8)  # fitted values orthogonal to residuals
```

These are identities, true in *every* sample no matter how badly specified the
model is. They are not evidence of anything.

# The geometry: $\bm{P}$ and $\bm{M}$

With $n = `r nrow(X)`$, the matrix $\bm{P}$ is
$`r nrow(X)` \times `r nrow(X)`$, about
`r round(nrow(X)^2 * 8 / 1e9, 1)` GB if we formed it. So we work with a random
subsample to demonstrate the properties, and use the implicit form for the full
data.

```{r}
#| label: projection
set.seed(1)
idx <- sample(n, 300)
Xs  <- X[idx, ]; ys <- y[idx]

P <- Xs %*% solve(crossprod(Xs)) %*% t(Xs)
M <- diag(nrow(Xs)) - P

## symmetric
all.equal(P, t(P))
## idempotent
all.equal(P %*% P, P)
all.equal(M %*% M, M)
## orthogonal to each other
max(abs(P %*% M))
## P X = X and M X = 0
all.equal(P %*% Xs, Xs, check.attributes = FALSE)
max(abs(M %*% Xs))
## traces are the dimensions of the two subspaces
c(trace_P = sum(diag(P)), k = ncol(Xs),
  trace_M = sum(diag(M)), n_minus_k = nrow(Xs) - ncol(Xs))
```

## Leverage

The diagonal of $\bm{P}$ is the leverage $h_{ii}$. On the full sample we get
it without ever forming $\bm{P}$:

```{r}
#| label: leverage
h <- hatvalues(fit)
c(mean = mean(h), theoretical = k / n, max = max(h))
```

```{r}
#| label: fig-leverage
#| fig-cap: "Leverage against years of schooling. The high-leverage observations are those with unusual *covariate* values, not unusual outcomes."
tibble::tibble(h = h, educ = gss$educ, exper = gss$exper) |>
  ggplot(aes(educ, h)) +
  geom_jitter(width = .18, alpha = .18, size = .5, colour = "navy") +
  geom_hline(yintercept = k / n, linetype = 2, colour = "firebrick") +
  labs(x = "Years of schooling", y = expression(h[ii])) +
  theme_minimal(base_size = 10)
```

## The sum-of-squares decomposition

```{r}
#| label: ssq
yhat <- fitted(fit)
TSS <- sum((y - mean(y))^2)
ESS <- sum((yhat - mean(y))^2)
RSS <- sum(residuals(fit)^2)

c(TSS = TSS, ESS_plus_RSS = ESS + RSS)
c(R2_manual = ESS / TSS, R2_lm = summary(fit)$r.squared)
```

And the model-comparison reading of $R^2$: it is $PRE$ against the
intercept-only model.

```{r}
#| label: pre
fit_c <- lm(lnearn ~ 1, data = gss)              # MODEL C
SSE_C <- sum(residuals(fit_c)^2)
SSE_A <- sum(residuals(fit)^2)                    # MODEL A
c(PRE = (SSE_C - SSE_A) / SSE_C, R2 = summary(fit)$r.squared)
```

# Frisch--Waugh--Lovell

We now verify FWL on the return to schooling. Partition
$\bm{X} = [\,\bm{X}_1 \ \ \bm{X}_2\,]$ with $\bm{X}_1 =$ `educ`
and $\bm{X}_2 =$ everything else.

```{r}
#| label: fwl
## Step 1: residualize the treatment on the controls
d_tilde <- residuals(lm(educ ~ exper + I(exper^2) + female + black, data = gss))

## Step 2: residualize the outcome on the controls
y_tilde <- residuals(lm(lnearn ~ exper + I(exper^2) + female + black, data = gss))

## Step 3: bivariate regression of the residuals
b_fwl <- coef(lm(y_tilde ~ d_tilde))["d_tilde"]

c(full_regression = coef(fit)["educ"], fwl = b_fwl)
```

Both forms of the regression-anatomy formula:

```{r}
#| label: anatomy
c(cov_ytilde_dtilde = cov(y_tilde, d_tilde) / var(d_tilde),
  cov_y_dtilde      = cov(gss$lnearn, d_tilde) / var(d_tilde))
```

In fact *all three* numerators agree, for the same reason:

```{r}
#| label: anatomy-three
c(both_residualized = cov(y_tilde, d_tilde) / var(d_tilde),
  only_regressor    = cov(gss$lnearn, d_tilde) / var(d_tilde),
  only_outcome      = cov(y_tilde, gss$educ) / var(d_tilde))
```

What you *cannot* change is the **denominator**:

```{r}
#| label: anatomy-wrong
c(correct = coef(lm(y_tilde ~ d_tilde))["d_tilde"],
  wrong   = coef(lm(y_tilde ~ gss$educ))[2])
```

::: callout-note
**Why does the second one fail?** Regressing $\tilde Y$ on raw `educ` divides by
$\Var_n(\texttt{educ})$ instead of $\Var_n(\widetilde{\texttt{educ}})$.
Since residualizing can only shrink the variance, the denominator is too large
and the estimate is attenuated. The numerators are all equal because the part
removed from $Y$ lies in the span of the controls and is orthogonal to
$\tilde D$. Work this out on paper; it is Question 2 of Problem Set 1.
:::

The residuals are also identical, as the theorem promises:

```{r}
#| label: fwl-resid
all.equal(unname(residuals(lm(y_tilde ~ d_tilde))), unname(residuals(fit)))
```

## The added-variable plot

```{r}
#| label: fig-avp
#| fig-cap: "Added-variable plot for years of schooling. The slope of the fitted line *is* the multivariate coefficient on `educ`."
tibble::tibble(d_tilde, y_tilde) |>
  ggplot(aes(d_tilde, y_tilde)) +
  geom_point(alpha = .12, size = .5, colour = "navy") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = .7) +
  labs(x = "Years of schooling | controls", y = "Log earnings | controls") +
  theme_minimal(base_size = 10)
```

Note how much of the spread in `educ` survives residualization
(`sd(d_tilde)` $=$ `r round(sd(d_tilde), 2)` versus
`sd(educ)` $=$ `r round(sd(gss$educ), 2)`). When that ratio gets small, the
coefficient is being estimated off a thin slice of the data, and its standard
error will show it.

# Omitted variable bias, numerically

Short regression omits verbal ability and parental education; long regression
includes them.

```{r}
#| label: ovb
long  <- lm(lnearn ~ educ + exper + I(exper^2) + female + black +
              wordsum + pareduc, data = gss)
short <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)

## auxiliary regressions: each omitted variable on the included ones
aux_w <- lm(wordsum ~ educ + exper + I(exper^2) + female + black, data = gss)
aux_p <- lm(pareduc ~ educ + exper + I(exper^2) + female + black, data = gss)

beta_short <- coef(short)["educ"]
beta_long  <- coef(long)["educ"]
bias_pred  <- coef(long)["wordsum"] * coef(aux_w)["educ"] +
              coef(long)["pareduc"] * coef(aux_p)["educ"]

round(c(short = beta_short,
        long = beta_long,
        long_plus_bias = beta_long + bias_pred,
        bias = bias_pred), 5)
```

The identity $\tilde\beta_1 = \beta_1 + \sum_j \beta_{2j}\delta_{1j}$ holds
exactly, to machine precision. Note the direction: ability and family background
are positively associated with schooling and positively associated with
earnings, so omitting them **overstates** the return to schooling.

```{r}
#| label: ovb-table
modelsummary(list("(1) Bivariate" = lm(lnearn ~ educ, data = gss),
                  "(2) + demographics" = short,
                  "(3) + ability, background" = long),
             coef_map = c("educ" = "Years of schooling",
                          "wordsum" = "Vocabulary score",
                          "pareduc" = "Parental education"),
             gof_map = c("nobs", "r.squared"),
             stars = TRUE, output = "markdown")
```

::: callout-important
Everything in this section is *algebra*. It tells you how the estimate moves
when you add a control. It does **not** tell you whether the long regression
recovers the causal return to schooling, for that you would need to argue
that nothing else is omitted, which is a substantive claim, not a statistical
one. That argument is what Weeks 5--11 are about.
:::

# Interpretation: functional form

## Logs

With $\ln Y$ on the left, a coefficient $\beta$ is approximately the
proportional change in $Y$ per unit of $X$. The exact percentage change is
$100(e^{\beta}-1)$.

```{r}
#| label: logs
b <- coef(long)["educ"]
c(approx_pct = 100 * b, exact_pct = 100 * (exp(b) - 1))
```

For small $\beta$ the two agree; by $\beta = 0.5$ they do not
($50\%$ vs $64.9\%$).

## Quadratics

`exper` enters quadratically, so the marginal effect depends on where you
evaluate it:

```{r}
#| label: quadratic
b1 <- coef(long)["exper"]; b2 <- coef(long)["I(exper^2)"]
sapply(c(0, 10, 20, 30, 40), function(x) b1 + 2 * b2 * x) |>
  setNames(paste0("exper=", c(0, 10, 20, 30, 40))) |> round(4)

-b1 / (2 * b2)      # turning point, in years of experience
```

## Interactions

```{r}
#| label: interaction
fit_int <- lm(lnearn ~ educ * female + exper + I(exper^2) + black +
                wordsum + pareduc, data = gss)
round(coef(fit_int)[c("educ", "female", "educ:female")], 4)
```

The return to schooling is `educ` for men and `educ + educ:female` for women.
Never read the `female` main effect as "the gender gap": with an interaction it
is the gap at `educ == 0`, which is outside the data.

## Saturating the key covariate

```{r}
#| label: saturate
fit_sat <- lm(lnearn ~ factor(educ) + exper + I(exper^2) + female + black +
                wordsum + pareduc, data = gss)
c(linear_R2 = summary(long)$r.squared,
  saturated_R2 = summary(fit_sat)$r.squared)
anova(long, fit_sat)
```

The $F$-test here is exactly the model comparison from *Social Statistics I*:
is the reduction in $SSE$ from letting every year of schooling have its own
coefficient large enough to justify the extra parameters?

# Exercises

Work these during lab; they are not collected, but Problem Set 1 builds directly
on them.

1.  **Rank.** Construct a design matrix that includes a full set of nine `region`
    dummies *and* an intercept. Confirm with `qr()$rank` that it is rank
    deficient, and confirm what `lm()` does about it.

2.  **The trace identity.** On the 300-observation subsample, verify numerically
    that $\E[\hat{\bm e}'\hat{\bm e}] = \sigma^2(n-k)$ by simulation: fix
    `Xs`, draw $\bm\epsilon \sim N(0, 4\bm{I})$, generate
    $\bm y = \bm{X}_s\bm{\beta} + \bm\epsilon$ for some $\bm{\beta}$ you
    choose, and average $\hat{\bm e}'\hat{\bm e}$ over 2,000 replications.
    Compare to $4 \times (300 - k)$.

3.  **FWL as fixed effects.** Regress `lnearn` on `educ` and a full set of
    `region` dummies. Then (a) demean `lnearn` and `educ` within `region` and
    regress the demeaned outcome on the demeaned regressor. Confirm the slopes
    match. This is the within estimator we will meet in Week 10.

4.  **Sign the bias.** Suppose you omit `hours` from the long regression. Predict
    the sign of the bias in the `educ` coefficient *before* running anything,
    using the OVB formula and the two auxiliary regressions. Then check.

5.  **Leverage.** Find the ten highest-leverage observations. What makes them
    unusual? Refit dropping them, how much does the `educ` coefficient move?
    (Compare to its standard error before concluding anything.)

6.  **Variance inflation.** Compute $\Var(\hat\beta_{\texttt{educ}})$ from the
    formula $s^2 / \sum_i \tilde d_i^2$ using the residualized `d_tilde` from
    the FWL section, and confirm it matches the `educ` entry of
    $s^2(\bm{X}'\bm{X})^{-1}$.

# Session information

```{r}
#| label: session
sessionInfo()
```
