---
title: "\\textbf{Lab 1 Exercise Solutions}"
subtitle: "SOCIOL 723, Social Statistics II"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Cov}{Cov}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.2, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

```{r}
#| label: packages
library(dplyr)
gss <- readRDS("../Data/gss_earnings.rds")
fit <- lm(lnearn ~ educ + exper + I(exper^2) + female + black, data = gss)
```

# Exercise 1: The dummy variable trap

```{r}
#| label: ex1
length(coef(lm(lnearn ~ factor(region), data = gss)))   # intercept + 8 dummies

## force all nine indicators in alongside an intercept
X9 <- model.matrix(~ 0 + factor(region), data = gss)
colnames(X9) <- paste0("r", 1:9)
coef(lm(gss$lnearn ~ X9))
```

`lm()` fits nine coefficients, not ten: with `factor(region)` it drops one
category automatically, and when we force all nine indicators in, it returns
`NA` for the last one. The nine dummies sum to one for every respondent ---
exactly the intercept column --- so once eight dummies and the intercept are
in, the ninth is a perfect linear combination of them and carries no new
information. `R` resolves the trap by refusing to estimate the redundant
coefficient; the dropped category becomes the reference against which the
others are measured.

# Exercise 2: The $n-2$, by simulation

```{r}
#| label: ex2
set.seed(1)
n_sim <- 2000
xs <- sample(8:20, n_sim, replace = TRUE)

rss <- replicate(2000, {
  y <- 8.2 + 0.12 * xs + rnorm(n_sim, 0, 2)     # sigma^2 = 4
  sum(residuals(lm(y ~ xs))^2)
})
c(mean_rss = mean(rss),
  sigma2_x_nminus2 = 4 * (n_sim - 2),
  wrong_guess = 4 * n_sim)
```

The average of $\sum_i \hat e_i^2$ sits on top of $\sigma^2(n-2)$, not
$\sigma^2 n$. Fitting the two coefficients absorbs exactly two degrees of
freedom's worth of error variation --- the residuals are systematically a
little smaller than the errors they estimate --- and dividing by $n-2$
instead of $n$ is precisely what makes $s^2$ unbiased.

# Exercise 3: Sign the bias

Prediction first. The bias from omitting `hours` is
$\beta_{\texttt{hours}} \times \delta_{\texttt{educ}}$, where
$\delta_{\texttt{educ}}$ is the `educ` coefficient from the auxiliary
regression of `hours` on the included regressors. More hours should mean
higher earnings ($\beta_{\texttt{hours}} > 0$); whether the more educated
work more or fewer hours is the genuinely uncertain sign.

```{r}
#| label: ex3
fit_h <- update(fit, . ~ . + hours)
aux   <- lm(hours ~ educ + exper + I(exper^2) + female + black, data = gss)

c(beta_hours = coef(fit_h)["hours"],
  delta_educ = coef(aux)["educ"])
c(short_minus_long = coef(fit)["educ"] - coef(fit_h)["educ"],
  product          = coef(fit_h)["hours"] * coef(aux)["educ"])
```

Both pieces come out positive --- conditional on the controls, the more
educated work slightly more hours, and hours pay --- so the short regression
*overstates* the schooling coefficient. The OVB identity holds to machine
precision, but note the magnitude: the bias is well under one percent of the
coefficient. Signing a bias and sizing it are different exercises.

# Exercise 4: The FWL standard error

```{r}
#| label: ex4
d_tilde <- residuals(lm(educ ~ exper + I(exper^2) + female + black, data = gss))
s2 <- sum(residuals(fit)^2) / (nrow(gss) - length(coef(fit)))

c(fwl     = sqrt(s2 / sum(d_tilde^2)),
  summary = coef(summary(fit))["educ", "Std. Error"])
```

Identical. The standard error on `educ` in the full regression runs on the
*residualized* variation $\sum_i \tilde d_i^{\,2}$ --- the part of schooling
the controls cannot predict --- which is why adding controls correlated with
`educ` inflates its standard error even as it sharpens its interpretation.

# Exercise 5: Saturate the key covariate

```{r}
#| label: ex5
fit_sat <- lm(lnearn ~ factor(educ) + exper + I(exper^2) + female + black,
              data = gss)
anova(fit, fit_sat)
```

The model comparison rejects decisively: freeing every year of schooling to
have its own coefficient buys more fit than its extra parameters cost, so
the earnings--schooling relationship is not exactly linear (the jumps at 12
and 16 years --- credential effects --- are the usual culprits). That is not
a verdict against the linear model: the single slope remains an excellent
one-number summary, and Week 1's lesson applies --- a line fit to a curved
relationship estimates the best linear approximation to it. Saturating is
how you check what that approximation is smoothing over.
