---
title: "\\textbf{Lab 3 Exercise Solutions}"
subtitle: "SOCIOL 723, Social Statistics II"
author: "Wenhao Jiang"
format:
  pdf:
    pdf-engine: pdflatex
    documentclass: article
    classoption: ["11pt"]
    geometry: "top=1in, bottom=1in, left=1.1in, right=1.1in"
    colorlinks: true
    include-in-header:
      text: |
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath,amssymb,bm}
        \usepackage{booktabs}
        \DeclareMathOperator{\Var}{Var}
        \newcommand{\E}{E}
        \newcommand{\loglik}{\ell}
        \newcommand{\Info}{\mathcal{I}}
        \usepackage{setspace}
        \setstretch{1.12}
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.width = 6,
                      fig.height = 3.2, fig.align = "center", dpi = 200)
options(digits = 4, scipen = 999)
```

```{r}
#| label: packages
library(dplyr)
gss <- readRDS("../Data/gss_earnings.rds")
```

# Exercise 1: Bernoulli information

```{r}
#| label: ex1
set.seed(723)
n <- 200
y <- rbinom(n, size = 1, prob = 0.3)

nll <- function(theta) -sum(dbinom(y, size = 1, prob = theta, log = TRUE))
fit <- optim(par = 0.5, fn = nll, method = "Brent", lower = 0.001,
             upper = 0.999, hessian = TRUE)
th  <- fit$par

c(mle = th, ybar = mean(y),
  se_hessian = sqrt(1 / as.numeric(fit$hessian)),
  se_by_hand = sqrt(th * (1 - th) / n))
```

The MLE is the sample proportion, and the two standard errors agree to the
precision of the numerical derivative. The Hessian is a finite-difference
approximation to $-\loglik''(\hat\theta) = n/[\hat\theta(1-\hat\theta)]$, the
expression derived in lecture, so this is a check on the software, not on the
theory.

# Exercise 2: Curvature is information

```{r}
#| label: fig-ex2
#| fig-cap: "Poisson log-likelihoods for four sample sizes, each shifted to peak at zero."
set.seed(723)
th_grid <- seq(1.2, 3.0, length.out = 300)
ll_shift <- function(n) {
  y <- rpois(n, 2)
  ll <- sapply(th_grid, function(t) sum(dpois(y, t, log = TRUE)))
  ll - max(ll)
}
curves <- sapply(c(25, 100, 400, 1600), ll_shift)

matplot(th_grid, curves, type = "l", lty = 1, lwd = 2,
        col = c("grey70", "steelblue", "navy", "firebrick"),
        ylim = c(-40, 0), xlab = expression(theta), ylab = "log-likelihood, peak at 0")
legend("bottomright", legend = paste0("n = ", c(25, 100, 400, 1600)),
       col = c("grey70", "steelblue", "navy", "firebrick"), lwd = 2, bty = "n")
```

Each time $n$ quadruples the peak gets visibly sharper: moving the same
distance from $\hat\theta$ costs four times as much fit. That is
$\Info_n = n\,\Info_1$ read off the picture: the second derivative is a sum
over observations, so it scales with $n$, and the width of the peak, which is
$1/\sqrt{\Info_n}$, halves each time.

# Exercise 3: Newton--Raphson by hand

```{r}
#| label: ex3
S  <- function(th) 5 / th - 1
H  <- function(th) -5 / th^2

th <- 1.5
out <- NULL
for (t in 0:4) {
  step <- -S(th) / H(th)
  out  <- rbind(out, c(guess = th, slope = S(th), minus_l2 = -H(th), step = step))
  th   <- th + step
}
round(out, 2)
```

The table matches the slide. The step is the ratio slope $\div(-\loglik'')$.
Near the peak $-\loglik''$ is at its smallest, so the "learning rate"
$1/(-\loglik'')$ is at its largest, but the slope is heading to zero faster
than the learning rate grows, and their product shrinks; the last step lands
on $\hat\theta = 5$ almost exactly.

# Exercise 4: Not unbiased, but consistent

```{r}
#| label: ex4
set.seed(723)
sig2_mle <- function(n, reps = 5000) {
  replicate(reps, {
    x <- rnorm(n); y <- 1 + 0.5 * x + rnorm(n)
    e <- residuals(lm(y ~ x))
    sum(e^2) / n
  })
}
s20  <- sig2_mle(20)
s200 <- sig2_mle(200)

rbind(n_20  = c(mean_mle = mean(s20),  expected = (20 - 2) / 20),
      n_200 = c(mean_mle = mean(s200), expected = (200 - 2) / 200))
```

At $n = 20$ the MLE of $\sigma^2$ averages about $0.9$, not $1$: the
$(n-2)/n$ from lecture. At $n = 200$ the shortfall is one percent. The bias is
real and it disappears as $n$ grows, which is exactly what maximum likelihood
promises: consistency, not unbiasedness.

# Exercise 5: Where the $z$-test fails

```{r}
#| label: ex5
n <- 20; th0 <- 0.1
ll <- function(theta, s) ifelse(s == 0, 0, s * log(theta)) + (n - s) * log(1 - theta)
rulers <- function(s) {
  th <- s / n
  se <- sqrt(th * (1 - th) / n)
  c(theta_hat = th, se = se, z = (th - th0) / se, W = ((th - th0) / se)^2,
    LR = 2 * (ll(th, s) - ll(th0, s)))
}
round(rbind(zero_successes = rulers(0), one_success = rulers(1)), 3)
```

With zero successes the standard error is zero and the Wald ratio is
undefined: the test cannot even be formed. The LR statistic exists,
$LR = 4.21$, and rejects $\theta_0 = 0.1$. With one success both can be
computed, and both accept, but at different values, $W = 1.05$ against
$LR = 0.67$, because the log-likelihood at $\hat\theta = 0.05$ is nowhere
near a parabola: it drops steeply toward zero and gently toward larger
$\theta$. The Wald test extrapolates the width at the peak; the LR test
measures the actual loss of fit between $\hat\theta$ and $\theta_0$.

Which to trust: lecture's exact table for $n = 20$ says neither is at
its nominal 5% here. Zero successes has probability $0.9^{20} = 12\%$ under
the null, so any test that rejects it (LR does) has size at least 12%, and
any test that cannot score it (Wald) is not a test at all. With two expected
successes there is no parabola to read, and the honest answer is an exact
binomial test. What survives is the ranking: the Wald ruler is the one that
breaks, in both directions, because it reads only the peak; from $n = 50$
onward LR sits nearest 5% and is the one to believe when they disagree.
